---
title: "Lebesgue Integration & Measure-Theoretic Probability for Econometricians"
author: "Eliab Luvanda"
date: "2025-08-09"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: true
    code-fold: false
    lof: true
    lot: true
---

\pagebreak
# Motivation: Why Measure Theory in Econometrics?

- **Econometric models** require integration over probability spaces, especially for expectations and variances.
- Classical Riemann integration is limited; Lebesgue integration generalizes the concept, handling more complex random variables and probability spaces.
- Measure-theoretic probability provides a rigorous framework for advanced econometric theory.

# Measure Theory Basics

## Sets & Sigma-Algebras

- **Sample Space ($\Omega$):** Universe of outcomes.
- **Sigma-Algebra ($\mathcal{F}$):** A collection of subsets of $\Omega$ (including $\Omega$ and $\varnothing$) closed under countable unions and complements. Ensures "measurability" of events.

## Measures

- **Measure $\mu$:** Function $\mu: \mathcal{F} \to [0, \infty]$, assigning "size" to sets.
- **Probability Measure $P$:** Special case, $P(\Omega) = 1$, representing certainty.

# Lebesgue Integration

## Motivation

- Riemann integral partitions the domain (horizontal slices).
- Lebesgue integral partitions the range (vertical slices); better suited for complex/discontinuous functions.

## Definition

### Simple Function

A simple function $s(x)$ is:
$$
s(x) = \sum_{k=1}^n a_k \mathbb{I}_{A_k}(x)
$$
where $A_k \in \mathcal{F}$, $a_k \geq 0$.

### Lebesgue Integral (non-negative $f$)
$$
\int_\Omega f \, d\mu = \sup \left\{ \int_\Omega s \, d\mu : 0 \leq s \leq f, ~ s \text{ is simple} \right\}
$$

### Key Properties

- **Linearity:** $\int(f + g)d\mu = \int f d\mu + \int g d\mu$
- **Monotonicity:** If $f \leq g$, then $\int f d\mu \leq \int g d\mu$
- **Dominated Convergence Theorem (DCT)**
- **Fatou’s Lemma**
- **Fubini’s Theorem**

## Probabilistic Interpretation

- **Expectation:**
$$
E[X] = \int_\Omega X(\omega) ~ P(d\omega)
$$
Applies to discrete, continuous, and mixed distributions.

# Measure-Theoretic Probability

## Probability Space

- **Triple $(\Omega, \mathcal{F}, P)$:**
  - $\Omega$: Sample space
  - $\mathcal{F}$: Sigma-algebra
  - $P$: Probability measure

## Random Variables

- **Random Variable ($X$):** Measurable function $X : \Omega \to \mathbb{R}$
- **Measurability:** For all Borel sets $B$, $X^{-1}(B) \in \mathcal{F}$

## Distribution

- **Pushforward Measure $\mu_X$ of $X$:**
$$
\mu_X(B) = P(X \in B) = P(X^{-1}(B))
$$

# Practical Econometric Applications

- **Expectations, variances, covariances**: All defined as Lebesgue integrals.
- **Laws of Large Numbers and Central Limit Theorem:** Measure-theoretic convergence (almost sure, in probability) is key.
- **Likelihood estimation:** Supports both continuous and mixed models.

# Key Theorems for Econometricians

## Dominated Convergence Theorem (DCT)
If $f_n \to f$ and $|f_n| \leq g$ (integrable),
$$
\lim_{n \to \infty} \int f_n\,d\mu = \int f\,d\mu
$$

## Fubini’s Theorem
Interchange order of integration:
$$
\int_\Omega \int_\Theta f(x,y)\,d\nu(y)\,d\mu(x) = \int_\Theta \int_\Omega f(x,y)\,d\mu(x)\,d\nu(y)
$$

| Feature                  | Riemann Integral         | Lebesgue Integral           |
|--------------------------|-------------------------|-----------------------------|
| Partitioning             | Domain (subintervals)   | Range (value slices)        |
| Functions handled        | Continuous              | Measurable                  |
| Probabilistic Models     | Limited                 | Natural & robust            |
| Convergence Theorems     | Few                     | DCT, Fatou’s, etc.          |
| Use for Expectations     | Simple RVs only         | General probability models  |

: Summary Table: Riemann vs Lebesgue Integration

# Essential Reading

- Billingsley, P. *Probability and Measure*
- Dudley, R. *Real Analysis and Probability*
- Chung, K.L. *A Course in Probability Theory*

# Concluding Remarks

- **Lebesgue integration** and **measure-theoretic probability** are foundational for econometric theory, supporting advanced models, convergence concepts, expectations, and asymptotics.

# Practice Problems

1. **Probability Spaces:** Construct a probability space for a Markov process, identify measurable events.
2. **Expectation Calculation:** If $X = \mathbb{I}_A + 2\mathbb{I}_B$, $A, B \in \mathcal{F}$, compute $E[X]$.
3. **Dominated Convergence:** State and apply the Dominated Convergence Theorem to a sequence of expected values.

---

**End of Notes**

# Visualizations

## 1. Venn Diagram (Classical/Frequentist Events)
```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false
#| fig.width: 3
#| fig.height: 3
#| out.width: "60%"
#| fig.cap: "Venn Diagram of Events A and B"
# Install packages if needed
if (!require("VennDiagram")) install.packages("VennDiagram", repos = "https://cloud.r-project.org")
library(VennDiagram)

venn.plot <- draw.pairwise.venn(area1 = 60, area2 = 40, cross.area = 20,
              category = c("Event A", "Event B"),
              fill = c("skyblue", "orange"), alpha = 0.5,
              cat.pos = c(-20, 20), cat.cex = 1.2)
grid.draw(venn.plot)
```


**Purpose:**  
A Venn diagram visually demonstrates how events and their relationships (intersection, union, disjointness) are handled in classical and frequentist probability.

**Detailed Explanation:**  
- **Events:** Each circle represents a possible event (e.g., getting a head in a coin toss, or an ace in a card draw).
- **Intersection:** The overlapping region shows outcomes shared by both events (e.g., outcomes where both "A" and "B" occur).
- **Union:** The total area covered by circles represents "A or B."
- **Disjointness:** If circles don’t overlap, events are mutually exclusive.
- Use this plot to discuss how probabilities of unions and intersections are calculated (e.g., $P(A \cup B) = P(A) + P(B) - P(A \cap B)$).

## Relative Frequency Plot (Law of Large Numbers)

```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false
#| fig.width: 6
#| fig.height: 4
#| out.width: "80%"
#| fig.cap: "Empirical Probability Converges to True Probability"
set.seed(42)
n_trials <- 1000
p_heads <- 0.5
results <- rbinom(n_trials, 1, p_heads)
cum_freq <- cumsum(results) / (1:n_trials)

# Plotting the relative frequency of heads over trials
# Use ggplot2 for better aesthetics
library(ggplot2)
ggplot(data.frame(trials=1:n_trials, freq=cum_freq), aes(x=trials, y=freq)) +
  geom_line(color="darkblue", linewidth = 1) +
  geom_hline(yintercept=p_heads, linetype="dashed", color="red") +
  labs(
#    title="Empirical Probability Converges to True Probability",
       x="Number of Trials", y="Relative Frequency") +
  theme_classic()
```

**Purpose:**  
Shows how empirical (experimentally observed) frequencies converge to theoretical probabilities as the number of trials increases—a foundation of frequentist probability and statistics.

**Detailed Explanation:**  
- **Empirical Line:** The blue line tracks the proportion of heads in repeated coin tosses.
- **Theoretical Probability:** The dashed red line marks the true value ($p = 0.5$ for a fair coin).
- **Convergence:** As the number of trials increases, the empirical proportion stabilizes around the theoretical probability, visually illustrating the Law of Large Numbers.

## Bayesian Updating (Prior and Posterior PDFs)

```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false
#| fig.width: 6
#| fig.height: 4
#| out.width: "80%"
#| fig.cap: "Bayesian Updating: Prior, Likelihood, and Posterior Distributions"
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)
library(dplyr)

x <- seq(0, 1, length=200)
prior <- dbeta(x, 2, 5)
likelihood <- dbeta(x, 8, 2)
posterior <- dbeta(x, 10, 7)

df <- data.frame(
  Probability=x,
  Prior=prior,
  Likelihood=likelihood,
  Posterior=posterior
)

df_long <- df %>% tidyr::pivot_longer(cols=-Probability, names_to="Type", values_to="Density")

ggplot(df_long, aes(x=Probability, y=Density, color=Type)) +
  geom_line(linewidth = 1.2) +
#  ggtitle("Bayesian Updating: Prior, Likelihood, and Posterior Distributions") +
  theme_classic()
```

**Purpose:**
Visualizes how Bayesian updating modifies beliefs (prior) based on new evidence (likelihood) to produce a posterior distribution.

**Detailed Explanation:**

- **Prior Distribution:** Represents initial beliefs about a parameter (e.g., probability of success).
- **Likelihood:** Represents the evidence or data observed (e.g., number of successes in trials).
- **Posterior Distribution:** Updated beliefs after considering the likelihood, combining prior and evidence.
- **Interpretation:** The posterior reflects how the prior is adjusted by the likelihood, showing the Bayesian approach to probability and inference.

## Probability Space Diagram (Measure-Theoretic)

Measure-Theoretic Probability Space can handle infinite & uncountable sample spaces, such as a dart board.

Let's consider a **dart board** as a classic example of an infinite and uncountable probability space, which is a fundamental concept in measure-theoretic probability.

### Description

- **Sample Space ($\Omega$)**: The set of all points on the dart board, mathematically represented as the set of all ordered pairs $$(x, y)$$ in the unit disk:
  $$
  \Omega = \{(x, y) \mid x^2 + y^2 \leq 1\}
  $$
  This set is **uncountably infinite**, meaning there are infinitely many points, and you cannot list them in a sequence.

- **Events ($A \subset \Omega$)**: An event is any measurable subset of the dart board — for instance, a specific region, such as a sector or an annulus.

- **Probability Measure ($\mathbb{P}$)**: The probability of "hitting" a subset $A$ is defined by the area of $A$ divided by the total area of the dart board. Formally,
  $$
  \mathbb{P}(A) = \frac{\text{Area}(A)}{\pi}
  $$
  (assuming the dart lands uniformly at random on the board of area $\pi$.)





```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false
#| fig.width: 2
#| fig.height: 2
#| out.width: "50%"
#| fig.cap: "Dartboard Visualization: Sample Space Ω and Event A = {x < 0}"
# Use ggplot2 to create a dartboard visualization, and add legend
library(ggplot2)

theta <- seq(0, 2*pi, length.out=500)
x <- cos(theta)
y <- sin(theta)
theta_left <- seq(pi/2, 3*pi/2, length.out=250)
x_left <- c(0, cos(theta_left), 0)
y_left <- c(0, sin(theta_left), 0)

df_disk <- data.frame(x = x, y = y)
df_left <- data.frame(x = x_left, y = y_left)

ggplot() +
  geom_polygon(data = df_disk, aes(x = x, y = y), fill = "lightblue", alpha = 0.5) +
  geom_polygon(data = df_left, aes(x = x, y = y), fill = "red", alpha = 0.5) +
  coord_fixed() +
#  labs(title = "Dartboard: Sample Space Ω and Event A") +
  theme_void() +
  theme(legend.position = "topright") +
  scale_fill_manual(values = c("lightblue" = "lightblue", "red" = "red"),
                    labels = c(expression(Omega), "A")) +
  guides(fill = guide_legend(title = "Regions"))

```

### Key Ideas from Probability Theory:

-	Sample Space (Ω):
The unit disk shows every possible outcome—the set of all points where a dart can land. This is an infinite, uncountable sample space.
	
-	Event ($A = {(x, y): x < 0}$):
The left half is highlighted in red—representing a measurable event (a subset of possible outcomes).
	
-	Probability Measure:
In continuous spaces, probability is assigned by area:

For this symmetric case, the probability of hitting the left half is 50%.
This visualization demonstrates the essence of measure-theoretic probability, where events are regions of a continuous space, and likelihoods are determined by their size relative to the whole sample space.

## Probability of Hitting a Subset: Measure-Theoretic Perspective

In measure-theoretic probability theory, the **sample space** ($\Omega$) can be a continuous, uncountable set of outcomes—such as every possible point where a dart can land on a circular board. Suppose we are interested in the probability that the dart lands inside a **smaller circle** (event $A$) that lies within the larger circle (the sample space).

This setting formalizes probability in terms of **measures** (such as area), rather than counting discrete outcomes. Specifically:

- **Sample Space ($\Omega$):**  
  The entire large circle, representing all possible points where the dart may land.

- **Event ($A$):**  
  Any measurable subset of $\Omega$. In this scenario, $A$ is the smaller circle (possibly offset from the center) entirely or partially within the larger one.

- **Probability Measure:**  
  The probability of hitting any point in $A$ is given by the ratio of the area of $A$ to the area of $\Omega$—assuming the dart lands uniformly at random:
  $$
  \mathbb{P}(A) = \frac{\text{Area}(A)}{\text{Area}(\Omega)}
  $$
  If the smaller circle (radius $r$) is entirely inside the larger circle (radius $R$):
  $$
  \mathbb{P}(A) = \frac{\pi r^2}{\pi R^2} = \left(\frac{r}{R}\right)^2
  $$
  If the inner circle overlaps the edge of the larger one, $\mathbb{P}(A)$ is the ratio of the overlapping area to the area of $\Omega$.

**Interpretation:**  
This approach encapsulates the essence of measure-theoretic probability:  
- The probability of an event in a continuous space is computed by comparing its "size" (area or measure) to that of the full sample space.
- Events need not be centered, nor do they have to be simple shapes—they can be any measurable region within $\Omega$.

**Visualization:**  
The following graph (see next section) depicts this relationship, with the **larger circle** representing all possible outcomes and the **smaller, possibly offset circle** highlighting the subset of interest whose probability we are computing.




```{r}
#| echo: false
#| warning: false
#| message: false
#| error: false
#| fig.width: 2
#| fig.height: 2
#| out.width: "50%"
#| fig.cap: "Dartboard Visualization of computing a probability measure of an event A"
# Dartboard Visualization: Offset Inner Circle Embedded in a Square

# Define radii
R <- 1      # Radius of big circle (sample space Ω)
r <- 0.1    # Radius of small circle (event A)

# Inner circle center coordinates (not at the origin)
cx <- 0.2   # x-coordinate of inner circle center
cy <- 0.8     # y-coordinate of inner circle center

# Set up coordinates for the big circle
theta <- seq(0, 2*pi, length.out=500)
x_big <- R * cos(theta)
y_big <- R * sin(theta)

# Coordinates for the offset small circle
x_small <- cx + r * cos(theta)
y_small <- cy + r * sin(theta)

# Set square boundaries
square_min <- -1.5
square_max <- 1.5

# Plot in a square box, no axes, no outer plot box
par(bty="n", mar=c(2,2,2,2))
plot(NA, NA,
     xlim=c(square_min, square_max), ylim=c(square_min, square_max),
     xlab="", ylab="", asp=1,
#     main="Probability Space: Offset Circle in Square",
     axes=FALSE)

# Draw the square boundary
rect(square_min, square_min, square_max, square_max,
     border="black", lwd=0.1)

# Fill the big circle (sample space Ω)
polygon(x_big, y_big, col=rgb(0.6,0.9,1,0.5), border=NA)

# Fill the offset inner circle (event A)
polygon(x_small, y_small, col=rgb(1,0,0,0.5), border=NA)

# Add a smaller legend inside the square
legend("topright",
       legend=c(expression(Omega), "A"),
       fill=c(rgb(0.6,0.9,1,0.5), rgb(1,0,0,0.5)),
       bty="n", cex=0.55)

```


### Key Measure-Theoretic Probability Ideas Illustrated by the Graph

- **Sample Space ($\Omega$):** The larger circle represents all possible outcomes where the dart can land.
- **Event ($A$):** The smaller circle, possibly offset, represents a specific measurable event within the sample space.
- **Probability Measure:** The shaded area of the smaller circle relative to the larger circle illustrates how we compute the probability of event $A$ by comparing its area to that of $\Omega$.

This visualization captures the essence of measure-theoretic probability, where events are defined in terms of their measurable size within a continuous space.
