---
title: "Is Spurious Regression a Problem in Stationary Time Series?"
subtitle: "A Theoretical Framework and Monte Carlo Evidence"
author: "Eliab Luvanda"
date: today
lang: en
bibliography: references.bib

# Print every entry in references.bib, including background reading that is not
# cited at a specific point in the text, so that the list below is a
# bibliography rather than a bare reference list.
nocite: |
  @*

abstract: |
  Regression between independent random walks is spurious: the OLS slope is inconsistent, and has a
  nondegenerate limit and the conventional $t$-ratio diverges. Stationarity removes that
  inconsistency, and it is tempting to conclude that conventional inference is then safe.
  This handout shows why that conclusion is wrong, and exactly how wrong. Under 
  stationarity, ergodicity and weak dependence, OLS is consistent for the population
  projection slope, but the conventional $t$-ratio converges to $N(0,\lambda)$, where
  $\lambda$ is the ratio of the long-run variance of the regression score to the product of
  the two variances the conventional formula uses. The test over-rejects, under-rejects, or
  is correctly sized according to whether $\lambda$ exceeds, falls below, or equals one.
  We decompose $\lambda$ into a contemporaneous and a serial component, which unifies three
  otherwise separate cases: independent AR(1) pairs, general independent linear processes,
  and dependent series with common volatility and a zero projection slope. Monte Carlo
  evidence covers nine data-generating processes, including a design in which both series
  are visibly serially correlated yet the conventional test is exactly correctly sized.
  Further experiments examine HAC inference, size-adjusted local power, local-to-unity
  sequences, and distributed-lag tests under omitted dynamics.

format:
  pdf:
    pdf-engine: lualatex
    documentclass: article
    classoption:
      - twoside
    fontsize: 11pt
    papersize: a4
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    fig-pos: "htbp"

geometry:
  - inner=1.2in
  - outer=1.0in
  - top=1.1in
  - bottom=1.2in
  - headsep=0.3in

mainfont: "TeX Gyre Termes"
sansfont: "TeX Gyre Heros"
monofont: "Inconsolata"
mathfont: "TeX Gyre Termes Math"

header-includes:
  - \usepackage{amsmath, amssymb, amsthm, mathtools}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{setspace}
  - \usepackage{framed}
  - \singlespacing
  - |
    \theoremstyle{plain}
    \newtheorem{theorem}{Theorem}
    \newtheorem{proposition}[theorem]{Proposition}
    \newtheorem{lemma}[theorem]{Lemma}
  - |
    \theoremstyle{definition}
    \newtheorem{assumption}[theorem]{Assumption}
    \newtheorem{exercise}{Exercise}
  - |
    \theoremstyle{remark}
    \newtheorem{remark}[theorem]{Remark}
  - |
    % Explanatory environments. These are unnumbered and set in a smaller sans
    % head so they are visually distinct from the formal results they follow:
    % "Logic" gives the skeleton of the argument, "Intuition" gives the plain-
    % language meaning. Neither is required to follow the formal development.
    \newtheoremstyle{explain}
      {8pt}{8pt}{\small}{}{\bfseries\sffamily}{.}{.5em}{}
    \theoremstyle{explain}
    \newtheorem*{logic}{Logic of the argument}
    \newtheorem*{intuition}{Intuition}
  - \renewcommand{\qedsymbol}{$\blacksquare$}
  - |
    \newcommand{\E}{\mathbb{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\Cov}{\mathrm{Cov}}
    \newcommand{\LRV}{\mathrm{LRV}}
  - \usepackage{fancyhdr}
  - |
    \pagestyle{fancy}
    \fancyhf{}
    \fancyhead[LE,RO]{\thepage}
    \fancyhead[LO]{Spurious Regression in Stationary Time Series}
    \fancyhead[RE]{Econometrics}

execute:
  echo: false
  warning: true
  message: false
---

```{r simulation-data, include=FALSE, cache=TRUE, cache.extra=tools::md5sum("scripts/spurious_stationary_sim.R")}
source("scripts/spurious_stationary_sim.R")
```

# Objectives and Intended Audience {#sec:objectives}

## Purpose

The question in the title is one that almost every empirical economist has met in an informal
form: *my series are stationary, so am I safe?* The usual answer given in coursework is that
spurious regression is a unit-root problem, that differencing or cointegration analysis
disposes of it, and that stationary data may therefore be regressed in the ordinary way. That
answer is incomplete in a way that matters for published work.

The objective of this handout is to replace it with a precise one. Specifically, the handout
aims to establish three things: that stationarity secures the *estimate* but says nothing about
the *standard error*; that whether conventional inference is trustworthy is settled by a single
quantity, $\lambda$, which can be computed for any given data-generating process; and that
$\lambda$ can go wrong in two independent ways, only one of which is detected by the diagnostics
most researchers actually run. A test for serial correlation is not sufficient, and the handout
constructs explicit counterexamples in both directions --- data that are visibly serially
correlated yet require no correction at all, and data with no serial correlation whatsoever that
reject a true null more than twice as often as they should.

The treatment is deliberately self-contained. Every claim is stated as a formal result with its
assumptions written out, explained again in words, and then verified by simulation whose code is
supplied and modifiable. Nothing is asserted that is not either proved or demonstrated
numerically.

## Two audiences

This handout is written for two overlapping groups, who will use it differently.

**Readers building towards econometric theory and teaching.** For staff members undertaking
doctoral work with a view to specialising in econometric theory, and to teaching the subject at
MA and PhD level, the material is intended as a compact case study in how asymptotic arguments
are actually constructed and where they fail. The topic is well chosen for that purpose: it is
elementary enough to be worked through completely, yet it involves a genuine limit theorem, a
non-trivial variance calculation, a distinction between pointwise and uniform approximation, and
a place where an intuitively appealing claim turns out to be false. Readers in this group should
work through the proofs, attempt the exercises, and modify the simulation code --- deriving
$\lambda$ for a process of their own choosing and confirming it numerically is the single most
useful thing they can do with this handout.

Those preparing to teach may also find the *structure* of the handout worth borrowing
independently of its content. Each formal result is followed by a **Logic of the argument** block
giving the skeleton of the proof, and an **Intuition** block giving its meaning in ordinary
language. Separating the three registers --- what is claimed, why it is true, what it means ---
lets students who cannot yet follow the algebra still follow the economics, and lets those who
can follow the algebra see what it is for. The device is reusable in any technical course.

**Readers applying econometrics in their own empirical research.** For academic staff whose
concern is the credibility of applied work, the objective is narrower and more immediate: to be
able to look at a regression on time-series data and judge whether the reported standard errors
mean what they appear to mean. The practical content is concentrated in the Intuition blocks,
in the results of Section \ref{sec:conventional} onwards, and in the final discussion. Readers
in this group may skip every proof without loss. What they should not skip is the decomposition
in Section \ref{sec:decomposition}, because it supplies the two questions that ought to be asked
of any time-series regression, and Section \ref{sec:nearunit}, because it explains why passing a
unit-root test is weaker reassurance than it appears.

## What a reader should be able to do afterwards

On completing the handout, a reader should be able to:

1. state precisely what stationarity does and does not guarantee about OLS inference, and
   explain why the random-walk case is a different problem in kind rather than a more severe
   version of the same one;
2. compute $\lambda$ for a given pair of processes, and recognise from the autocorrelation
   structure alone whether conventional inference will over-reject, under-reject, or be correct;
3. distinguish the two independent sources of failure --- contemporaneous and serial --- and
   identify which diagnostics detect which;
4. choose sensibly among conventional, HC and HAC standard errors, and say what each does not
   fix;
5. explain why a process that is technically stationary may still be badly approximated by
   stationary asymptotics, and why this makes near-unit-root diagnostics complementary to robust
   inference rather than redundant;
6. separate a specification failure from a covariance-estimation failure in a distributed-lag or
   Granger-causality setting, and know that only the second can be repaired by a robust
   covariance matrix; and
7. reproduce and extend the accompanying Monte Carlo evidence.

## Prerequisites and technical level

The handout assumes familiarity with ordinary least squares, the notions of consistency and
asymptotic normality, and the idea of an autocorrelation function. It does not assume
measure-theoretic probability, and no result below requires it. Where a limit theorem is needed
it is stated as an assumption in verifiable form and attributed, rather than proved from
primitives; the mathematics that is carried out in full is confined to variance calculations
that require no more than algebra and the summation of a geometric series.

Readers who prefer to see the argument before the formalism may read the Intuition blocks in
Section \ref{sec:theory} in sequence, ignoring everything between them, and will obtain a
complete if informal account of the entire result.

# Introduction

The problem is older than it is usually made to appear. Yule [-@yule1926nonsense] coined the
term *nonsense-correlation* and, in what is essentially the first Monte Carlo study in
econometrics, drew repeated samples to trace the distribution of the correlation coefficient
across three classes of series. For independent random series it concentrated near zero, as
expected. For *conjunct* series --- serially correlated, but with differences that were not
--- the distribution was noticeably more dispersed. For conjunct series whose differences were
also conjunct, which is essentially the integrated case, it became U-shaped, with most of its
mass near $\pm1$: his motivating example, a correlation of $0.95$ between the Church of England
marriage rate and the standardised mortality rate in England and Wales over 1866--1911, was no
accident of that particular pair of series but the typical outcome for such data.

Both of Yule's pathological classes reappear below. The U-shaped case anticipates, in
simulation, the nondegenerate limit distribution later derived analytically; the intermediate
*conjunct* case is the subject of this handout, and Yule's observation that mere serial
correlation widens the sampling distribution is precisely the phenomenon quantified in Section
\ref{sec:decomposition} by the ratio $\lambda$.

Granger and Newbold [-@granger1974spurious] returned to the question and showed that a
regression between two independent random walks can produce a high $R^2$, a low
Durbin--Watson statistic, and apparently significant coefficients even though the series are
unrelated. Phillips [-@phillips1986understanding] then established the nonstandard asymptotic
theory: the OLS slope has a nondegenerate Brownian-functional limit rather than converging to
zero, while its conventional $t$-ratio diverges and the probability of a false rejection
approaches one. Yule's U-shaped histogram and Phillips's limiting functional are, sixty years
apart, descriptions of the same object.

It is tempting to infer that stationarity makes conventional regression inference harmless.
That inference is too broad. For jointly stationary series with a zero population projection
slope, OLS is consistent for zero under standard regularity conditions, but the conventional
OLS standard error need not estimate the slope's long-run sampling variance. Full independence
is sufficient but unnecessary: cross-lag dependence, common volatility, and other nonlinear
dependence can coexist with a zero contemporaneous projection. The resulting test may
over-reject, under-reject, or have correct asymptotic size. Which case occurs is determined by
the long-run variance of the regression score, not by serial correlation alone.

This handout develops that distinction and evaluates it by Monte Carlo simulation. The
experiment includes equal positive AR coefficients, unequal positive coefficients, an
opposite-sign pair, an i.i.d. benchmark, an AR/MA design in which both series are serially
correlated yet the conventional test is exactly correctly sized, dependent series with common
stochastic volatility, local-to-unity sequences, and independent random walks. It compares
conventional OLS inference with heteroskedasticity-consistent (HC) and
heteroskedasticity-and-autocorrelation-consistent (HAC) standard errors
[@white1980heteroskedasticity; @neweywest1987], measures what HAC costs in size-adjusted local
power, and uses distributed-lag tests to examine Granger noncausality and the consequences of
omitting the dependent variable's dynamics.

The results here are not new. The stationary over-rejection phenomenon was documented by
Granger, Hyung and Jeon [-@ghj2001spurious], and Ventosa-Santaulària [-@ventosa2009spurious]
surveys the wider spurious-regression literature. The purpose of this handout is to assemble
the argument in one place, with the assumptions stated explicitly and the simulation code
available for inspection and modification.

The principal conclusions are:

1. Under suitable stationarity, ergodicity, and weak-dependence conditions, OLS consistently
   estimates the population projection slope; under a zero-projection null it converges to zero.
2. The conventional $t$-ratio converges to $N(0,\lambda)$, where $\lambda$ is a long-run variance
   ratio. Values above, below, and equal to one imply over-rejection, under-rejection, and
   correct asymptotic size, respectively.
3. $\lambda$ decomposes into a contemporaneous component $\kappa$ and a serial component.
   Both must be accounted for: a serially uncorrelated score does *not* by itself deliver
   correct size, and a serially correlated score does not by itself destroy it.
4. For independent AR(1) series, $\lambda$ depends on the product $\phi_x\phi_y$. Equal positive
   persistence produces over-rejection, whereas opposite-sign persistence can produce severe
   under-rejection.
5. HAC inference is asymptotically valid under additional conditions ensuring consistent
   long-run variance estimation, although highly persistent processes can exhibit substantial
   finite-sample distortion. Once critical values are size-adjusted, HAC costs remarkably
   little power.
6. Stationary asymptotics are not uniform near the unit-root boundary, and dynamic
   misspecification can generate false distributed-lag and Granger-causality findings even when
   the candidate predictor is generated independently.

# The Classical Random-Walk Benchmark

Consider two independent random walks,
$$
x_t=x_{t-1}+\varepsilon_t, \qquad y_t=y_{t-1}+\eta_t,
$$
with independent zero-mean i.i.d. innovations of finite variance, and regress $y_t$ on an
intercept and $x_t$. Let
$\bar W_j=\int_0^1W_j(r)\,dr$. Functional central limit theory gives
$$
\hat\beta \Rightarrow
\frac{\int_0^1\{W_x(r)-\bar W_x\}\{W_y(r)-\bar W_y\}\,dr}
     {\int_0^1\{W_x(r)-\bar W_x\}^2\,dr}.
$$
Thus $\hat\beta=O_p(1)$ and does not converge to the true value zero. The conventional standard
error is $O_p(T^{-1/2})$, so the absolute $t$-ratio diverges at rate $T^{1/2}$ and
$$
\Pr(|t|>1.96)\longrightarrow1.
$$
This random-walk benchmark is the classical Granger--Newbold form of spurious regression. The
analysis below contrasts it with fixed-parameter, short-memory stationary processes; it does
not claim that unit roots are the only possible source of nonstandard regression asymptotics.

\begin{intuition}
A random walk wanders. Over any finite stretch it drifts up or down and traces out something
that looks like a trend, even though nothing is pushing it in either direction. Regressing one
random walk on another is therefore not really estimating a relationship between two variables;
it is comparing the shape of one wandering path with the shape of another. Whether the two
happen to drift the same way is a matter of luck in the particular pair of paths you drew.

Collecting more data does not settle the matter, because each new observation adds another step
of wandering rather than another independent piece of evidence about a fixed relationship. That
is why $\hat\beta$ never converges: there is nothing for it to converge to. The $t$-statistic
grows without bound because the formula behind it counts every observation as fresh evidence,
so with $T=1000$ it believes it has a thousand times more information than it really does.
\end{intuition}

# Stationary Regression Theory {#sec:theory}

## Setup and regularity conditions

Consider the population linear projection
$$
y_t=\alpha_0+\beta_0x_t+u_t, \qquad t=1,\ldots,T,
$$
where
$$
\beta_0=\frac{\Cov(x_t,y_t)}{\Var(x_t)},
$$
and test $H_0:\beta_0=0$. Write $\mu_x=\E x_t$, $\mu_y=\E y_t$,
$\sigma_x^2=\Var(x_t)>0$, $\sigma_u^2 = \Var(u_t)$, and define the **regression score**
$$
w_t=(x_t-\mu_x)u_t.
$$

\begin{assumption}[Stationarity, moments, and projection orthogonality]\label{as:stat}
The joint process $\{(x_t,y_t)\}$ is strictly stationary and ergodic, has finite moments of
order $4+\delta$ for some $\delta>0$, and satisfies the population projection condition
$\E[(x_t-\mu_x)u_t]=0$. Under $H_0$, $u_t=y_t-\mu_y$.
\end{assumption}

\begin{assumption}[Weak dependence and central limit theorem]\label{as:clt}
The joint process is sufficiently weakly dependent that
$$
T^{-1/2}\sum_{t=1}^T w_t\Rightarrow N(0,\Omega), \qquad
\Omega=\sum_{k=-\infty}^{\infty}\Cov(w_t,w_{t+k})\in(0,\infty),
$$
and the autocovariances required below are absolutely summable.
\end{assumption}

Standard strong-mixing conditions with suitable moment and mixing-rate restrictions are
sufficient for Assumption \ref{as:clt} [@davidson1994stochastic; @davidson2000econometric].

\begin{assumption}[HAC consistency]\label{as:hac}
For HAC inference, a positive-semidefinite kernel and a bandwidth $L_T$ are used under
conditions ensuring uniform consistency of the sample autocovariances, with
$L_T\to\infty$ and $L_T/T\to0$. These conditions imply
$\widehat{\LRV}(w)\xrightarrow{p}\Omega$.
\end{assumption}

Primitive conditions for Assumption \ref{as:hac} are given by @andrews1991heteroskedasticity.
Note that it is additional to the score central limit theorem: a CLT alone does not
guarantee that a particular HAC estimator consistently estimates $\Omega$.

\begin{intuition}[what the three assumptions are each for]
Read the three assumptions as a division of labour, each buying one thing.

Assumption \ref{as:stat} says the process does not change its character over time and that
averages taken from it settle down as the sample grows. This is what makes the \emph{estimate}
reliable. The moment condition is bookkeeping: we will be averaging products of four random
quantities, so those products need to have a finite average in the first place.

Assumption \ref{as:clt} says that a particular average --- the average of the score $w_t$ ---
is not just settling down but is approximately normally distributed, with a spread measured by
$\Omega$. This is what makes \emph{testing} possible at all, since every critical value we use
comes from a normal distribution. The quantity $\Omega$ is the true spread of that average.

Assumption \ref{as:hac} says we can \emph{estimate} $\Omega$ from the data. This is a genuinely
separate requirement and it is the one most often assumed away. Knowing that a quantity has a
normal distribution with some spread is not the same as being able to measure that spread from
a finite sample. Assumption \ref{as:clt} gives us the first; only Assumption \ref{as:hac} gives
us the second.
\end{intuition}

## Consistency and the conventional test

\begin{proposition}[Consistency]\label{prop:consistent}
Under Assumption \ref{as:stat},
$$
\hat\beta\xrightarrow{p}\beta_0,
$$
and hence $\hat\beta\xrightarrow{p}0$ under $H_0$.
\end{proposition}

\begin{proof}
The OLS slope is the ratio of the sample covariance of $x_t$ and $y_t$ to the sample variance
of $x_t$. By ergodicity, these quantities converge to $\Cov(x_t,y_t)$ and
$\sigma_x^2>0$, respectively. Their ratio is $\beta_0$; under the null it is zero.
\end{proof}

\begin{intuition}
The estimated slope is built from two averages: an average measuring how $x$ and $y$ move
together, and an average measuring how much $x$ moves on its own. If both averages settle down
to the right numbers, then so does their ratio, and that is the whole proof.

Notice how little is being asked. Nothing here rules out serial correlation; the data may be
strongly dependent from one period to the next. Dependence makes the averages settle down more
slowly, so you need a longer sample to get close, but it does not stop them settling. This is
exactly what fails for random walks: their averages never settle at all, which is why the
random-walk case is a different problem in kind rather than a more severe version of the same
one.

The practical message is a narrow one, and it is easy to over-read. Stationarity protects the
\emph{number} you report as your estimate. It says nothing whatsoever about the standard error
printed next to it, which is the subject of the rest of this section.
\end{intuition}

\begin{proposition}[Slope and conventional $t$-ratio]\label{prop:tstat}
Under Assumptions \ref{as:stat}--\ref{as:clt} and $H_0$,
$$
\sqrt{T}\hat\beta\Rightarrow N\!\left(0,\frac{\Omega}{\sigma_x^4}\right)
$$
and the conventional OLS $t$-ratio satisfies
$$
t_{OLS}\Rightarrow N(0,\lambda), \qquad
\lambda=\frac{\Omega}{\sigma_x^2\sigma_u^2}
=\frac{\LRV\{(x_t-\mu_x)u_t\}}{\Var(x_t)\Var(u_t)}.
$$
\end{proposition}

\begin{proof}
Demeaning changes the scaled score numerator by $o_p(1)$, while
$T^{-1}\sum_t(x_t-\bar x)^2\xrightarrow{p}\sigma_x^2$. Assumption \ref{as:clt} and Slutsky's
theorem give the slope limit. The conventional standard error obeys
$$
T\,\mathrm{se}_{OLS}(\hat\beta)^2\xrightarrow{p}\frac{\sigma_u^2}{\sigma_x^2},
$$
which yields the stated $t$-ratio limit.
\end{proof}

For a two-sided nominal 5% test,
$$
\Pr(|t_{OLS}|>1.96)\longrightarrow
2\Phi\!\left(-\frac{1.96}{\sqrt\lambda}\right).
$$
Therefore $\lambda>1$ implies asymptotic over-rejection, $\lambda<1$ implies asymptotic
under-rejection, and $\lambda=1$ implies correct asymptotic size.

An equivalent and often more useful statement: the *size-adjusted* two-sided critical value for
the conventional statistic converges to $1.96\sqrt\lambda$. Section \ref{sec:power} verifies
this directly in simulation, which is a sharper test of Proposition \ref{prop:tstat} than any
tail rejection probability.

\begin{logic}
A $t$-statistic is an estimate divided by an estimate of its own spread. Proposition
\ref{prop:tstat} works out each piece separately and then compares them.

The \emph{true} spread of the slope estimate is governed by $\Omega$, the spread of the summed
score. The \emph{assumed} spread, the one the conventional formula computes, is
$\sigma_x^2\sigma_u^2$ --- the variance of the regressor multiplied by the variance of the
error, as if the two could be handled separately and as if time-ordering were irrelevant.

If those two quantities agree, the $t$-statistic has the standard normal distribution we all
assume it has. If they disagree, the statistic is still normal and still centred at zero, but
its spread is wrong by the factor $\lambda$, the ratio of truth to assumption. Everything else
in this handout is an investigation of when that ratio is not one.
\end{logic}

\begin{intuition}
The conventional standard error assumes that every observation is a fresh piece of information.
When the data are dependent, that is false, and the formula misprices the evidence.

Suppose consecutive observations tend to repeat each other. Then a sample of 200 does not carry
200 independent pieces of information --- perhaps it carries the equivalent of 50. The
conventional formula still divides as though there were 200, producing a standard error that is
too small, a $t$-statistic that is too large, and rejections far more often than 5\% of the
time. In this reading $\lambda$ is a penalty for dependence, and $T/\lambda$ is the
\emph{effective} sample size: with $\lambda=4$, a sample of 200 is worth about 50.

The reverse is equally possible and much less well known. If consecutive observations tend to
offset each other, the sample carries \emph{more} information than its length suggests, the
conventional standard error is too large, and the test rejects too rarely. Such a test is not
"safe" --- it will miss relationships that are really there. Section \ref{sec:conventional}
reports a case that rejects essentially never.

So $\lambda>1$ means "you have less evidence than you think", $\lambda<1$ means "you have more",
and $\lambda=1$ means the conventional formula happens to be right. Note that a $t$-statistic
of 2.5 is not evidence of anything by itself: with $\lambda=4$ the honest cutoff is
$1.96\times2=3.92$, so the same number that looks convincing under one $\lambda$ is
unremarkable under another.
\end{intuition}

## What makes $\lambda$ differ from one {#sec:decomposition}

The following decomposition is the organising device for everything that follows. It is
elementary, but it prevents a mistake that is easy to make.

\begin{lemma}[Contemporaneous and serial components]\label{lem:decomp}
Under Assumptions \ref{as:stat}--\ref{as:clt},
$$
\lambda=\underbrace{\frac{\E[(x_t-\mu_x)^2u_t^2]}{\sigma_x^2\sigma_u^2}}_{\displaystyle \kappa}
\;+\;\underbrace{\frac{2}{\sigma_x^2\sigma_u^2}\sum_{k=1}^{\infty}\Cov(w_t,w_{t+k})}_{\displaystyle \text{serial component}} .
$$
\end{lemma}

\begin{proof}
$\E w_t=0$ by Assumption \ref{as:stat}, so $\Cov(w_t,w_t)=\E[(x_t-\mu_x)^2u_t^2]$. Split
$\Omega=\sum_k\Cov(w_t,w_{t+k})$ into its lag-zero term and the symmetric sum over $k\ne0$,
then divide by $\sigma_x^2\sigma_u^2$.
\end{proof}

The quantity $\kappa$ measures contemporaneous dependence between the squared regressor
deviation and the squared error: $\kappa=1$ when $\E[(x_t-\mu_x)^2u_t^2]$ factorises, and
$\kappa\ne1$ under conditional heteroskedasticity. The serial component measures the
autocovariance of the score.

\begin{logic}
$\Omega$ is a sum of terms, one for each possible gap between two time periods. The lemma does
nothing more than split that sum into the single term with no gap and everything else, then
divide through so that each part is expressed relative to what the conventional formula
assumes.
\end{logic}

\begin{intuition}
This is the most useful thing in the handout, so it is worth stating in words.

There are exactly two ways the conventional standard error can be wrong, and they have nothing
to do with each other.

The first is a failure \emph{at a single point in time}, measured by $\kappa$. The conventional
formula multiplies the typical size of the regressor by the typical size of the error, treating
them as unrelated. Suppose instead that periods with unusually large $x$ tend to be the same
periods with unusually large errors. Then the products $x_tu_t$ have occasional very large
values, the sum of them is far more volatile than the formula expects, and $\kappa$ exceeds
one. Nothing about the ordering of the data is involved --- shuffle the observations into random
order and this problem survives untouched.

The second is a failure \emph{across time}, measured by the serial component. Here each product
$x_tu_t$ may be perfectly well behaved on its own, but consecutive products tend to have the
same sign, so they accumulate instead of cancelling. Shuffle the data and this problem
disappears entirely.

The mistake this handout is written to prevent is checking only the second. A researcher who
tests for serial correlation, finds none, and concludes the conventional standard error is safe
has checked one of the two terms and ignored the other. Section \ref{sec:volatility} gives a
process where the serial term is exactly zero and the test still rejects two and a half times
too often.

A useful mental image: $\kappa$ asks whether the trouble is visible in a scatter plot, and the
serial term asks whether it is visible in a time plot. You have to look at both.
\end{intuition}

\begin{remark}[The condition for correct size]\label{rem:size}
Correct asymptotic size requires $\lambda=1$, that is, $\Omega=\sigma_x^2\sigma_u^2$. This
needs \emph{both} components to be accounted for. In particular:
\begin{itemize}
\item A serially uncorrelated score is \textbf{not} sufficient. If the serial component
      vanishes then $\lambda=\kappa$, and $\kappa=1$ only under score conditional
      homoskedasticity. Section \ref{sec:volatility} constructs a process whose score is
      exactly serially uncorrelated with $\kappa=e^{v_h}>1$.
\item A serially correlated score is \textbf{not} fatal. Positive and negative autocovariances
      can cancel, leaving $\lambda=1$ with both series visibly autocorrelated. Section
      \ref{sec:lambda-one} constructs such a case and Section \ref{sec:conventional} confirms
      by simulation that it is correctly sized.
\end{itemize}
A sufficient condition for $\lambda=1$ is that $w_t$ be serially uncorrelated \emph{and}
$u_t$ be independent of $x_t$ --- for example, $u_t$ i.i.d. and independent of the regressor.
\end{remark}

\begin{remark}[Exact Gaussian benchmark]\label{rem:exact}
If the regression errors are i.i.d. Gaussian and independent of the complete regressor matrix,
the usual $t$-statistic has an exact Student-$t$ distribution conditional on the regressors.
The i.i.d. Gaussian Monte Carlo DGP below satisfies these stronger conditions. Without them,
$\lambda=1$ is an asymptotic statement and does not imply exact finite-sample size.
\end{remark}

## Independent AR(1) processes with unequal coefficients

\begin{proposition}[General AR(1) example]\label{prop:ar1}
Let
$$
x_t=\phi_xx_{t-1}+\varepsilon_t, \qquad
y_t=\phi_yy_{t-1}+\eta_t,
$$
where $|\phi_x|<1$, $|\phi_y|<1$, and $\{\varepsilon_t\}$ and $\{\eta_t\}$ are mutually
independent i.i.d. sequences with zero mean and finite fourth moments. Then $\kappa=1$ and
$$
\lambda=\frac{1+\phi_x\phi_y}{1-\phi_x\phi_y}.
$$
\end{proposition}

\begin{proof}
Independence of the two processes gives
$$
\Cov(w_t,w_{t+k})
=\gamma_x(k)\gamma_y(k)
=(\phi_x\phi_y)^{|k|}\sigma_x^2\sigma_y^2 .
$$
At $k=0$ this is $\sigma_x^2\sigma_y^2$, so $\kappa=1$ by Lemma \ref{lem:decomp}. Summing over
all lags,
$$
\Omega=\sigma_x^2\sigma_y^2
\left\{1+2\sum_{k=1}^{\infty}(\phi_x\phi_y)^k\right\}
=\sigma_x^2\sigma_y^2\frac{1+\phi_x\phi_y}{1-\phi_x\phi_y}. \qedhere
$$
\end{proof}

Note that Gaussianity is not required: mutual independence of the two innovation *processes*
plus finite fourth moments is enough. Equal positive coefficients reproduce the familiar
over-rejection result. Unequal positive coefficients show that the relevant quantity is their
product. Opposite signs make that product negative and can cause substantial under-rejection,
despite serial correlation in both series.

\begin{intuition}[why the \emph{product} of the two coefficients is what matters]
The score is the product $x_ty_t$. For that product to carry information from one period to the
next, \emph{both} series must carry information from one period to the next. If either series
forgets its past immediately, the product forgets too.

This has a consequence worth memorising: a regression whose regressor is serially uncorrelated
is safe no matter how badly autocorrelated the error is. Set $\phi_x=0$ and $\lambda=1$ for
every value of $\phi_y$. Textbook warnings about autocorrelated errors are really warnings about
autocorrelated errors \emph{together with} a persistent regressor.

The sign works the same way. If both series are positively autocorrelated, then a large $x$ this
period tends to be followed by a large $x$ next period, and likewise for $y$, so consecutive
products line up and reinforce one another: $\lambda$ rises and the test over-rejects. If one
series is positively and the other negatively autocorrelated, consecutive products tend to
\emph{flip} sign, so they cancel; with $(0.8,-0.8)$ the cancellation overshoots so far that
$\lambda$ falls to about $0.22$ and the test almost never rejects.

The numbers escalate sharply. Moving from $\phi=0.5$ to $\phi=0.95$ in both series raises
$\lambda$ from $1.7$ to $19.5$, which means an effective sample size roughly twenty times
smaller than the nominal one. Persistence of $0.95$ is entirely ordinary in macroeconomic and
financial data.
\end{intuition}

## Independent linear processes

The AR(1) calculation extends naturally. Let
$$
x_t=\sum_{j=0}^{\infty}a_j\varepsilon_{t-j}, \qquad
y_t=\sum_{j=0}^{\infty}b_j\eta_{t-j},
$$
where the coefficient sequences are absolutely summable and the two innovation sequences are
mutually independent. If $\gamma_x(k)$ and $\gamma_y(k)$ denote their autocovariances and
$\rho_x(k),\rho_y(k)$ the corresponding autocorrelations, then independence implies $\kappa=1$
and
$$
\Omega=\sum_{k=-\infty}^{\infty}\gamma_x(k)\gamma_y(k),
\qquad
\lambda=\sum_{k=-\infty}^{\infty}\rho_x(k)\rho_y(k).
$$
This formula covers short-memory ARMA processes, unequal lag structures, seasonal dependence,
and negative autocorrelation. It also shows why the *sign pattern* of the two autocorrelation
functions, rather than persistence in either series separately, governs conventional size.

\begin{intuition}
Each series has an autocorrelation function: a list of numbers recording how strongly it
resembles itself one period back, two periods back, and so on. The formula says: line the two
lists up side by side, multiply them lag by lag, and add everything up. That total is $\lambda$.

Read this way, $\lambda$ measures the \emph{alignment} of the two autocorrelation functions,
not the size of either. Two series can each be strongly autocorrelated and still produce
$\lambda=1$, provided their patterns fail to line up --- where one has a positive correlation
the other has a negative one, and the contributions cancel. Equally, two mildly autocorrelated
series that happen to share the same pattern will produce $\lambda>1$.

So the question to ask about a regression is not "are my series persistent?" but "do my series
resemble themselves over time \emph{in the same way}?". Section \ref{sec:lambda-one} builds a
pair that deliberately fails to line up, and the conventional test on that pair is exactly
correct.
\end{intuition}

## A correctly sized test with serially correlated data {#sec:lambda-one}

The formula above makes the "not necessary" half of Remark \ref{rem:size} constructive. Take
$x_t$ to be AR(1) with coefficient $\phi$, so $\rho_x(k)=\phi^{|k|}$, and $y_t$ to be an
invertible MA(2),
$$
y_t=\eta_t+\theta_1\eta_{t-1}+\theta_2\eta_{t-2},
$$
whose autocorrelations vanish beyond lag two. Then
$$
\lambda=1+2\left\{\phi\rho_y(1)+\phi^2\rho_y(2)\right\},
$$
so $\lambda=1$ exactly whenever $\rho_y(1)+\phi\rho_y(2)=0$, which reduces to
$$
\theta_1(1+\theta_2)+\phi\theta_2=0
\qquad\Longleftrightarrow\qquad
\theta_1=-\frac{\phi\theta_2}{1+\theta_2}.
$$
The simulation uses $\phi=`r lambda1_phi`$ and $\theta_2=`r lambda1_theta2`$, hence
$\theta_1=`r sprintf("%.4f", lambda1_theta1)`$. The implied autocorrelations are
$\rho_x(1)=`r sprintf("%.3f", lambda1_phi)`$, $\rho_y(1)=`r sprintf("%.3f", ma2_acf(lambda1_theta1, lambda1_theta2, 1))`$
and $\rho_y(2)=`r sprintf("%.3f", ma2_acf(lambda1_theta1, lambda1_theta2, 2))`$: both series are
plainly serially correlated, and the MA roots have modulus $\sqrt2>1$ so the process is
invertible. Yet the two contributions cancel exactly and the conventional test is correctly
sized. This is the cleanest available demonstration that serial correlation is not the
diagnostic --- the sign pattern is.

\begin{intuition}
We are building $y$ so that its autocorrelation pattern deliberately fails to line up with
$x$'s.

The regressor $x$ is a simple persistent series: it resembles itself with correlation $0.5$ one
period back and $0.25$ two periods back, both positive. We then choose $y$ so that it resembles
itself \emph{negatively} one period back and \emph{positively} two periods back. When the two
lists are multiplied lag by lag, the negative contribution at lag one and the positive
contribution at lag two are equal and opposite, and they cancel to nothing.

The result is a pair of series that would fail any test for serial correlation you ran on them
individually --- a Durbin--Watson statistic on either would look alarming --- yet the ordinary
$t$-test on the regression between them is exactly right, with no correction of any kind
needed. A student who had learned "autocorrelation means your standard errors are wrong" would
apply a correction here and make the answer worse rather than better.

The cancellation is of course exact only because we engineered it. Real data will not oblige.
The point is not that this configuration is common but that it is possible, which is enough to
show that the presence of serial correlation cannot by itself tell you what to do.
\end{intuition}

## Dependence without a contemporaneous projection {#sec:volatility}

Full independence is stronger than the null $\beta_0=0$. For example, let
$$
x_t=s_t\varepsilon_t, \qquad y_t=s_t\eta_t,
$$
where $s_t>0$ is a common persistent stochastic scale and the two innovations are mutually
independent, serially independent, and independent of $s_t$. Then
$\Cov(x_t,y_t)=0$, but the series are dependent through their common volatility. The score
$w_t=s_t^2\varepsilon_t\eta_t$ has $\Cov(w_t,w_{t+k})=0$ for every $k\ne0$, so the serial
component of Lemma \ref{lem:decomp} vanishes identically. If $h_t=\log s_t^2$ is Gaussian with
variance $v_h$, then
$$
\lambda=\kappa=\frac{\E(s_t^4)}{\E(s_t^2)^2}=e^{v_h}>1.
$$
This is the promised counterexample to "serially uncorrelated score implies correct size". The
distortion is entirely contemporaneous heteroskedasticity. An HC standard error is sufficient
asymptotically; HAC is also valid but can be less precise in small samples.

\begin{intuition}
Think of $s_t$ as the general level of turbulence in the market at time $t$. Both series are
calm in calm periods and wild in wild periods, but which \emph{direction} each moves is decided
independently by its own coin flip. So the two variables are genuinely unrelated in the sense
the null cares about --- knowing $x$ tells you nothing about the direction of $y$ --- while
being obviously related in another sense, since knowing that $x$ was wild tells you $y$ was wild
too.

Now consider the products $x_ty_t$ that the standard error is built from. In calm periods they
are all tiny. In wild periods both factors are large at once, so the product is enormous. The
sum is therefore dominated by a handful of turbulent episodes, which makes it far more volatile
than a formula assuming steady conditions would predict. That formula multiplies the average
size of $x$ by the average size of $y$ and never notices that the large values arrive
\emph{together}. Here the mismatch is a factor of about $1.65$, and the test rejects roughly
$12.7\%$ of the time instead of $5\%$.

The decisive feature is what happens if you shuffle the observations into random order. Serial
correlation would be destroyed by shuffling; this problem would not, because it lives entirely
in the pairing of $x$ and $y$ within each period. That is why the correct remedy is HC, which
looks at each period's product on its own, rather than HAC, which is designed to detect
relationships across periods that here do not exist.

The pattern is not exotic. Volatility clustering of exactly this kind is among the most robust
regularities in financial and macroeconomic data.
\end{intuition}

## Near-unit roots and non-uniformity {#sec:nearunit}

Fixed-parameter stationary theory holds $|\phi|<1$ constant as $T$ grows. It need not approximate
sequences such as
$$
\phi_T=1-\frac{c}{T}, \qquad c>0,
$$
which approach the unit-root boundary with the sample size. Local-to-unity asymptotics
[@phillips1987towards; @chanwei1987] are nonstandard: $\hat\beta$ again has a nondegenerate
limit --- now a functional of independent Ornstein--Uhlenbeck processes rather than Brownian
motions --- and the $t$-ratio again diverges at rate $T^{1/2}$, so the rejection probability
tends to one exactly as in the unit-root case. The fixed-$\phi$ plateau formula is therefore
not uniform over such sequences. Consequently, classifying a fitted process as technically
stationary does not guarantee that ordinary stationary or HAC approximations are accurate in
realistic samples. This non-uniformity is the source of well-known difficulties in predictive
regression [@cavanagh1995inference; @elliott1998impact; @stambaugh1999predictive].

\begin{intuition}[what "non-uniform" means, and why it is not a technicality]
Asymptotic theory is a way of producing approximations, not a description of what actually
happens as you collect data. When we say "let $T\to\infty$ with $\phi$ fixed at $0.99$", we are
choosing to approximate our situation by one in which the sample is enormous relative to how
close $\phi$ sits to one. Whether that is a good approximation depends on our actual sample.

For $\phi=0.99$ and $T=200$, it is a poor one. Such a series has barely had time to revert
towards its mean even once, so within the sample it behaves for all practical purposes like a
random walk. The local-to-unity device is a different approximation, built to describe exactly
this situation: it lets $\phi$ approach one as $T$ grows so that the two stay in a realistic
relationship to each other.

"Non-uniform" is the precise statement of the difficulty. For \emph{every} fixed $\phi$ below
one, the ordinary stationary theory is eventually correct --- that much is true and is not in
dispute. But "eventually" arrives later and later as $\phi$ approaches one, and there is no
single sample size that is large enough for all $\phi<1$ at once. So the reassurance "my series
is stationary, therefore stationary theory applies" is empty unless you also ask \emph{how}
stationary, relative to how much data you have.

The practical upshot is that a unit-root test which fails to reject is not a licence to proceed.
An estimated $\hat\phi$ of $0.97$ in 150 observations should be treated as a warning, not a
clearance, whichever side of the critical value it falls.
\end{intuition}

## HAC inference

Using the Bartlett kernel, define
$$
\widehat{\LRV}(w)=\hat\gamma_0+
2\sum_{k=1}^{L_T}\left(1-\frac{k}{L_T+1}\right)\hat\gamma_k,
\qquad
\hat\gamma_k=T^{-1}\sum_{t=k+1}^T\hat w_t\hat w_{t-k},
$$
where $\hat w_t=(x_t-\bar x)\hat u_t$.

\begin{proposition}[HAC validity]\label{prop:hac}
Under Assumptions \ref{as:stat}--\ref{as:hac},
$$
t_{HAC}=\frac{\hat\beta}{\mathrm{se}_{HAC}(\hat\beta)}
\Rightarrow N(0,1).
$$
\end{proposition}

The proof is standard and is omitted [@neweywest1987; @andrews1991heteroskedasticity].

This result is asymptotic, and three caveats matter in practice. First, when persistence is
high a finite sample may contain too little information to estimate the long autocovariance
tail accurately, so bandwidth choice is consequential even though any valid bandwidth sequence
delivers consistency. Second, prewhitening [@andrewsmonahan1992] often improves matters
substantially in exactly the persistent cases where the plain estimator struggles. Third, the
$N(0,1)$ limit ignores the sampling variability of $\widehat{\LRV}$; fixed-$b$ asymptotics
[@kiefervogelsang2005] retains it and delivers larger, better-calibrated critical values. The
small-sample HAC over-rejection documented in Section \ref{sec:hac} is the canonical motivation
for that literature; @lazarus2018har give practical recommendations.

\begin{logic}
Only one piece of the conventional $t$-statistic is broken. The estimate in the numerator is
fine, by Proposition \ref{prop:consistent}. The denominator is wrong because it uses
$\sigma_x^2\sigma_u^2$ where it should use $\Omega$. So replace it: estimate $\Omega$ directly
from the data and divide by that instead. If the estimate is any good, the resulting statistic
has the standard normal distribution the conventional one was supposed to have.
\end{logic}

\begin{intuition}[why the bandwidth is a genuine dilemma]
To estimate $\Omega$ we must estimate how strongly the score resembles itself one period back,
two periods back, and so on --- and then add all of those up. In principle the list is infinite;
in practice we can only use a finite number of terms, and $L_T$ is how many we use.

Both directions fail, which is what makes this awkward. Use too few lags and you miss the tail
of the list, understate $\Omega$, and remain oversized --- you have corrected part of the problem
and declared victory. Use too many and each individual term is estimated from very little data:
the correlation at lag 40 in a sample of 200 rests on 160 overlapping pairs and is mostly noise.
Adding up many noisy numbers gives a noisy total, and a standard error that is itself
unreliable produces unreliable tests. There is no setting that avoids both errors; the bandwidth
rule is a compromise.

This is why HAC is not a switch you flip. When $\phi=0.95$ the score's memory is so long that no
bandwidth a sample of even 5000 can support captures the whole tail, which is exactly what the
HAC comparison in Section \ref{sec:hac} shows: HAC still rejects about twice as often as it
should. It is also why
HAC can be \emph{worse} than doing nothing at very small $T$ --- at $T=50$ even the
independent-data benchmark, where there is nothing whatever to correct, over-rejects under HAC
simply because the correction itself is estimated noisily.
\end{intuition}

## Distributed lags, Granger causality, and omitted dynamics {#sec:dlgranger}

A distributed-lag regression asks whether past values of $x_t$ contain predictive information
for $y_t$. In the dynamic model
$$
y_t=\alpha+\sum_{j=1}^{p}\rho_jy_{t-j}
    +\sum_{j=1}^{q}\beta_jx_{t-j}+e_t,
$$
$x$ does not Granger-cause $y$ relative to the stated information set when
$H_0:\beta_1=\cdots=\beta_q=0$ [@granger1969investigating]. This is a predictive definition,
not a claim about structural causality.

Correct specification matters as much as the covariance estimator. If relevant lags of $y_t$
are omitted, their persistence enters the regression error. Persistent $x$ lags multiplied by
that serially correlated error produce a dependent score vector, so the conventional joint
Wald test can falsely detect predictive content.

It is worth being precise about *which* failure this is, because two distinct problems are
easily conflated. Omitting a relevant regressor biases the remaining coefficients only if the
omitted variable is correlated with the included ones. In the experiment of Section
\ref{sec:granger} the two series are generated independently, so the omitted $y_{t-1}$ is
orthogonal to $x_{t-1}$ and $x_{t-2}$: the population projection coefficients on the $x$ lags
remain exactly zero, there is no bias, and the entire distortion is a failure of the
conventional covariance estimator. That is why HAC repairs most of it. In applications where
the omitted dynamics *are* correlated with the included regressors, the coefficients are biased
as well, and no covariance estimator can help. Including an adequate dynamic specification
remains the first line of defence; robust covariance estimation is not a substitute for
specifying the conditional mean.

\begin{intuition}[two different failures that look alike in the output]
Leaving a variable out of a regression can go wrong in two quite separate ways, and students
routinely merge them into one.

The first is \emph{bias}: the estimate itself moves to the wrong place. This happens only when
the omitted variable is correlated with the variables you kept, so that the regression credits
the missing variable's influence to whichever included variable resembles it. No standard error
can repair this, because the number being tested is already wrong.

The second is a \emph{mismeasured standard error}: the estimate is fine, but the uncertainty
attached to it is not. This is what happens when you omit something uncorrelated with your
regressors. Its influence goes into the error term, which becomes persistent, and persistent
errors combined with a persistent regressor inflate $\lambda$ exactly as in Section
\ref{sec:decomposition}.

The experiment in Section \ref{sec:granger} is deliberately of the second kind: $x$ and $y$ are
built independently, so dropping $y_{t-1}$ cannot bias anything. Everything you see there ---
a test rejecting a true null about a third of the time --- comes purely from a mismeasured
standard error, which is why HAC repairs most of it. Do not read the experiment as showing that
HAC fixes omitted-variable bias. It does not, and in applications where the omitted dynamics
\emph{are} correlated with your regressors you have both problems and HAC addresses only one.

The useful diagnostic question is therefore: is the thing I left out correlated with the things
I kept? If no, you have a standard-error problem and a robust covariance estimator is the right
tool. If yes, you have a specification problem and no covariance estimator will save you.
\end{intuition}

\begin{theorem}[Summary of the stationary case]\label{thm:answer}
For fixed-parameter, short-memory stationary series satisfying Assumptions
\ref{as:stat}--\ref{as:clt}, OLS is consistent for the population projection slope. Under
$H_0:\beta_0=0$, conventional inference is governed by $\lambda=\kappa+{}$serial component.
The test over-rejects when $\lambda>1$, under-rejects when $\lambda<1$, and has correct
asymptotic size when $\lambda=1$. Under the additional HAC-consistency conditions in
Assumption \ref{as:hac}, HAC inference has correct asymptotic size. Thus stationarity prevents
the random-walk inconsistency considered here, but does not by itself validate conventional OLS
standard errors.
\end{theorem}

\begin{framed}
\noindent\textbf{What to take away.} Stationarity buys you consistency of $\hat\beta$, and
nothing else. Whether the conventional $t$-test is trustworthy is a separate question, answered
by a single number $\lambda$, which has two independent sources of trouble: contemporaneous
heteroskedasticity of the score ($\kappa\ne1$) and serial correlation of the score. Neither
serial correlation in the data nor its absence tells you what $\lambda$ is. Diagnose the score,
not the series.
\end{framed}

# Monte Carlo Design

The simulation uses independent Gaussian innovations and nine DGPs:

1. an i.i.d. Gaussian benchmark;
2. equal positive AR coefficients $(\phi_x,\phi_y)=(0.5,0.5)$, $(0.8,0.8)$,
   $(0.9,0.9)$, and $(0.95,0.95)$;
3. unequal positive coefficients $(0.9,0.5)$;
4. opposite-sign coefficients $(0.8,-0.8)$;
5. the AR/MA pair of Section \ref{sec:lambda-one}, for which $\lambda=1$ exactly despite
   serial correlation in both series; and
6. independent random walks as a nonstationary benchmark.

For every stationary AR(1) process, the pre-sample state is drawn independently from its
invariant distribution $N\{0,1/(1-\phi^2)\}$; the MA(2) process is stationary from its first
observation. Every retained observation is therefore drawn from the stationary process. One
trajectory of length `r Tmax` is generated in each replication, and nested prefixes are
evaluated on the grid
$$
T\in\{50,100,150,\ldots,`r Tmax`\}.
$$
Nesting correlates adjacent estimates and produces a smoother comparison across $T$; it does
not eliminate Monte Carlo sampling variation.

The experiment uses $N=`r Nsim_main`$ replications per DGP. The maximum Monte Carlo standard
error of a rejection proportion is $\sqrt{0.25/`r Nsim_main`}\approx`r max_mcse(Nsim_main)`$
percentage points, and reported proportions should be read with that in mind. The conventional
statistic is evaluated over the complete grid. HAC statistics are evaluated at
$T\in\{`r paste(selected_T, collapse=",")`\}$ using a Bartlett kernel and
$L_T=\lfloor2T^{1/3}\rfloor$, which satisfies $L_T\to\infty$ and $L_T/T\to0$. This rate is the
MSE-optimal one for the Bartlett kernel and is deliberately more generous than the
Newey--West rule of thumb $4(T/100)^{2/9}$, because the scores generated by persistent AR(1)
pairs have a long autocovariance tail; Section \ref{sec:hac} shows that even this bandwidth is
not generous enough at $\phi=0.95$.

The standalone script `scripts/spurious_stationary_sim.R` is the single computational source
for the document. It generates all data, figures, and CSV tables, and every numerical value
quoted in the text below is read from its output at render time rather than transcribed by
hand. A single Bartlett estimator, implemented in that script, is used for every HAC number
reported here, so the tables are mutually comparable.

Four supplementary experiments broaden the design:

1. **Common stochastic volatility.** A persistent Gaussian log-volatility with autoregressive
   coefficient 0.95 and stationary variance `r common_volatility_variance` multiplies
   independent Gaussian innovations in both series. The projection slope is zero, but the
   series are dependent and the theoretical variance ratio is
   $e^{`r common_volatility_variance`}\approx`r sprintf("%.2f", common_volatility_lambda)`$.
   Conventional, HC0, and HAC inference are compared using `r Nsim_vol` replications.
2. **Local-to-unity sequences.** At each selected $T$, independent AR(1) series use
   $\phi_T=1-`r local_to_unity_c`/T$. Because the DGP changes with $T$, this is a
   triangular-array experiment rather than a nested fixed-parameter trajectory. The pre-sample
   value is drawn from the invariant distribution, so this is the *stationary-start* design;
   a zero-start design gives materially different finite-sample rejection frequencies, and the
   initialisation is a modelling choice rather than a technicality. Conventional and HAC tests
   use `r Nsim_l2u` replications.
3. **Distributed-lag Granger tests.** Independent AR(1) series with
   $(\phi_x,\phi_y)=(0.8,0.8)$ are used to test jointly whether two lags of $x$ predict $y$.
   The correctly specified regression includes $y_{t-1}$; the misspecified regression omits it.
   Conventional and HAC Wald tests are compared over `r Nsim_gr` replications at
   $T\in\{`r paste(granger_sample_sizes, collapse=",")`\}$.
4. **Size-adjusted local power.** Under $y_t=(b/\sqrt T)x_t+v_t$ with $x$ and $v$ independent
   AR(1) processes with $\phi=`r power_phi`$, conventional and HAC tests are compared at
   $T\in\{`r paste(power_sample_sizes, collapse=",")`\}$ over `r Nsim_pow` replications.
   Critical values are size-adjusted: each test is compared with the empirical 95th percentile
   of its own $|t|$ under $b=0$. Size adjustment is what makes the comparison meaningful, since
   an oversized test mechanically has more raw power and a raw power curve would merely restate
   the size distortion.

# Results

## Conventional inference {#sec:conventional}

@fig-conventional shows conventional rejection frequencies. Equal positive persistence
generates plateaus above 5%, and the plateau rises with $\phi_x\phi_y$. The unequal positive
case $(0.9,0.5)$ also over-rejects, but less severely than $(0.9,0.9)$. The opposite-sign case
$(0.8,-0.8)$ under-rejects sharply and approaches its theoretical rejection probability near
zero. The random-walk rejection frequency continues toward one rather than settling at a
stationary plateau, reaching `r rp("I(1)", 5000)` at $T=`r 5000`$.

Most instructive is the AR/MA design. Both of its series are serially correlated --- the AR
component has $\rho_x(1)=`r sprintf("%.2f", lambda1_phi)`$ --- yet its rejection frequency sits
on the nominal line, at `r rp("AR/MA lambda=1", 5000)` at $T=5000$. Serial correlation alone is
therefore not a diagnosis; it is the sign pattern of the two autocorrelation functions, through
$\lambda$, that matters.

```{r}
#| label: fig-conventional
#| fig-cap: "Conventional OLS rejection frequencies. Labels report the two AR coefficients where applicable; 'AR/MA lambda=1' is the AR(1)/MA(2) design for which lambda equals one exactly. The dashed line is the nominal 5% level. Note the logarithmic sample-size axis."
#| fig-width: 6.0
#| fig-height: 4.4
#| out-width: 100%
p1
```

```{r tbl-conventional}
# Build the wide table keyed on dgp only: lambda and asymptotic are NA for the
# I(1) row, and NA values in reshape's idvar are unreliable.
show_T <- c(50, 500, 5000)
conv_wide <- reshape(
  subset(comparison_table, T %in% show_T, select = c("dgp", "T", "reject")),
  idvar = "dgp", timevar = "T", direction = "wide"
)
conv_wide <- conv_wide[match(dgp_levels, as.character(conv_wide$dgp)), ]
bench <- dgp_spec[match(dgp_levels, dgp_spec$dgp), ]
fmt_lambda <- ifelse(is.na(bench$lambda), "---", sprintf("%.2f", bench$lambda))
fmt_asy <- ifelse(
  is.na(bench$asymptotic), "---",
  ifelse(bench$asymptotic < 0.001,
         sprintf("%.1e", bench$asymptotic),
         sprintf("%.3f", bench$asymptotic)))
conv_wide <- data.frame(
  DGP = dgp_levels, lambda = fmt_lambda, Asymptotic = fmt_asy,
  conv_wide[, paste0("reject.", show_T), drop = FALSE],
  check.names = FALSE, stringsAsFactors = FALSE
)
names(conv_wide) <- c("DGP", "$\\lambda$", "Asymptotic",
                      "$T=50$", "$T=500$", "$T=5000$")
knitr::kable(conv_wide, booktabs = TRUE, row.names = FALSE, escape = FALSE,
      align = c("l", "r", "r", "r", "r", "r"),
      digits = 3,
      caption = paste0("Conventional rejection frequencies and stationary asymptotic ",
                       "benchmarks. Entries are proportions; the maximum Monte Carlo ",
                       "standard error is ", max_mcse(Nsim_main),
                       " percentage points. The I(1) row has no stationary benchmark."))
```

@tbl-conventional confirms the theoretical classification. At $T=5000$, the conventional
rejection rates are `r rp("(0.5,0.5)", 5000)` for $(0.5,0.5)$, `r rp("(0.8,0.8)", 5000)` for
$(0.8,0.8)$, `r rp("(0.9,0.9)", 5000)` for $(0.9,0.9)$, and `r rp("(0.95,0.95)", 5000)` for
$(0.95,0.95)$, against asymptotic benchmarks of `r asy_of("(0.5,0.5)")`,
`r asy_of("(0.8,0.8)")`, `r asy_of("(0.9,0.9)")` and `r asy_of("(0.95,0.95)")` respectively.
The unequal positive case $(0.9,0.5)$ rejects `r rp("(0.9,0.5)", 5000)` of the time. For
$(0.8,-0.8)$ the theoretical rejection probability is `r asy_of("(0.8,-0.8)", 4)`, so among
`r Nsim_main` replications one expects essentially none; the simulated frequency at $T=5000$ is
`r rp("(0.8,-0.8)", 5000)`. The i.i.d. and AR/MA frequencies fluctuate around the exact 5%
benchmark rather than equalling 5% mechanically at every sample size.

## HAC inference {#sec:hac}

@fig-hac compares conventional and HAC rejection frequencies for the stationary DGPs. At
moderate and large $T$, HAC inference moves rejection rates toward 5% in every serially
correlated case, including both the over-rejecting and the under-rejecting ones. At $T=50$,
however, the correction is dominated by long-run variance estimation noise and can be worse
than no correction at all: for $(0.5,0.5)$ the conventional test rejects
`r rp("(0.5,0.5)", 50)` while HAC rejects `r rp("(0.5,0.5)", 50, "HAC")`, and even the i.i.d.
benchmark, where there is nothing to correct, rejects `r rp("iid", 50, "HAC")` under HAC.
Convergence is slowest for $(0.95,0.95)$, whose HAC rejection rate is still
`r rp("(0.95,0.95)", 5000, "HAC")` at $T=5000$. This is a finite-sample bandwidth and long-run
variance estimation problem, not evidence against Proposition \ref{prop:hac}.

```{r}
#| label: fig-hac
#| fig-cap: "Conventional and Bartlett-HAC rejection frequencies for the stationary DGPs at selected sample sizes. The dotted line is the nominal 5% level. The bandwidth is floor(2 T^(1/3))."
#| fig-width: 6.0
#| fig-height: 4.6
#| out-width: 100%
p2
```

```{r tbl-hac}
hac_end <- subset(hac_comparison_table, T %in% c(50, 5000),
                  select = c(dgp, T, reject_conventional, reject_hac))
hac_end <- reshape(hac_end, idvar = "dgp", timevar = "T", direction = "wide")
hac_end <- hac_end[match(setdiff(dgp_levels, "I(1)"), hac_end$dgp), ]
names(hac_end) <- c("DGP", "OLS $T=50$", "HAC $T=50$",
                    "OLS $T=5000$", "HAC $T=5000$")
knitr::kable(hac_end, booktabs = TRUE, row.names = FALSE, escape = FALSE,
      align = c("l", "r", "r", "r", "r"), digits = 3,
      caption = paste0("Conventional and HAC rejection frequencies at the endpoints. ",
                       "Entries are proportions; maximum Monte Carlo standard error ",
                       max_mcse(Nsim_main), " percentage points."))
```

The HAC results caution against describing robust inference as an automatic finite-sample cure.
At small $T$, estimating a long-run variance is itself noisy. By $T=5000$ the HAC frequencies
are close to 5% for every stationary DGP except the most persistent equal-coefficient case.
Prewhitening or fixed-$b$ critical values would be the natural next step for that case.

## Size-adjusted local power {#sec:power}

Robust inference is often described as trading size for power. @fig-power quantifies the trade
for the $(0.8,0.8)$ design, and the answer is milder than the folklore suggests. Once each test
is compared with its own size-adjusted critical value, the two power curves are close: at
$T=`r power_sample_sizes[1]`$ and $b=4$ the conventional test has size-adjusted power
`r pw("Conventional OLS", power_sample_sizes[1], 4)` against
`r pw("HAC", power_sample_sizes[1], 4)` for HAC, and at $T=`r power_sample_sizes[2]`$ the
figures are `r pw("Conventional OLS", power_sample_sizes[2], 4)` and
`r pw("HAC", power_sample_sizes[2], 4)`. The cost of HAC, properly measured, is the modest loss
from estimating the long-run variance rather than knowing it --- not the dramatic sacrifice
that raw power comparisons suggest.

The size-adjusted critical values are themselves worth reporting, because they test
Proposition \ref{prop:tstat} directly. For the $(0.8,0.8)$ design, $\lambda=`r sprintf("%.3f", power_lambda)`$,
so the conventional two-sided 5% critical value should converge to
$1.96\sqrt\lambda=`r sprintf("%.2f", qnorm(0.975) * sqrt(power_lambda))`$ rather than to 1.96.
The simulated values are `r pw_cv("Conventional OLS", power_sample_sizes[1])` at
$T=`r power_sample_sizes[1]`$ and `r pw_cv("Conventional OLS", power_sample_sizes[2])` at
$T=`r power_sample_sizes[2]`$, and they do not shrink with $T$ --- the distortion is a feature
of the limit distribution, not a small-sample artefact. The corresponding HAC critical values,
`r pw_cv("HAC", power_sample_sizes[1])` and `r pw_cv("HAC", power_sample_sizes[2])`, do decline
toward 1.96, as Proposition \ref{prop:hac} requires.

```{r}
#| label: fig-power
#| fig-cap: "Size-adjusted local power against the alternative slope b / sqrt(T), for independent AR(1) series with phi = 0.8. Each test uses its own empirical 95th percentile under b = 0 as critical value, so all curves pass through 5% at b = 0 by construction."
#| fig-width: 6.0
#| fig-height: 3.8
#| out-width: 100%
p6
```

## Dependent zero-projection series and common volatility

@fig-common-volatility relaxes full independence. Although $x_t$ and $y_t$ share the same
stochastic scale, their population projection slope is zero because their innovations are
independent, and their score is exactly serially uncorrelated. The conventional rejection
frequency reaches `r cv_rej("Conventional OLS", 5000)` at $T=5000$, against the
`r pc(common_volatility_asymptotic)` asymptotic value implied by
$\lambda=\kappa=e^{`r common_volatility_variance`}$. HC0 rejects `r cv_rej("HC0", 5000)` and
HAC `r cv_rej("HAC", 5000)` at the same sample size. This is the counterexample of Remark
\ref{rem:size} in action: a size distortion generated entirely by contemporaneous
second-moment dependence, with no serial autocovariance in the score whatsoever. HC suffices;
HAC also works but pays for estimating autocovariances that are zero.

```{r}
#| label: fig-common-volatility
#| fig-cap: "Rejection frequencies for dependent zero-projection series sharing a persistent stochastic volatility. The dotted line is the nominal 5% level."
#| fig-width: 5.6
#| fig-height: 3.8
#| out-width: 90%
p3
```

## Local-to-unity sequences

@fig-local-unity shows that fixed-parameter stationary conclusions are not uniform near the
unit-root boundary. As $T$ rises, $\phi_T=1-`r local_to_unity_c`/T$ moves closer to one.
Conventional rejection rises from `r l2u_rej("Conventional OLS", 50)` at $T=50$ to
`r l2u_rej("Conventional OLS", 5000)` at $T=5000$ rather than approaching a fixed stationary
plateau, and HAC deteriorates along the same sequence, from `r l2u_rej("HAC", 50)` to
`r l2u_rej("HAC", 5000)`. Each finite-$T$ DGP is stationary, yet the sequence behaves
increasingly like the unit-root benchmark: at $T=5000$ the local-to-unity conventional
rejection frequency of `r l2u_rej("Conventional OLS", 5000)` is close to the I(1) benchmark's
`r rp("I(1)", 5000)`. As noted in Section \ref{sec:nearunit}, the theoretical rejection
probability tends to
one along this sequence, exactly as under a unit root. A unit-root or near-unit-root diagnostic
is therefore complementary to robust covariance estimation, not substitutable for it.

```{r}
#| label: fig-local-unity
#| fig-cap: "Rejection frequencies for local-to-unity AR(1) sequences with phi_T = 1 - c/T. Each finite-sample DGP is stationary, but the parameter approaches unity with T."
#| fig-width: 5.6
#| fig-height: 3.8
#| out-width: 90%
p4
```

## Distributed-lag and Granger-causality tests {#sec:granger}

The null in @fig-granger is that two lags of $x$ jointly have no predictive content for $y$.
Because the two AR processes are generated independently, this null is true --- and, as
explained in Section \ref{sec:dlgranger}, it is true in *both* specifications, since the
omitted $y_{t-1}$ is
orthogonal to the included $x$ lags. There is no coefficient bias anywhere in this experiment.

When the model includes $y_{t-1}$, the conventional joint Wald rejection frequency is
`r gr_rej("Correct dynamic model", "Conventional OLS", 100)` at $T=100$ and
`r gr_rej("Correct dynamic model", "Conventional OLS", 2000)` at $T=2000$. HAC is more distorted
in the smallest samples, at `r gr_rej("Correct dynamic model", "HAC", 100)` at $T=100$, but
declines toward nominal size.

When $y_{t-1}$ is omitted, the conventional test rejects
`r gr_rej("Omit lagged y", "Conventional OLS", 100)` at $T=100$ and
`r gr_rej("Omit lagged y", "Conventional OLS", 2000)` at $T=2000$, creating persistent false
evidence of Granger causality that does not diminish with sample size. HAC reduces this to
`r gr_rej("Omit lagged y", "HAC", 2000)` at $T=2000$. Readers should check whether that figure
has settled or is still declining across the last two sample sizes in the plot: convergence to
nominal size is guaranteed asymptotically but can be slow when the induced error is as
persistent as it is here.

The comparison separates two remedies. Including $y$'s dynamics correctly specifies the
conditional mean; HAC addresses the covariance of the remaining score. Here only the second
problem is present, and HAC accordingly does most of the work. In applications where the
omitted dynamics are correlated with the regressors, the first problem is present too and HAC
cannot help. Applied work generally needs both.

```{r}
#| label: fig-granger
#| fig-cap: "Joint Wald tests of no Granger causality from x to y. The correct dynamic model includes lagged y; the misspecified model omits it. The dotted line is the nominal 5% level."
#| fig-width: 6.0
#| fig-height: 3.8
#| out-width: 100%
p5
```

```{r tbl-extensions}
pick <- function(df, value_col, ...) {
  conds <- list(...)
  keep <- rep(TRUE, nrow(df))
  for (nm in names(conds)) keep <- keep & df[[nm]] == conds[[nm]]
  v <- df[[value_col]][keep]
  if (length(v) != 1L) stop("tbl-extensions: ambiguous selection")
  v
}
extension_table <- data.frame(
  Experiment = c("Common volatility, $T=5000$",
                 "Local-to-unity, $T=5000$",
                 "Granger, correct dynamics, $T=2000$",
                 "Granger, omitted dynamics, $T=2000$"),
  Conventional = c(
    pick(common_volatility_res, "reject", method = "Conventional OLS", T = 5000),
    pick(local_to_unity_res,   "reject", method = "Conventional OLS", T = 5000),
    pick(granger_res, "reject", method = "Conventional OLS",
         specification = "Correct dynamic model", T = 2000),
    pick(granger_res, "reject", method = "Conventional OLS",
         specification = "Omit lagged y", T = 2000)),
  HC0 = c(pick(common_volatility_res, "reject", method = "HC0", T = 5000),
          NA, NA, NA),
  HAC = c(
    pick(common_volatility_res, "reject", method = "HAC", T = 5000),
    pick(local_to_unity_res,   "reject", method = "HAC", T = 5000),
    pick(granger_res, "reject", method = "HAC",
         specification = "Correct dynamic model", T = 2000),
    pick(granger_res, "reject", method = "HAC",
         specification = "Omit lagged y", T = 2000)),
  check.names = FALSE
)
old_na <- getOption("knitr.kable.NA")
options(knitr.kable.NA = "---")
knitr::kable(extension_table, booktabs = TRUE, row.names = FALSE, escape = FALSE,
             digits = 3, align = c("l", "r", "r", "r"),
             caption = "Selected rejection frequencies from the scope-extension experiments. HC0 is reported only where contemporaneous heteroskedasticity is the operative distortion.")
options(knitr.kable.NA = old_na)
```

# Discussion and Conclusion

The theory and simulation together support a more precise answer than the assertion that
spurious regression either disappears or necessarily persists under stationarity.

First, joint short-memory stationarity and projection orthogonality make OLS consistent for the
population projection slope under the stated assumptions. Full independence is unnecessary.
The common-volatility experiment shows that dependent series can have a zero slope while still
invalidating the conventional homoskedastic standard error. This sharply distinguishes the
stationary setup from the independent random-walk regression, in which the slope has a
nondegenerate limit and the conventional $t$-ratio diverges.

Second, consistency of the coefficient does not validate its conventional standard error. The
limiting conventional test depends on $\lambda$, which by Lemma \ref{lem:decomp} has two
independent sources: a contemporaneous component $\kappa$ and a serial component. Correct
asymptotic size requires the two together to sum to one. Neither the presence nor the absence
of serial correlation settles the matter, as the AR/MA design and the common-volatility design
demonstrate from opposite directions. The equal-positive AR(1) case is economically important,
but it is an example rather than a universal law.

Third, increasing the sample size removes transient finite-sample effects but does not eliminate
a non-nominal conventional plateau. The size-adjusted critical values in Section
\ref{sec:power} make this concrete: they converge to $1.96\sqrt\lambda$, not to 1.96, and they
do not shrink as $T$ grows.

Fourth, HAC inference is the appropriate asymptotic remedy when its additional consistency
conditions hold. The simulation confirms movement toward nominal size and shows that the power
cost, correctly measured against size-adjusted critical values, is modest. It also shows that
high persistence makes convergence slow and bandwidth choice consequential, and that at very
small $T$ the correction can be worse than no correction. Prewhitening and fixed-$b$ critical
values are the standard responses.

Fifth, fixed-parameter stationary asymptotics can be misleading near the unit-root boundary.
The local-to-unity experiment produces rejection probabilities tending toward one even though
every finite-sample autoregressive coefficient is below one. Stationarity is therefore not a
uniform guarantee over empirically plausible parameter sequences.

Finally, robust inference cannot replace dynamic specification. Distributed-lag and
Granger-causality tests require an information set rich enough to capture the dependent
variable's own dynamics. Omitting those dynamics leaves serial dependence in the error and can
create persistent false predictive relationships. In the design used here that failure is
purely one of covariance estimation and HAC largely repairs it; when the omitted dynamics are
correlated with the regressors, the coefficients are biased too and no covariance estimator can
help.

The practical conclusion is conditional but clear: short-memory stationarity and projection
orthogonality protect slope consistency, not conventional inference. Researchers should
diagnose dependence and heteroskedasticity in the regression score, assess proximity to the
unit-root boundary, specify relevant dynamics, and use a defensible covariance estimator rather
than assume that either stationarity or a large sample makes the usual OLS $t$-ratio or
Granger-causality Wald test reliable.

# Exercises {.unnumbered}

\begin{exercise}[The product, not the persistence]
Two independent AR(1) series have $\phi_x=0.9$ and $\phi_y=-0.9$. Compute $\lambda$ and the
asymptotic rejection probability of a nominal 5\% two-sided $t$-test. Compare with
$\phi_y=+0.9$. Explain in one sentence why persistence in each series separately is not the
relevant statistic.
\end{exercise}

\begin{exercise}[Constructing correct size]
Using $\lambda=\sum_k\rho_x(k)\rho_y(k)$, find a value of $\theta_2$ such that an AR(1)
regressor with $\phi=0.7$ paired with an invertible MA(2) yields $\lambda=1$. Verify
invertibility. Then modify \texttt{lambda1\_phi} and \texttt{lambda1\_theta2} in the
simulation script and confirm the rejection frequency numerically.
\end{exercise}

\begin{exercise}[Where does the distortion come from?]
For the common-volatility design of Section \ref{sec:volatility}, verify that
$\Cov(w_t,w_{t+k})=0$ for all $k\neq0$, and that $\kappa=\E(s_t^4)/\{\E(s_t^2)\}^2$. Explain
why HC0 is sufficient here and what would change if the innovations $\varepsilon_t$ were
themselves serially correlated.
\end{exercise}

\begin{exercise}[Reading a size-adjusted critical value]
Section \ref{sec:power} reports a size-adjusted conventional critical value near
$1.96\sqrt\lambda$ that does not shrink with $T$. Explain why this is a sharper check on
Proposition \ref{prop:tstat} than the rejection frequency, and what you would expect to see if
Proposition \ref{prop:tstat} were false.
\end{exercise}

\begin{exercise}[Bandwidth sensitivity]
Replace \texttt{hac\_bandwidth} in the simulation script with the Newey--West rule of thumb
$\lfloor4(T/100)^{2/9}\rfloor$ and re-run the HAC comparison. At $T=5000$ this changes the
bandwidth from 34 to 9. Predict, before running, whether the $(0.95,0.95)$ distortion will get
better or worse, and explain the result you obtain.
\end{exercise}

\begin{exercise}[Diagnosis in practice]
You estimate a regression on quarterly data with $T=120$ and obtain a $t$-statistic of 2.6.
List the diagnostics you would run, in order, before reporting this as evidence of a
relationship. For each, state what result would change your conclusion.
\end{exercise}

# Bibliography {.unnumbered}

The list below includes both the works cited in the text and general background reading.
Students meeting this material for the first time will find @hamilton1994time the most useful
single reference for the underlying time-series asymptotics, and @engle1982autoregressive the
origin of the conditional-heteroskedasticity models that motivate the common-volatility design
of Section \ref{sec:volatility}.

::: {#refs}
:::

# Computational Appendix {.unnumbered}

The complete, executable simulation is maintained in
`scripts/spurious_stationary_sim.R`. It depends only on `ggplot2` and `scales`; the Bartlett
HAC estimator used throughout is implemented in the script itself so that a single estimator
produces every robust number reported above. Running

```{.bash}
Rscript scripts/spurious_stationary_sim.R
```

recreates all figures, the CSV result files, and a `sessionInfo.txt` recording the package
versions used. Setting `SPURIOUS_QUICK=1` in the environment runs a fast, noisier pass suitable
for drafting; `SPURIOUS_OUTPUT_DIR` redirects the output location. Each experiment is seeded
independently, so changing the replication count of one does not silently alter the results of
another. The Quarto document sources that same script, so the handout and the standalone
computation share one implementation, and every number in the text is read from the simulation
output at render time.
