---
title: "**Time Series Econometrics**"
subtitle: "**Properties of Stationary and Integrated Time Series**"
author: "**Eliab Luvanda**"
date: today
lang: en

format:
  pdf:
    pdf-engine: lualatex
    mainfont: Palatino
    documentclass: article
    classoption:
      - twoside
    fontsize: 11pt
    papersize: a4
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    lof: true

execute:
  echo: false
  warning: false
  message: false


geometry:
  - inner=1.3in
  - outer=1.0in
  - top=1.1in
  - bottom=1.1in
  - headsep=0.3in
header-includes:
  - \usepackage{amsmath, amssymb, amsthm, mathtools}
  - \usepackage{bm}
  - \usepackage{bbm}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{setspace}
  - \onehalfspacing
  - |
    \theoremstyle{plain}
    \newtheorem{theorem}{Theorem}[section]
    \newtheorem{lemma}[theorem]{Lemma}
    \newtheorem{proposition}[theorem]{Proposition}
    \newtheorem{corollary}[theorem]{Corollary}
  - |
    \theoremstyle{definition}
    \newtheorem{definition}[theorem]{Definition}
    \newtheorem{example}[theorem]{Example}
    \newtheorem{exercise}[theorem]{Exercise}
  - |
    \theoremstyle{remark}
    \newtheorem{remark}[theorem]{Remark}
  - \renewcommand{\qedsymbol}{$\blacksquare$}
  - |
    \newcommand{\E}{\mathbb{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\Cov}{\mathrm{Cov}}
    \newcommand{\plim}{\mathrm{plim}}
    \newcommand{\R}{\mathbb{R}}
    \newcommand{\1}{\mathbbm{1}}
  - \usepackage{titlesec}
  - |
    \titleformat{\section}
      {\normalfont\large\bfseries}{\thesection}{1em}{}
  - |
    \titleformat{\subsection}
      {\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
  - \usepackage{fancyhdr}
  - |
    \pagestyle{fancy}
    \fancyhf{}
    \fancyhead[LE,RO]{\thepage}
    \fancyhead[LO]{Stationary and Integrated Processes}
    \fancyhead[RE]{Stationary and Integrated Processes}
  - \usepackage{hyperref}
  - |
    \hypersetup{
      colorlinks=true,
      linkcolor=blue,
      citecolor=blue,
      urlcolor=blue
    }
---
## Introduction


Time series econometrics is a branch of econometrics that focuses on the analysis of data that is collected over time. Time series econometrics is distinct from conventional econometrics in that it explicitly accounts for the temporal dependence and dynamic structure of the data. For this reason, time series econometrics is often regarded as a specialized subfield of econometrics, with its own unique set of tools and techniques for analyzing time series data. The reason for this distinction is that time series data often exhibit properties such as autocorrelation, non-stationarity, and seasonality, which renders the application of the standard laws of large numbers and the central limit theorem  inapplicable without appropriate modifications. As a result, time series econometrics has developed a rich theoretical framework and a wide array of methodologies to address these challenges, making it an essential area of study for economists and statisticians working with temporal data.


The rationale for this distinction lies in the fact that time series data frequently display characteristics such as autocorrelation, non-stationarity, and seasonality. These inherent features preclude the direct application of the classical Law of Large Numbers and the Central Limit Theorem, unless appropriate methodological adjustments are implemented to account for such dependencies and structural patterns.  Consequently, time series econometrics has evolved into a highly sophisticated discipline characterized by a robust theoretical foundation and an extensive repertoire of methodological approaches designed to address these complexities, thereby establishing itself as an indispensable field of inquiry for researchers engaged in the analysis of temporal data.




```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#| caption: "Simulated Stationary Time Series"
#| fig.cap: "A simulated stationary time series"
#| fig.width: 4
#| fig.height: 3
library(ggplot2)
set.seed(123)
n <- 100
library(ggplot2)  # Load required library
library(ggpubr)  # Load ggpubr for theme_classic2

# Function to generate a stationary time series
# y_t = c + \phi *y_{t-1} + \epsilon_t
ar1_time_series <- function(n, c, phi) {
  # Generate white noise
  epsilon <- rnorm(n, mean = 0, sd = 1)
  
  # Initialize the time series
  y <- numeric(n)
  y[1] <- c + epsilon[1]  # Initial value
  
  # Generate the time series
  for (t in 2:n) {
    y[t] <- c + phi * y[t - 1] + epsilon[t]
  }  # Added missing closing brace
  
  return(y)
}  # Fixed brace placement

# Define n before using it
n <- 200  # Add this line
c <- 0.2  # Constant term
phi <- 0.6  # Autoregressive coefficient

# Generate an AR(1) time series
data <- data.frame(
  time = 1:n,
  time_series = ar1_time_series(n, c, phi)
)

# Plot the time series using ggplot2
ggplot(data, aes(x = time, y = time_series)) +
  geom_line(color = "navy") +
  labs(
    x = "Time",
    y = "Value"
  ) +
  theme_classic2() +
  theme(plot.title = element_text(hjust = 0.5))

```

\begin{definition}
A time series is stationary if its mean, variance, and autocovariance do not change over time.
\end{definition}

\begin{example}
Consider a time series \( Y_t \) defined as follows:
\[Y_t = \mu + \epsilon_t \]
where \( \mu \) is a constant and \( \epsilon_t \) is white noise. This time series is stationary because its mean is constant (\( \mu \)), its variance is constant, and its autocovariance does not depend on time.
\end{example}

### Unit Roots
A unit root is a characteristic of a time series that indicates non-stationarity. A time series has a unit root if it can be represented as an autoregressive process with a root equal to one. The presence of a unit root implies that shocks to the time series have permanent effects, making it non-stationary.

\begin{definition}
A time series has a unit root if it can be represented as an autoregressive process with a root equal to one.
\end{definition}

\begin{remark}
The presence of a unit root in a time series indicates that the series is non-stationary and that shocks to the series have permanent effects.
\end{remark}

## Example: Visualizing Weak Stationarity

The figure below illustrates a simulated weakly stationary AR(1) process, which is a common example of a stationary time series. The plot demonstrates the constancy of the mean (indicated by the red dashed line) and the stable variance over time, visually confirming the properties of weak stationarity.

- Constancy of mean
- Stable variance


```{r, fig.cap="A simulated stationary AR(1) process is illustrated. The mean (red dashed line) is shown to be constant, and the variance remains stable over time. This stability is further depicted by fluctuations being contained within a nearly uniform band", fig.width=5, fig.height=3}
# Simulate a stationary AR(1) process
#| echo: false
set.seed(123)
n <- 500
phi <- 0.7
mu <- 5
sigma <- 1
x <- numeric(n)
x[1] <- rnorm(1, mean = mu, sd = sigma / sqrt(1 - phi^2))
for (t in 2:n) {
  x[t] <- mu + phi * (x[t-1] - mu) + rnorm(1, 0, sigma)
}
library(ggplot2)

mean_x <- mean(x)
sd_x <- sd(x)
upper_band <- mean_x + 2 * sd_x
lower_band <- mean_x - 2 * sd_x

ggplot(data.frame(Time = 1:n, Value = x), aes(Time, Value)) +
  geom_line(color = 'steelblue', linewidth = 1) +
  geom_hline(yintercept = mean_x, linetype = 'dashed', color = 'red', linewidth = 1) +
  geom_hline(yintercept = upper_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  geom_hline(yintercept = lower_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  labs(y = expression(X(t)), x = 'Time') +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  scale_x_continuous(breaks = seq(0, n, by = 50)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```


\begin{remark}
The concept of weak stationarity is often contrasted with strong stationarity (or strict stationarity), which requires that the entire joint distribution of the process be invariant under time shifts, not just the first two moments. While strong stationarity is a more stringent condition, weak stationarity is often sufficient for many practical applications in econometrics and time series analysis, particularly when the focus is on linear models and second-order properties.
\end{remark}

## Strong Stationarity


A strong (or strictly) stationary stochastic process is defined as a process whose complete probabilistic structure remains invariant under temporal shifts. In other words, for any displacement of the time index—whether forward or backward—the joint distribution of any finite collection of observations remains unchanged. This property implies that the statistical behavior of the process is entirely independent of the specific time at which it is observed. In contrast to weak stationarity, which requires only the constancy of the mean, variance, and autocovariance structure over time, strong stationarity imposes the more stringent condition that all finite-dimensional distributions are invariant with respect to time shifts.

In other words, the process is characterized not merely by stability in its first two moments, but by invariance in its entire distributional structure, thereby ensuring that its stochastic behavior remains fully time-invariant. This stronger condition establishes a rigorous theoretical framework for modeling processes in which the complete probability distribution is preserved over time, and thus constitutes a fundamental concept in probability theory and stochastic modeling.

\begin{definition}[strong stationarity]
Let $ (\Omega, \mathcal{F}, \mathbb{P}) $ be a probability space, and let $ \{X_t : t \in T\} $ be a stochastic process indexed by $ T \subseteq \mathbb{R} $. The process $ \{X_t\} $ is said to be strongly stationary (or strictly stationary) if, for every integer $ n \geq 1 $, for every choice of time points $ t_1, t_2, \ldots, t_n \in T $, and for every shift $ h \in \mathbb{R} $ such that $ t_i + h \in T $, we have
$ (X_{t_1}, X_{t_2}, \ldots, X_{t_n}) \overset{d}{=} (X_{t_1+h}, X_{t_2+h}, \ldots, X_{t_n+h}), $
where $ \overset{d}{=} $ denotes equality in distribution.
\end{definition}


**Key Points**

• Entire distribution invariance: Not just the mean, variance, or covariance, but the full joint distribution of any finite collection of random variables is unchanged under time shifts.

• Stronger than weak stationarity: Weak stationarity only requires invariance of first and second moments, while strong stationarity requires invariance of all finite-dimensional distributions.

• Implication: Every strong stationary process is weakly stationary (if moments exist), but not every weakly stationary process is strongly stationary.

The underlying rationale of strong stationarity is that a stochastic process exhibits identical behavior regardless of the temporal point at which observation begins. Rather than merely requiring the stability of moments such as the mean, variance, or covariance, strong stationarity necessitates that the complete joint distribution of any set of observations remains invariant under temporal shifts.

Intuitively, this implies that the randomness governing the process is entirely invariant with respect to time. Specifically, if one observes the process at time points $t_1, t_2, \ldots, t_n$ and subsequently shifts this collection of observations forward by an arbitrary increment $h$, the resulting joint probability distribution remains unchanged. This definition therefore characterizes the strongest form of temporal consistency, ensuring that the process is not merely stable in terms of its moments, but is fundamentally identical in its full distributional structure under any temporal translation.

\begin{example}
A simple example of a strongly stationary process is i.i.d. white noise. Suppose $\{X_t\}$ is a sequence of independent and identically distributed random variables, each with distribution $F$. Because the variables are identically distributed, the joint distribution of any finite collection $ (X_{t_1}, X_{t_2}, \ldots, X_{t_n}) $ is the same as that of $ (X_{t_1+h}, X_{t_2+h}, \ldots, X_{t_n+h}) $ for any time shift $h$. Independence ensures that shifting the time indices does not change the joint law, and identical distribution ensures that each variable has the same marginal distribution. Thus, i.i.d. white noise is strictly stationary. 
\end{example}

#```{r}
#| echo: true
#| include: true
#| warning: false
#| message: false
#| caption: "Gaussian white noise: strong stationarity example"
#| fig.cap: "Gaussian white noise (i.i.d. N(0,1)) — an example of a strongly stationary stochastic process"
#| fig.width: 5
#| fig.height: 3
set.seed(42)
n <- 200
wn <- rnorm(n, mean = 0, sd = 1)
library(ggplot2)
ggplot(data.frame(Time = 1:n, Value = wn), aes(x = Time, y = Value)) +
  geom_line(color = 'black') +
  labs(x = 'Time', y = expression(epsilon[t]), title = 'Gaussian White Noise (i.i.d. N(0,1))') +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

\begin{example}
 Another simple example is a Bernoulli process, where each $X_t$ is an independent Bernoulli random variable with the same success probability $p$. Again, the joint distribution is invariant under time shifts, making it a strongly stationary process. 
\end{example}

These examples highlight the intuition: when randomness is identically distributed across time and unaffected by shifts, the process satisfies strong stationarity.

A classic example of a process that is weakly stationary but not strongly stationary is a Gaussian process with time-invariant mean and covariance compared to a non-Gaussian process with the same covariance structure. To make it concrete, consider the following:
\begin{example}
Let $\{X_t\}$ be a sequence of random variables defined as
$$X_t = Z_t \cdot Y,$$
 
where $\{Z_t\}$ is i.i.d. white noise with mean zero and variance one, and $Y$ is an independent random variable taking values $\pm 1$ with equal probability.
Why weakly stationary?
The mean of  $X_t$ is zero for all $t$, the variance is constant, and the covariance between $X_s$ and $X_t$ depends only on the lag because the underlying $\{Z_t\}$ is i.i.d. Thus, the process satisfies weak stationarity.
Why not strongly stationary?
The joint distribution of $X_{t_1}, X_{t_2}, \ldots, X_{t_n}$ depends on the random sign $Y$. For example, the distribution of $(X_1, X_2)$ is not the same as that of $(-X_1, -X_2)$, even though their means and covariances are identical. This shows that the finite-dimensional distributions are not invariant under time shifts, so the process fails strong stationarity.
\end{example}

In short: weak stationarity only cares about mean, variance, and covariance stability, while strong stationarity requires full distributional invariance. This example illustrates how a process can meet the weaker conditions without satisfying the stronger ones.
