---
title: "**Time Series Econometrics**"
subtitle: "**Properties of Stationary and Integrated Time Series**"
author: "**Eliab Luvanda**"
date: today
lang: en
bibliography: references.bib
abstract: |
  This handout reviews properties of stationary and integrated time
  series, covering definitions of stationarity, examples, and visual
  illustrations. It is intended as a concise reference for students and
  researchers in time series econometrics.
keywords:
  - time series
  - stationarity
  - econometrics

format:
  pdf:
    pdf-engine: lualatex
    mainfont: Palatino
    documentclass: article
    classoption:
      - twoside
    fontsize: 11pt
    papersize: a4
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    lof: true

execute:
  echo: false
  warning: false
  message: false


geometry:
  - inner=1.3in
  - outer=1.0in
  - top=1.1in
  - bottom=1.1in
  - headsep=0.3in
header-includes:
  - \usepackage{amsmath, amssymb, amsthm, mathtools}
  - \usepackage{bm}
  - \usepackage{bbm}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{setspace}
  - \onehalfspacing
  - |
    \theoremstyle{plain}
    \newtheorem{theorem}{Theorem}[section]
    \newtheorem{lemma}[theorem]{Lemma}
    \newtheorem{proposition}[theorem]{Proposition}
    \newtheorem{corollary}[theorem]{Corollary}
  - |
    \theoremstyle{definition}
    \newtheorem{definition}[theorem]{Definition}
    \newtheorem{example}[theorem]{Example}
    \newtheorem{exercise}[theorem]{Exercise}
  - |
    \theoremstyle{remark}
    \newtheorem{remark}[theorem]{Remark}
  - \renewcommand{\qedsymbol}{$\blacksquare$}
  - |
    \newcommand{\E}{\mathbb{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\Cov}{\mathrm{Cov}}
    \newcommand{\plim}{\mathrm{plim}}
    \newcommand{\R}{\mathbb{R}}
    \newcommand{\1}{\mathbbm{1}}
  - |
  
  - \usepackage{titlesec}
  - |
    \titleformat{\section}
      {\normalfont\large\bfseries}{\thesection}{1em}{}
  - |
    \titleformat{\subsection}
      {\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
  - \usepackage{fancyhdr}
  - |
    \pagestyle{fancy}
    \fancyhf{}
    \fancyhead[LE,RO]{\thepage}
    \fancyhead[LO]{Stationary and Integrated Processes}
    \fancyhead[RE]{Stationary and Integrated Processes}
  - \usepackage{hyperref}
  - |
    \hypersetup{
      colorlinks=true,
      linkcolor=blue,
      citecolor=blue,
      urlcolor=blue
    }
---
## Introduction


Time series econometrics is a branch of econometrics that focuses on the analysis of data collected over time. It is distinct from conventional econometrics because it explicitly accounts for temporal dependence and the dynamic structure of data. Time series data often exhibit properties such as autocorrelation, non‑stationarity, and seasonality, which can render the standard laws of large numbers and the central limit theorem inapplicable without appropriate modifications. As a result, time series econometrics has developed a rich theoretical framework and a wide array of methodologies to address these challenges, making it an essential area of study for economists and statisticians working with temporal data.

This distinction arises because time series data frequently display features such as autocorrelation, non‑stationarity, and seasonality.

The distinction between stationary time series and integrated time series is fundamental in time series econometrics. Understanding the properties of these types of series is crucial for selecting appropriate modeling techniques and for making valid inferences about the underlying data-generating processes. The underlying concept central to understanding stationary and integrated time series is the notion of a stochastic process. Building on foundational principles of probability theory, a stochastic process can be defined as follows:

\begin{definition}
Let $(\Omega, \mathcal{F}, \mathbb{P})$ be a probability space, and let $T$ be an index set (often representing time, e.g. $T = \mathbb{N}$ for discrete time or $T = \mathbb{R}_{+}$ for continuous time).

A stochastic process is a family of random variables
$$\{X(\tau, \omega) : \tau \in T, \; \omega \in \Omega\}$$

such that for each fixed $\tau \in T$, the mapping
$$X(\tau, \cdot) : \Omega \to \mathbb{R}$$

is a random variable on $(\Omega, \mathcal{F}, \mathbb{P})$.
\end{definition}


Alternatively, a stochastic process can be defined as follows:

\begin{definition}
A stochastic process can equivalently be viewed as a function
$$X: \Omega \times T \to \mathbb{R},\qquad (\omega,\tau) \mapsto X(\omega,\tau),$$
where:

- For each $\tau \in T$, the mapping $X(\cdot,\tau):\Omega\to\mathbb{R}$ is measurable with respect to $\mathcal{F}$.

- For each $\omega\in\Omega$, the mapping $X(\omega,\cdot):T\to\mathbb{R}$ is a trajectory (sample path) of the process.
\end{definition}



 **Key Perspectives**

• Random variable view: For each $\tau \in T$, $X(\tau)$ is a random variable.

• Sample path view: For each $\omega \in \Omega$, $X(\omega, \cdot)$ is a deterministic function of time.

• Distributional view: The finite-dimensional distributions $X(\tau_1), \ldots, X(\tau_n)$ for $\tau_1, \ldots, \tau_n \in T$ characterize the process.


Figure 1 illustrates the concept of a stochastic process through the simulation of multiple sample paths of a simple random walk, a canonical example within stochastic process theory. Each trajectory corresponds to a distinct realization of the process, associated with a particular outcome $\omega \in \Omega$. The figure depicts the temporal evolution of the process values across these realizations, thereby exemplifying the intrinsic randomness and variability that are fundamental characteristics of stochastic processes.

```{r, fig.cap="Multiple simple random-walk sample paths (5)", fig.width=6, fig.height=3, echo=FALSE}
# Simulate and plot 5 simple random walks (T = 500)
set.seed(2026)
T <- 500
n_paths <- 5

# Generate iid normal increments and form cumulative sums (simple random walks)
increments <- matrix(rnorm(T * n_paths, mean = 0, sd = 1), ncol = n_paths)
paths <- apply(increments, 2, cumsum)

# Prepare long dataframe for ggplot
df_paths <- data.frame(
  Time = rep(1:T, times = n_paths),
  Value = as.vector(paths),
  Omega = factor(rep(1:n_paths, each = T))
)

library(ggplot2)

# Use showtext to ensure the glyphs (e.g. ω) are available in R graphics
if (!requireNamespace("showtext", quietly = TRUE)) {
  install.packages("showtext", repos = "https://cloud.r-project.org")
}
library(showtext)
# Register Google font 'Noto Sans' (covers Greek letters) and enable showtext
try({
  font_add_google("Noto Sans", "notosans")
}, silent = TRUE)
showtext_auto()

ggplot(df_paths, aes(x = Time, y = Value, color = Omega, group = Omega)) +
  geom_line(size = 0.6) +
  labs(x = 'Time', y = expression(X[t](omega)), color = expression(omega)) +
  theme_bw() +
  theme(text = element_text(family = "notosans"),
        legend.position = 'right',
        plot.title = element_text(hjust = 0.5)) +
    scale_color_discrete(labels = parse(text = paste0("omega[", 1:n_paths, "]"))) +
    scale_x_continuous(breaks = seq(1, T, by = 50))
```

A stochastic process may be most appropriately understood as a mathematical formalization of the concept of “random evolution over time.” Its definition is grounded in a probability space $(\Omega, \mathcal{F}, \mathbb{P})$, which characterizes the set of all possible outcomes and provides the underlying measure-theoretic structure for randomness. For each time point $\tau$ in the index set $T$, the process assigns a random variable $X(\tau, \cdot)$. Consequently, at every instant in time, there exists a well-defined random quantity associated with the process.

See the formal definition above; Figure 1 shows sample paths (realizations) for different outcomes $\\omega \\in \\Omega$.

Conversely, fixing an outcome $\\omega$ yields a deterministic function $X(\\omega, \\cdot)$, called a *sample path*, which describes the process for that realization. This dual perspective (random variables indexed by time vs.
deterministic trajectories indexed by outcomes) is central to the theory.

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#| caption: "Simulated Stationary Time Series"
#| fig.cap: "A simulated stationary time series"
#| fig.width: 4
#| fig.height: 3
library(ggplot2)
set.seed(123)
n <- 100
library(ggplot2)  # Load required library
library(ggpubr)  # Load ggpubr for theme_classic2

# Function to generate a stationary time series
# y_t = c + \phi *y_{t-1} + \epsilon_t
ar1_time_series <- function(n, c, phi) {
  # Generate white noise
  epsilon <- rnorm(n, mean = 0, sd = 1)
  
  # Initialize the time series
  y <- numeric(n)
  y[1] <- c + epsilon[1]  # Initial value
  
  # Generate the time series
  for (t in 2:n) {
    y[t] <- c + phi * y[t - 1] + epsilon[t]
  }  # Added missing closing brace
  
  return(y)
}  # Fixed brace placement

# Define n before using it
n <- 200  # Add this line
c <- 0.2  # Constant term
phi <- 0.6  # Autoregressive coefficient

# Generate an AR(1) time series
data <- data.frame(
  time = 1:n,
  time_series = ar1_time_series(n, c, phi)
)

# Plot the time series using ggplot2
ggplot(data, aes(x = time, y = time_series)) +
  geom_line(color = "navy") +
  labs(
    x = "Time",
    y = "Value"
  ) +
  theme_classic2() +
  theme(plot.title = element_text(hjust = 0.5))

```


## Example: Visualizing Weak Stationarity

<!-- -->

The figure below illustrates a simulated weakly stationary AR(1) process, which is a common example of a stationary time series. The plot demonstrates the constancy of the mean (indicated by the red dashed line) and the stable variance over time, visually confirming the properties of weak stationarity.

- Constancy of mean
- Stable variance


```{r, fig.cap="Simulated stationary AR(1) process (mean dashed)", fig.width=5, fig.height=3}
# Simulate a stationary AR(1) process
#| echo: false
set.seed(123)
n <- 500
phi <- 0.8
mu <- 5
sigma <- 1
x <- numeric(n)
x[1] <- rnorm(1, mean = mu, sd = sigma / sqrt(1 - phi^2))
for (t in 2:n) {
  x[t] <- mu + phi * (x[t-1] - mu) + rnorm(1, 0, sigma)
}
library(ggplot2)

mean_x <- mean(x)
sd_x <- sd(x)
upper_band <- mean_x + 2 * sd_x
lower_band <- mean_x - 2 * sd_x

ggplot(data.frame(Time = 1:n, Value = x), aes(Time, Value)) +
  geom_line(color = 'black', linewidth = 0.6) +
  geom_hline(yintercept = mean_x, linetype = 'dashed', color = 'red', linewidth = 1) +
  geom_hline(yintercept = upper_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  geom_hline(yintercept = lower_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  labs(y = expression(X(t)), x = 'Time') +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  scale_x_continuous(breaks = seq(0, n, by = 50)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```


\begin{remark}
The concept of weak stationarity is often contrasted with strong stationarity (or strict stationarity), which requires that the entire joint distribution of the process be invariant under time shifts, not just the first two moments. While strong stationarity is a more stringent condition, weak stationarity is often sufficient for many practical applications in econometrics and time series analysis, particularly when the focus is on linear models and second-order properties.
\end{remark}

## Strong Stationarity

A strong (or strictly) stationary stochastic process is defined as a process whose complete probabilistic structure remains invariant under temporal shifts. For any displacement of the time index—whether forward or backward—the joint distribution of any finite collection of observations remains unchanged. This property implies that the statistical behavior of the process is independent of the specific time at which it is observed. In contrast to weak stationarity, which requires only the constancy of the mean, variance, and autocovariance structure over time, strong stationarity imposes the more stringent condition that all finite-dimensional distributions are invariant with respect to time shifts.

Thus, the process is characterized not merely by stability in its first two moments, but by invariance in its entire distributional structure, ensuring that its stochastic behavior remains fully time-invariant. This stronger condition establishes a rigorous theoretical framework for modeling processes in which the complete probability distribution is preserved over time, and thus constitutes a fundamental concept in probability theory and stochastic modeling.

\begin{definition}[strong stationarity]
Let $ (\Omega, \mathcal{F}, \mathbb{P}) $ be a probability space, and let $ \{X_t : t \in T\} $ be a stochastic process indexed by $ T \subseteq \mathbb{R} $. The process $ \{X_t\} $ is said to be strongly stationary (or strictly stationary) if, for every integer $ n \geq 1 $, for every choice of time points $ t_1, t_2, \ldots, t_n \in T $, and for every shift $ h \in \mathbb{R} $ such that $ t_i + h \in T $, we have
$ (X_{t_1}, X_{t_2}, \ldots, X_{t_n}) \overset{d}{=} (X_{t_1+h}, X_{t_2+h}, \ldots, X_{t_n+h}), $
where $ \overset{d}{=} $ denotes equality in distribution.
\end{definition}


**Key Points**

• Entire distribution invariance: Not just the mean, variance, or covariance, but the full joint distribution of any finite collection of random variables is unchanged under time shifts.

• Stronger than weak stationarity: Weak stationarity only requires invariance of first and second moments, while strong stationarity requires invariance of all finite-dimensional distributions.

• Implication: Every strong stationary process is weakly stationary (if moments exist), but not every weakly stationary process is strongly stationary.

The underlying rationale of strong stationarity is that a stochastic process exhibits identical behavior regardless of the temporal point at which observation begins. Rather than merely requiring the stability of moments such as the mean, variance, or covariance, strong stationarity necessitates that the complete joint distribution of any set of observations remains invariant under temporal shifts.

Intuitively, this implies that the randomness governing the process is entirely invariant with respect to time. Specifically, if one observes the process at time points $t_1, t_2, \ldots, t_n$ and subsequently shifts this collection of observations forward by an arbitrary increment $h$, the resulting joint probability distribution remains unchanged. This definition therefore characterizes the strongest form of temporal consistency, ensuring that the process is not merely stable in terms of its moments, but is fundamentally identical in its full distributional structure under any temporal translation.

\begin{example}
An everyday example is i.i.d. white noise: independence plus identical marginals imply that every finite-dimensional joint distribution is invariant under time shifts, so the process is strictly stationary.
\end{example}

```{r white_noise, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.cap="Gaussian white noise (i.i.d. N(0,1))", fig.width=5, fig.height=3}
set.seed(42)
n <- 500
wn <- rnorm(n, mean = 0, sd = 1)
library(ggplot2)

# compute summary lines
mean_wn <- mean(wn)
sd_wn <- sd(wn)
upper <- mean_wn + 2 * sd_wn
lower <- mean_wn - 2 * sd_wn

ggplot(data.frame(Time = 1:n, Value = wn), aes(x = Time, y = Value)) +
  geom_line(color = 'black') +
  geom_hline(yintercept = mean_wn, linetype = 'dashed', color = 'red', linewidth = 0.6) +
  geom_hline(yintercept = upper, linetype = 'dotted', color = 'darkgreen', linewidth = 0.5) +
  geom_hline(yintercept = lower, linetype = 'dotted', color = 'darkgreen', linewidth = 0.5) +
  labs(x = 'Time', y = expression(epsilon[t])) +
  scale_x_continuous(breaks = seq(0, n, by = 50)) +
  scale_y_continuous(breaks = seq(floor(min(wn)), ceiling(max(wn)), by = 1)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

\begin{example}
The Bernoulli process (i.i.d. Bernoulli($p$)) is likewise strictly stationary.
\end{example}

These examples highlight the intuition: when randomness is identically distributed across time and unaffected by shifts, the process satisfies strong stationarity.

A classic example of a process that is weakly stationary but not strongly stationary is a Gaussian process with time-invariant mean and covariance compared to a non-Gaussian process with the same covariance structure. To make it concrete, consider the following:
\begin{example}
Define $X_t=Z_tY$ with $Z_t$ i.i.d. mean-zero noise and $Y\in\{\pm1\}$ independent of the $Z_t$. The process has constant mean and lag-dependent covariance (weak stationarity), but finite-dimensional distributions depend on the random sign $Y$, so it is not strictly stationary.
\end{example}

In short: weak stationarity only cares about mean, variance, and covariance stability, while strong stationarity requires full distributional invariance. This example illustrates how a process can meet the weaker conditions without satisfying the stronger ones.
