---
title: "**Time Series Econometrics**"
subtitle: "**Properties of Stationary and Integrated Time Series**"
author: "**Eliab Luvanda**"
date: today
lang: en
bibliography: references.bib
abstract: |
  This handout reviews properties of stationary and integrated time
  series, covering definitions, examples, and visual illustrations. It
  provides a concise reference for students and researchers in
  time-series econometrics.

keywords:
  - time series
  - stationarity
  - econometrics

format:
  pdf:
    pdf-engine: lualatex
    mainfont: Palatino
    documentclass: article
    classoption:
      - twoside
    fontsize: 11pt
    papersize: a4
    toc: true
    toc-depth: 3
    number-sections: true
    colorlinks: true
    lof: true

execute:
  echo: false
  warning: false
  message: false


geometry:
  - inner=1.3in
  - outer=1.0in
  - top=1.1in
  - bottom=1.1in
  - headsep=0.3in
header-includes:
  - \usepackage{amsmath, amssymb, amsthm, mathtools}
  - \usepackage{bm}
  - \usepackage{bbm}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{setspace}
  - \onehalfspacing
  - |
    \theoremstyle{plain}
    \newtheorem{theorem}{Theorem}[section]
    \newtheorem{lemma}[theorem]{Lemma}
    \newtheorem{proposition}[theorem]{Proposition}
    \newtheorem{corollary}[theorem]{Corollary}
  - |
    \theoremstyle{definition}
    \newtheorem{definition}[theorem]{Definition}
    \newtheorem{example}[theorem]{Example}
    \newtheorem{exercise}[theorem]{Exercise}
  - |
    \theoremstyle{remark}
    \newtheorem{remark}[theorem]{Remark}
  - \renewcommand{\qedsymbol}{$\blacksquare$}
  - |
    \newcommand{\E}{\mathbb{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\Cov}{\mathrm{Cov}}
    \newcommand{\plim}{\mathrm{plim}}
    \newcommand{\R}{\mathbb{R}}
    \newcommand{\1}{\mathbbm{1}}
  - |
  
  - \usepackage{titlesec}
  - |
    \titleformat{\section}
      {\normalfont\large\bfseries}{\thesection}{1em}{}
  - |
    \titleformat{\subsection}
      {\normalfont\normalsize\bfseries}{\thesubsection}{1em}{}
  - \usepackage{fancyhdr}
  - |
    \pagestyle{fancy}
    \fancyhf{}
    \fancyhead[LE,RO]{\thepage}
    \fancyhead[LO]{Stationary and Integrated Processes}
    \fancyhead[RE]{Stationary and Integrated Processes}
  - \usepackage{hyperref}
  - |
    \hypersetup{
      colorlinks=true,
      linkcolor=blue,
      citecolor=blue,
      urlcolor=blue
    }
---

## Introduction

Time-series econometrics analyzes sequences of observations indexed by time and explicitly models temporal dependence. Time-series data commonly exhibit autocorrelation, non-stationarity, and seasonality. As a result, the standard i.i.d. assumptions often fail and require specialized methods. This handout summarizes core definitions and properties of stationary and integrated processes, and gives concise examples and illustrations.

Distinguishing stationary from integrated series is essential for choosing models and making valid inferences. Central to these ideas is the notion of a stochastic process, defined below.

\begin{definition}
Let $(\Omega,\mathcal{F},\mathbb{P})$ be a probability space, and let $T$ be an index set (often representing time, e.g. $T = \mathbb{N}$ for discrete time or $T = \mathbb{R}_{+}$ for continuous time).

A stochastic process is a family of random variables
$$\{X(\tau, \omega) : \tau \in T, \; \omega \in \Omega\}$$

such that for each fixed $\tau \in T$, the mapping
$$X(\tau, \cdot) : \Omega \to \mathbb{R}$$

is a random variable on $(\Omega, \mathcal{F}, \mathbb{P})$.
\end{definition}

Alternatively, we define a stochastic process as follows:

\begin{definition}
One can equivalently view a stochastic process as a function
$$X: \Omega \times T \to \mathbb{R},\qquad (\omega,\tau) \mapsto X(\omega,\tau),$$
where:

- For each $\tau \in T$, the mapping $X(\cdot,\tau):\Omega\to\mathbb{R}$ is measurable with respect to $\mathcal{F}$.

- For each $\omega\in\Omega$, the mapping $X(\omega,\cdot):T\to\mathbb{R}$ is a trajectory (sample path) of the process.
\end{definition}

**Key Perspectives**

• Random variable view: For each $\tau \in T$, $X(\tau)$ is a random variable.

• Sample path view: For each $\omega \in \Omega$, $X(\omega, \cdot)$ is a deterministic function of time.

• Distributional view: The finite-dimensional distributions $X(\tau_1), \ldots, X(\tau_n)$ for $\tau_1, \ldots, \tau_n \in T$ characterize the process.

Figure 1 illustrates the concept of a stochastic process through the simulation of multiple sample paths of a simple random walk, a canonical example within stochastic process theory. Each trajectory corresponds to a distinct realization of the process, associated with a particular outcome $\omega \in \Omega$. The figure depicts the temporal evolution of the process values across these realizations, thereby exemplifying the intrinsic randomness and variability that are fundamental characteristics of stochastic processes.

```{r, fig.cap="Multiple simple random-walk sample paths (5)", fig.width=6, fig.height=3, echo=FALSE}
# Simulate and plot 5 simple random walks (T = 500)
set.seed(2026)
T <- 500
n_paths <- 5

# Generate iid normal increments and form cumulative sums (simple random walks)
increments <- matrix(rnorm(T * n_paths, mean = 0, sd = 1), ncol = n_paths)
paths <- apply(increments, 2, cumsum)

# Prepare long dataframe for ggplot
df_paths <- data.frame(
  Time = rep(1:T, times = n_paths),
  Value = as.vector(paths),
  Omega = factor(rep(1:n_paths, each = T))
)

library(ggplot2)

# Use showtext to ensure glyphs (e.g. ω) are available in R graphics.
# NOTE: install required packages and fonts outside the document to avoid
# interactive installs during rendering. Recommended packages are listed in
# README.md.
library(showtext)
# If you want the Google font 'Noto Sans' available automatically, install
# it interactively (once) with: font_add_google("Noto Sans", "notosans").
showtext_auto()

ggplot(df_paths, aes(x = Time, y = Value, color = Omega, group = Omega)) +
  geom_line(size = 0.6) +
  labs(x = 'Time', y = expression(X[t](omega)), color = expression(omega)) +
  theme_bw() +
  theme(text = element_text(family = "notosans"),
        legend.position = 'right',
        plot.title = element_text(hjust = 0.5)) +
    scale_color_discrete(labels = parse(text = paste0("omega[", 1:n_paths, "]"))) +
    scale_x_continuous(breaks = seq(1, T, by = 50))
```

A stochastic process formalizes "random evolution over time": given a probability space $(\Omega,\mathcal{F},\mathbb{P})$, the process assigns a random variable to each time point and, for a fixed outcome $\omega$, a deterministic sample path. Figure 1 illustrates multiple sample paths (realizations) of a simple random walk. Each trajectory corresponds to a distinct realization. The figure depicts the temporal evolution of the process values across these realizations. It exemplifies the intrinsic randomness and variability that are fundamental characteristics of stochastic processes.

```{r}
#| echo: false
#| include: false
#| warning: false
#| message: false
#| caption: "Simulated Stationary Time Series"
#| fig.cap: "A simulated stationary time series"
#| fig.width: 4
#| fig.height: 3
library(ggplot2)
set.seed(123)
n <- 100
library(ggpubr)  # Load ggpubr for theme_classic2

# Function to generate a stationary time series
# y_t = c + \phi *y_{t-1} + \epsilon_t
ar1_time_series <- function(n, c, phi) {
  # Generate white noise
  epsilon <- rnorm(n, mean = 0, sd = 1)
  
  # Initialize the time series
  y <- numeric(n)
  y[1] <- c + epsilon[1]  # Initial value
  
  # Generate the time series
  for (t in 2:n) {
    y[t] <- c + phi * y[t - 1] + epsilon[t]
  }  # Added missing closing brace
  
  return(y)
}  # Fixed brace placement

# Define n before using it
n <- 200  # Add this line
c <- 0.2  # Constant term
phi <- 0.6  # Autoregressive coefficient

# Generate an AR(1) time series
data <- data.frame(
  time = 1:n,
  time_series = ar1_time_series(n, c, phi)
)

# Plot the time series using ggplot2
ggplot(data, aes(x = time, y = time_series)) +
  geom_line(color = "navy") +
  labs(
    x = "Time",
    y = "Value"
  ) +
  theme_classic2() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Example: Visualizing Weak Stationarity

<!-- -->

The figure below illustrates a simulated weakly stationary AR(1) process, which is a common example of a stationary time series. The plot demonstrates the constancy of the mean (indicated by the red dashed line) and the stable variance over time, visually confirming the properties of weak stationarity.

- Constancy of mean
- Stable variance

```{r, fig.cap="Simulated stationary AR(1) process (mean dashed)", fig.width=5, fig.height=3}
# Simulate a stationary AR(1) process
#| echo: false
set.seed(123)
n <- 500
phi <- 0.8
mu <- 5
sigma <- 1
x <- numeric(n)
x[1] <- rnorm(1, mean = mu, sd = sigma / sqrt(1 - phi^2))
for (t in 2:n) {
  x[t] <- mu + phi * (x[t-1] - mu) + rnorm(1, 0, sigma)
}
library(ggplot2)

mean_x <- mean(x)
sd_x <- sd(x)
upper_band <- mean_x + 2 * sd_x
lower_band <- mean_x - 2 * sd_x

ggplot(data.frame(Time = 1:n, Value = x), aes(Time, Value)) +
  geom_line(color = 'black', linewidth = 0.6) +
  geom_hline(yintercept = mean_x, linetype = 'dashed', color = 'red', linewidth = 1) +
  geom_hline(yintercept = upper_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  geom_hline(yintercept = lower_band, linetype = 'dotted', color = 'darkgreen', linewidth = 0.8) +
  labs(y = expression(X(t)), x = 'Time') +
  theme_bw(base_size = 14) +
  theme(
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8)
  ) +
  scale_x_continuous(breaks = seq(0, n, by = 50)) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 8))
```

\begin{remark}
Weak stationarity requires only constant mean and autocovariance, whereas strong stationarity requires full distributional invariance under time shifts. Although strict stationarity is a stronger condition, weak stationarity is often sufficient in practice, especially for linear models and analyses focused on second-order properties.
\end{remark}

## Strong Stationarity

A strongly (strictly) stationary process has the same finite-dimensional distributions under any time shift. This is a stronger requirement than weak stationarity, which only constrains the first two moments; strict stationarity demands full distributional invariance.

\begin{definition}[strong stationarity]
Let $ (\Omega, \mathcal{F}, \mathbb{P}) $ be a probability space, and let $ \{X_t : t \in T\} $ be a stochastic process indexed by $ T \subseteq \mathbb{R} $. The process $ \{X_t\} $ is said to be strongly stationary (or strictly stationary) if, for every integer $ n \geq 1 $, for every choice of time points $ t_1, t_2, \ldots, t_n \in T $, and for every shift $ h \in \mathbb{R} $ such that $ t_i + h \in T $, we have
$ (X_{t_1}, X_{t_2}, \ldots, X_{t_n}) \overset{d}{=} (X_{t_1+h}, X_{t_2+h}, \ldots, X_{t_n+h}), $
where $ \overset{d}{=} $ denotes equality in distribution.
\end{definition}

**Key Points**

• Entire distribution invariance: Not just the mean, variance, or covariance, but the full joint distribution of any finite collection of random variables remains the same under time shifts.

• Stronger than weak stationarity: Weak stationarity only requires invariance of first and second moments, while strong stationarity requires invariance of all finite-dimensional distributions.

• Implication: Every strong stationary process is weakly stationary (if moments exist), but not every weakly stationary process is strongly stationary.

Intuitively, strict stationarity means that time shifts do not change the joint distribution of observations.

\begin{example}
An everyday example is i.i.d. white noise: independence plus identical marginals imply that every finite-dimensional joint distribution is invariant under time shifts, so the process is strictly stationary.
\end{example}

```{r white_noise, echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.cap="Gaussian white noise (i.i.d. N(0,1))", fig.width=5, fig.height=3}
set.seed(42)
n <- 500
wn <- rnorm(n, mean = 0, sd = 1)
library(ggplot2)

# compute summary lines
mean_wn <- mean(wn)
sd_wn <- sd(wn)
upper <- mean_wn + 2 * sd_wn
lower <- mean_wn - 2 * sd_wn

ggplot(data.frame(Time = 1:n, Value = wn), aes(x = Time, y = Value)) +
  geom_line(color = 'black') +
  geom_hline(yintercept = mean_wn, linetype = 'dashed', color = 'red', linewidth = 0.6) +
  geom_hline(yintercept = upper, linetype = 'dotted', color = 'darkgreen', linewidth = 0.5) +
  geom_hline(yintercept = lower, linetype = 'dotted', color = 'darkgreen', linewidth = 0.5) +
  labs(x = 'Time', y = expression(epsilon[t])) +
  scale_x_continuous(breaks = seq(0, n, by = 50)) +
  scale_y_continuous(breaks = seq(floor(min(wn)), ceiling(max(wn)), by = 1)) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

\begin{example}
The Bernoulli process (i.i.d. Bernoulli($p$)) is likewise strictly stationary.
\end{example}

These examples illustrate that identical marginal distributions and independence across time imply strict stationarity.

A classic example of a process that is weakly stationary but not strongly stationary is a Gaussian process with time-invariant mean and covariance compared to a non-Gaussian process with the same covariance structure. To make it concrete, consider the following:

\begin{example}
Define $X_t=Z_tY$ with $Z_t$ i.i.d. mean-zero noise and $Y\in\{\pm1\}$ independent of the $Z_t$. The process has constant mean and lag-dependent covariance (weak stationarity), but finite-dimensional distributions depend on the random sign $Y$, so it is not strictly stationary.
\end{example}

In short: weak stationarity only cares about mean, variance, and covariance stability, while strong stationarity requires full distributional invariance. This example illustrates how a process can meet the weaker conditions without satisfying the stronger ones.


Table: Summary comparison of weak (second‑order) and strong (strict) stationarity: definitions, requirements, implications, and practical considerations.

| Aspect | Weak (second‑order) stationarity | Strong (strict) stationarity |
|---|---:|---|
|Definition| Mean constant and covariance depends only on lag: $\mathbb{E}[X_t]=\mu$ and $\mathrm{Cov}(X_{t+h},X_t)=\gamma(h)$ for all $t,h$. | All finite‑dimensional distributions are invariant under time shifts: for every $k$, $t_1,\dots,t_k$ and shift $h$, $(X_{t_1+h},\dots,X_{t_k+h}) \stackrel{d}{=} (X_{t_1},\dots,X_{t_k})$. |
|Required moments| Requires existence of first and second moments. | No moment requirement in the definition. |
|Condition on distributions| Only first two moments must be time‑invariant. | Entire joint distributions must be time‑invariant. |
|Implication relation| Does not imply strong stationarity in general. | Implies weak stationarity if first and second moments exist. |
|Gaussian case| For Gaussian processes, weak ⇔ strong (mean & covariance determine law). | Same as left column under Gaussianity. |
|Typical examples| ARMA processes with finite variance. | IID processes; time‑homogeneous Markov processes. |
|Detectability / tests| Tests focus on mean/covariance (ADF, KPSS, sample ACF). | Hard to test directly; compares finite‑dimensional distributions. |
|Practical use| Sufficient for spectral analysis and linear forecasting. | Stronger theoretical guarantees; rarely required in practice. |
|Vulnerabilities| Can hide nonstationary higher moments or non‑Gaussian features. | More robust but harder to verify. |

