---
title: "Finite-Sample and Asymptotic Properties of the Linear GMM Estimator"
subtitle: "Theory and Monte Carlo Evidence"
author: "Eliab Luvanda"
date: today
lang: en
bibliography: references.bib

abstract: |
  We give a self-contained treatment of the linear Generalized Method of
  Moments (GMM) estimator in an overidentified instrumental-variables model.
  We state the identification, consistency, and asymptotic-normality results
  from first principles, derive the efficiency bound attained by the optimal
  weighting matrix, and derive the limiting chi-squared distribution of
  Hansen's $J$ statistic for testing overidentifying restrictions. We then
  contrast this asymptotic theory with finite-sample behaviour using a Monte
  Carlo experiment: an overidentified, conditionally heteroskedastic
  instrumental-variables design in which a one-step (2SLS-weighted) and a
  two-step efficient GMM estimator are compared across sample sizes ranging
  from 50 to 10,000. The simulation traces out consistency (bias and variance
  shrinking with $n$), the efficiency gain of two-step over one-step GMM,
  the approach of the studentized estimator's finite-sample distribution to
  its $N(0,1)$ limit, the coverage of sandwich-based confidence intervals,
  and the size of the $J$ test, and discusses where each departs from its
  asymptotic approximation at small $n$.

format:
  pdf:
    pdf-engine: lualatex
    documentclass: article
    fontsize: 11pt
    papersize: a4
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    fig-pos: "htbp"

geometry:
  - inner=1.2in
  - outer=1.0in
  - top=1.1in
  - bottom=1.2in
  - headsep=0.3in

mainfont: "TeX Gyre Termes"
sansfont: "TeX Gyre Heros"
monofont: "Inconsolata"
mathfont: "TeX Gyre Termes Math"

header-includes:
  - \usepackage{amsmath, amssymb, amsthm, mathtools}
  - \newcommand{\bm}[1]{\symbf{#1}}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{setspace}
  - \singlespacing
  - |
    \theoremstyle{plain}
    \newtheorem{proposition}{Proposition}
    \newtheorem{theorem}[proposition]{Theorem}
    \newtheorem{lemma}[proposition]{Lemma}
    \newtheorem{corollary}[proposition]{Corollary}
  - |
    \theoremstyle{definition}
    \newtheorem{assumption}{Assumption}
    \newtheorem{definition}[assumption]{Definition}
  - \renewcommand{\qedsymbol}{$\blacksquare$}
  - |
    \newcommand{\E}{\mathbb{E}}
    \newcommand{\Var}{\mathrm{Var}}
    \newcommand{\Cov}{\mathrm{Cov}}
    \newcommand{\plim}{\operatorname*{plim}}
    \newcommand{\bX}{\bm{X}}
    \newcommand{\bZ}{\bm{Z}}
    \newcommand{\by}{\bm{y}}
    \newcommand{\bbeta}{\bm{\beta}}
    \newcommand{\beps}{\bm{\varepsilon}}
  - \usepackage{fancyhdr}
  - |
    \pagestyle{fancy}
    \fancyhf{}
    \fancyhead[LE,RO]{\thepage}
    \fancyhead[LO]{Finite-Sample and Asymptotic Properties of Linear GMM}
    \fancyhead[RE]{Econometrics}

execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE, cache=TRUE, cache.extra=tools::md5sum("scripts/gmm_finite_sample_asymptotic_sim.R")}
source("scripts/gmm_finite_sample_asymptotic_sim.R")
library(knitr)
```

# Introduction

The Generalized Method of Moments (GMM) estimator of @Hansen2022 [see also
@Hayashi2000; @Hall2005] is the workhorse estimator for overidentified linear
models with endogenous regressors: it nests ordinary least squares, simple
instrumental variables, and two-stage least squares (2SLS) as special cases,
and it remains valid when the number of instruments exceeds the number of
parameters and when the structural error is conditionally heteroskedastic.
Its large-sample theory is standard textbook material -- consistency,
asymptotic normality, an efficiency bound attained by an optimal weighting
matrix, and a chi-squared test of overidentifying restrictions. What that
theory does *not* say is how large a sample has to be before its predictions
become a good guide to what an applied researcher actually sees.

This handout has two purposes. Section 2 develops the finite-sample and
asymptotic theory of the linear GMM estimator rigorously, from the moment
condition through the closed-form estimator, its consistency and asymptotic
normality, the efficiency bound, the two-step feasible efficient estimator,
and Hansen's $J$ test. Section 3 sets out a Monte Carlo design -- an
overidentified, conditionally heteroskedastic instrumental-variables model in
which the one-step (2SLS-weighted) and two-step efficient GMM estimators
genuinely differ -- that is built to make each asymptotic result in Section 2
independently checkable in finite samples. Section 4 reports and interprets
the simulation evidence, and Section 5 concludes.

# Theoretical Background

## The Linear Moment-Condition Model

Let $\{(y_i, \bm{x}_i, \bm{z}_i)\}_{i=1}^n$ be an independent and identically
distributed sample, where $y_i$ is a scalar outcome, $\bm{x}_i$ is a
$K \times 1$ vector of regressors (possibly including endogenous variables),
and $\bm{z}_i$ is an $L \times 1$ vector of instruments with $L \geq K$. The
model is linear,
$$
y_i = \bm{x}_i^\top \bbeta_0 + \varepsilon_i,
$$
and identification comes entirely from the population moment condition
\begin{equation}
\E\big[\bm{z}_i \varepsilon_i(\bbeta_0)\big] = \bm{0}_L, \qquad
\varepsilon_i(\bbeta) \equiv y_i - \bm{x}_i^\top \bbeta,
\end{equation}
which states that the instruments are uncorrelated with the structural error
at the true parameter value $\bbeta_0$. When $L > K$, (1) imposes $L - K$
*overidentifying restrictions*: more moment conditions than are needed to
pin down $\bbeta_0$, which is what makes the specification testable (Section
2.6) and what makes an efficiency question ("which linear combination of the
$L$ conditions should we use?") meaningful in the first place.

Stacking observations into $\bX$ ($n\times K$), $\bZ$ ($n\times L$), and $\by$
($n\times1$), define the *sample moment vector*
$$
\bar{\bm{g}}_n(\bbeta) \equiv \frac{1}{n}\bZ^\top(\by - \bX\bbeta)
= \frac{1}{n}\sum_{i=1}^n \bm{z}_i\big(y_i - \bm{x}_i^\top\bbeta\big).
$$
By the weak law of large numbers, $\bar{\bm g}_n(\bbeta_0) \xrightarrow{p}
\E[\bm z_i\varepsilon_i(\bbeta_0)] = \bm 0$, but for $L>K$ there is in general
no $\bbeta$ that sets $\bar{\bm g}_n(\bbeta)$ *exactly* to zero in a given
sample: the $L$ sample moment conditions are $L$ equations in only $K$
unknowns. GMM resolves this by choosing $\bbeta$ to make the sample moments
as close to zero as possible, in a quadratic metric.

## The GMM Objective Function and the Closed-Form Estimator

For a symmetric, positive definite weighting matrix $\bm{W}_n$, the GMM
estimator is
\begin{equation}
\hat{\bbeta}(\bm{W}_n) = \arg\min_{\bbeta}\; Q_n(\bbeta), \qquad
Q_n(\bbeta) \equiv \bar{\bm g}_n(\bbeta)^\top \bm{W}_n\, \bar{\bm g}_n(\bbeta).
\end{equation}
Because the model is linear, $Q_n$ is an exact quadratic form in $\bbeta$ and
(2) has a closed-form solution. Writing $\bar{\bm g}_n(\bbeta) =
\tfrac1n\bZ^\top\by - \tfrac1n\bZ^\top\bX\bbeta$, the first-order condition
$\partial Q_n/\partial\bbeta = -\tfrac2n \bX^\top\bZ\,\bm W_n\,
\bar{\bm g}_n(\bbeta) = \bm 0$ gives
$$
\bX^\top\bZ\,\bm{W}_n\,\bZ^\top\bX\,\hat{\bbeta}(\bm W_n)
= \bX^\top\bZ\,\bm{W}_n\,\bZ^\top\by,
$$
so that, whenever $\bX^\top\bZ\,\bm W_n\,\bZ^\top\bX$ is invertible,
\begin{equation}
\hat{\bbeta}(\bm W_n) = \big(\bX^\top\bZ\,\bm{W}_n\,\bZ^\top\bX\big)^{-1}
\bX^\top\bZ\,\bm{W}_n\,\bZ^\top\by.
\end{equation}
Two familiar estimators are special cases of (3): $\bm W_n = \bm I_L$ gives
unweighted GMM, and $\bm W_n = (\bZ^\top\bZ/n)^{-1}$ gives numerically the
2SLS estimator. Neither is asymptotically efficient in general; Section 2.5
derives the weighting matrix that is.

## Consistency

\begin{assumption}[Identification and Regularity]
\leavevmode
\begin{enumerate}
\item \textbf{(Correct specification)} $\E[\bm z_i\varepsilon_i(\bbeta)] = \bm 0$
   holds if and only if $\bbeta=\bbeta_0$, for $\bbeta_0$ in the interior of
   a compact parameter space.
\item \textbf{(Rank condition)} $\bm G \equiv \E[\bm z_i\bm x_i^\top]$ has full
   column rank $K$ (in particular $L\geq K$).
\item \textbf{(Weighting matrix)} $\bm W_n \xrightarrow{p} \bm W$, a finite,
   symmetric, positive definite $L\times L$ matrix.
\item \textbf{(Moments)} $\E\|\bm z_i\bm x_i^\top\| < \infty$ and
   $\E\|\bm z_i y_i\| < \infty$.
\end{enumerate}
\end{assumption}

\begin{proposition}[Consistency of Linear GMM]
Under Assumption 1,
$\hat{\bbeta}(\bm W_n) \xrightarrow{p} \bbeta_0$.
\end{proposition}

\begin{proof}
Substituting $y_i = \bm x_i^\top\bbeta_0+\varepsilon_i$ into
$\bar{\bm g}_n(\bbeta_0) = \tfrac1n\bZ^\top\beps$ and writing
$\hat{\bm G}_n \equiv \tfrac1n\bZ^\top\bX$, the closed form (3) rearranges to
\begin{equation}
\hat{\bbeta}(\bm W_n) - \bbeta_0
= \big(\hat{\bm G}_n^\top \bm W_n \hat{\bm G}_n\big)^{-1}
  \hat{\bm G}_n^\top \bm W_n\, \bar{\bm g}_n(\bbeta_0).
\end{equation}
By the weak law of large numbers, $\hat{\bm G}_n \xrightarrow{p} \bm G$ and
$\bar{\bm g}_n(\bbeta_0)\xrightarrow{p}\E[\bm z_i\varepsilon_i(\bbeta_0)] =
\bm 0$ by Assumption 1.1. Combined with $\bm W_n\xrightarrow{p}\bm W$ and the
continuous-mapping theorem, $\bm G^\top\bm W\bm G$ is invertible by
Assumption 1.2--1.3 (full column rank $\bm G$ and positive definite $\bm W$
imply $\bm G^\top\bm W\bm G$ is positive definite, hence invertible), so the
right-hand side of (4) converges in probability to
$(\bm G^\top\bm W\bm G)^{-1}\bm G^\top\bm W\cdot\bm 0=\bm 0$. $\blacksquare$
\end{proof}

Two features of this argument are worth flagging because the simulation is
built around them. First, consistency holds for *every* positive definite
limiting weight $\bm W$ -- the one-step and two-step estimators of Section
2.5 are both consistent, so any efficiency difference between them is a
finite-sample or asymptotic-variance phenomenon, not a difference in the
target they estimate. Second, the proof uses only a law of large numbers; no
distributional assumption on $\varepsilon_i$, and in particular no
homoskedasticity, is needed.

## Asymptotic Normality

\begin{assumption}[Central Limit Theorem for the Moments]
In addition to Assumption 1,
$\bm S \equiv \Var\big(\bm z_i\varepsilon_i(\bbeta_0)\big) =
\E[\bm z_i\bm z_i^\top\varepsilon_i^2]$ is finite and positive definite, and
$$
\sqrt n\,\bar{\bm g}_n(\bbeta_0) = \frac{1}{\sqrt n}\bZ^\top\beps
\xrightarrow{d} N(\bm 0, \bm S).
$$
\end{assumption}

Assumption 2 allows $\varepsilon_i$ to be conditionally heteroskedastic --
$\E[\varepsilon_i^2\mid \bm z_i]$ need not be constant -- which is exactly the
feature built into the Monte Carlo design below, and is why $\bm S$ is
written as $\E[\bm z_i\bm z_i^\top\varepsilon_i^2]$ rather than
$\sigma^2\E[\bm z_i\bm z_i^\top]$.

\begin{proposition}[Asymptotic Normality of Linear GMM]
Under Assumptions 1--2,
\begin{equation}
\sqrt n\big(\hat{\bbeta}(\bm W_n) - \bbeta_0\big) \xrightarrow{d}
N\Big(\bm 0,\; \bm V(\bm W)\Big), \qquad
\bm V(\bm W) = (\bm G^\top\bm W\bm G)^{-1}\bm G^\top\bm W\bm S\bm W\bm G(\bm G^\top\bm W\bm G)^{-1}.
\end{equation}
\end{proposition}

\begin{proof}
Multiply (4) by $\sqrt n$:
$$
\sqrt n\big(\hat{\bbeta}(\bm W_n)-\bbeta_0\big)
= \big(\hat{\bm G}_n^\top\bm W_n\hat{\bm G}_n\big)^{-1}
  \hat{\bm G}_n^\top\bm W_n\;\sqrt n\,\bar{\bm g}_n(\bbeta_0).
$$
By Assumption 2, $\sqrt n\,\bar{\bm g}_n(\bbeta_0)\xrightarrow{d}N(\bm0,\bm
S)$; by Assumption 1, $\hat{\bm G}_n\xrightarrow{p}\bm G$ and $\bm
W_n\xrightarrow{p}\bm W$. Slutsky's theorem applied to the product of the
(stochastically bounded, converging-in-probability) matrix
$(\hat{\bm G}_n^\top\bm W_n\hat{\bm G}_n)^{-1}\hat{\bm G}_n^\top\bm W_n$ and
the asymptotically normal vector $\sqrt n\,\bar{\bm g}_n(\bbeta_0)$ gives
$$
\sqrt n\big(\hat{\bbeta}(\bm W_n)-\bbeta_0\big) \xrightarrow{d}
(\bm G^\top\bm W\bm G)^{-1}\bm G^\top\bm W\,\bm u, \qquad \bm u\sim N(\bm0,\bm S),
$$
and the variance of a linear transformation $\bm A\bm u$ of $\bm u\sim
N(\bm0,\bm S)$ is $\bm A\bm S\bm A^\top$, which for $\bm A =
(\bm G^\top\bm W\bm G)^{-1}\bm G^\top\bm W$ gives exactly (5).
$\blacksquare$
\end{proof}

In a finite sample, (5) is implemented by plugging in consistent estimators
$\hat{\bm G}_n=\tfrac1n\bZ^\top\bX$, $\hat{\bm S}_n$ (Section 2.5), and $\bm
W_n$, giving the *sandwich* variance estimator
\begin{equation}
\widehat{\Var}\big(\hat{\bbeta}(\bm W_n)\big)
= \frac1n\big(\hat{\bm G}_n^\top\bm W_n\hat{\bm G}_n\big)^{-1}
  \hat{\bm G}_n^\top\bm W_n\hat{\bm S}_n\bm W_n\hat{\bm G}_n
  \big(\hat{\bm G}_n^\top\bm W_n\hat{\bm G}_n\big)^{-1}.
\end{equation}
This is the standard-error formula used for both estimators in the
simulation; it collapses to a simpler expression for the efficient estimator
below.

## Efficient GMM and the Two-Step Estimator

Equation (5) shows that the asymptotic variance depends on the choice of
$\bm W$. @Hansen2022's classical result identifies the efficient choice.

\begin{proposition}[GMM Efficiency Bound]
Among all weighting matrices $\bm W$ satisfying
Assumption 1.3, $\bm V(\bm W) - \bm V(\bm S^{-1})$ is positive semi-definite
for every $\bm W$, so $\bm W = \bm S^{-1}$ minimizes the asymptotic variance
(5) in the positive semi-definite ordering, and
\begin{equation}
\bm V(\bm S^{-1}) = (\bm G^\top\bm S^{-1}\bm G)^{-1}.
\end{equation}
\end{proposition}

\begin{proof}
Substituting $\bm W=\bm S^{-1}$ into (5) gives $\bm V(\bm S^{-1}) =
(\bm G^\top\bm S^{-1}\bm G)^{-1}\bm G^\top\bm S^{-1}\bm S\bm S^{-1}\bm G(\bm
G^\top\bm S^{-1}\bm G)^{-1} = (\bm G^\top\bm S^{-1}\bm G)^{-1}$, which is (7).
For optimality, let $\bm S^{1/2}$ be a symmetric square root of $\bm S$ and
define $\bm A \equiv (\bm G^\top\bm W\bm G)^{-1}\bm G^\top\bm W\bm S^{1/2}$
and $\bm B \equiv (\bm G^\top\bm S^{-1}\bm G)^{-1}\bm G^\top\bm S^{-1/2}$, so
that $\bm V(\bm W)=\bm A\bm A^\top$ and $\bm V(\bm S^{-1})=\bm B\bm B^\top$.
Direct multiplication using $\bm S^{-1/2}\bm S^{1/2}=\bm I_L$ verifies
$\bm B\bm A^\top = \bm B\bm B^\top$, i.e. $(\bm A-\bm B)\bm B^\top = \bm 0$,
so $\bm A\bm A^\top = \bm B\bm B^\top + (\bm A-\bm B)(\bm A-\bm B)^\top$, and
$(\bm A-\bm B)(\bm A-\bm B)^\top$ is positive semi-definite. Hence $\bm
V(\bm W) - \bm V(\bm S^{-1}) = (\bm A-\bm B)(\bm A-\bm B)^\top \succeq \bm
0$. $\blacksquare$
\end{proof}

The efficiency bound (7) is attractive but infeasible: $\bm S$ is unknown.
The standard solution is **two-step efficient GMM**:

1. **Step 1.** Choose any consistent, feasible weighting matrix -- in this
   handout, $\bm W_n^{(1)} = (\bZ^\top\bZ/n)^{-1}$, which makes
   $\hat{\bbeta}^{(1)} \equiv \hat{\bbeta}(\bm W_n^{(1)})$ numerically equal to
   the 2SLS estimator. By Proposition 1, $\hat{\bbeta}^{(1)}$ is consistent
   even though $\bm W_n^{(1)}$ is generally *not* efficient.
2. **Estimate $\bm S$.** Using the step-1 residuals $\hat\varepsilon_i^{(1)}
   = y_i - \bm x_i^\top\hat{\bbeta}^{(1)}$, form the heteroskedasticity-robust
   estimator
   $$
   \hat{\bm S}_n = \frac1n\sum_{i=1}^n \bm z_i\bm z_i^\top
   \big(\hat\varepsilon_i^{(1)}\big)^2 \xrightarrow{p} \bm S,
   $$
   which is consistent for $\bm S$ under Assumption 2 regardless of the
   form of the conditional heteroskedasticity [@White2000].
3. **Step 2.** Set $\bm W_n^{(2)} = \hat{\bm S}_n^{-1}$ and compute
   $\hat{\bbeta}^{(2)} \equiv \hat{\bbeta}(\bm W_n^{(2)})$.

Because $\bm W_n^{(2)}\xrightarrow{p}\bm S^{-1}$, Proposition 3 applies with
$\bm W=\bm S^{-1}$, so $\sqrt n(\hat{\bbeta}^{(2)}-\bbeta_0)\xrightarrow{d}
N\big(\bm0,(\bm G^\top\bm S^{-1}\bm G)^{-1}\big)$, and the sandwich formula
(6) collapses to
\begin{equation}
\widehat{\Var}\big(\hat{\bbeta}^{(2)}\big)
= \frac1n\big(\hat{\bm G}_n^\top\hat{\bm S}_n^{-1}\hat{\bm G}_n\big)^{-1}.
\end{equation}
This is the estimator the simulation calls the "two-step efficient GMM"
estimator, and (8) is the variance used for its confidence intervals.

## Testing Overidentifying Restrictions

When $L>K$, correct specification (Assumption 1.1) is testable: it implies
$\bar{\bm g}_n(\hat{\bbeta}^{(2)})$ should be close to zero even though nothing
in the estimation *forces* all $L$ sample moments to vanish. Hansen's $J$
statistic formalizes "close to zero."

\begin{proposition}[Hansen's $J$ Test]
Under Assumptions 1--2 and correct specification,
$$
J_n \equiv n\,\bar{\bm g}_n(\hat{\bbeta}^{(2)})^\top \hat{\bm S}_n^{-1}
\bar{\bm g}_n(\hat{\bbeta}^{(2)}) \xrightarrow{d} \chi^2(L-K).
$$
\end{proposition}

\begin{proof}
A first-order Taylor expansion of $\sqrt n\bar{\bm g}_n(\hat{\bbeta}^{(2)})$
around $\bbeta_0$, combined with the first-order condition defining
$\hat{\bbeta}^{(2)}$, gives
$$
\sqrt n\,\bar{\bm g}_n(\hat{\bbeta}^{(2)})
= \Big[\bm I_L - \bm G(\bm G^\top\bm S^{-1}\bm G)^{-1}\bm G^\top\bm
S^{-1}\Big]\sqrt n\,\bar{\bm g}_n(\bbeta_0) + o_p(1).
$$
Premultiplying by $\bm S^{-1/2}$ and writing $\bm u \equiv \bm
S^{-1/2}\sqrt n\,\bar{\bm g}_n(\bbeta_0) \xrightarrow{d} N(\bm 0,\bm I_L)$
(Assumption 2) and $\bm P \equiv \bm S^{-1/2}\bm G(\bm G^\top\bm S^{-1}\bm
G)^{-1}\bm G^\top\bm S^{-1/2}$, this is
$$
\bm S^{-1/2}\sqrt n\,\bar{\bm g}_n(\hat{\bbeta}^{(2)}) = (\bm I_L-\bm P)\,\bm u + o_p(1).
$$
$\bm P$ is idempotent ($\bm P^2=\bm P$, by direct multiplication) and
symmetric, with $\mathrm{rank}(\bm P) = \mathrm{tr}(\bm P) =
\mathrm{tr}\big((\bm G^\top\bm S^{-1}\bm G)^{-1}\bm G^\top\bm S^{-1}\bm
G\big) = \mathrm{tr}(\bm I_K) = K$, so $\bm I_L-\bm P$ is itself an
idempotent projection of rank $L-K$. Since $J_n \xrightarrow{d} \bm
u^\top(\bm I_L-\bm P)\bm u$ and $\bm u\sim N(\bm0,\bm I_L)$, this is the
quadratic form of a standard normal vector in an idempotent matrix of rank
$L-K$, which is $\chi^2(L-K)$ by the standard result for quadratic forms in
normal vectors. $\blacksquare$
\end{proof}

With one endogenous regressor and two excluded instruments (Section 3), the
design below has $L=3$, $K=2$, and a single overidentifying restriction, so
$J_n\xrightarrow{d}\chi^2(1)$ under correct specification.

## Finite-Sample Properties: What Asymptotic Theory Does Not Guarantee

Propositions 1--4 are exact statements about limits as $n\to\infty$; none of
them bounds how large $n$ must be in practice, and each can fail badly in a
finite sample that its own asymptotic approximation does not warn against.
Three caveats motivate the simulation design in Section 3:

1. **GMM estimators are, in general, biased in finite samples**, even though
   they are consistent. Unlike OLS under the classical Gauss--Markov
   assumptions, $\E[\hat{\bbeta}(\bm W_n)]\neq\bbeta_0$ at any fixed $n$ for
   the linear IV/GMM estimator, because $\hat{\bbeta}(\bm W_n)$ is a nonlinear
   (ratio-like) function of the sample moments even though the *population*
   moment condition is linear in $\bbeta$; the bias is $O(1/n)$ under
   standard regularity conditions and shrinks as identification strengthens
   [@Hall2005, ch.\ 3; @Hayashi2000, sec.\ 3.4].
2. **Two-step efficient GMM introduces an extra source of finite-sample
   noise**: $\hat{\bm S}_n$ in Section 2.5 is itself estimated, using
   step-1 residuals, and that estimation error propagates into
   $\hat{\bbeta}^{(2)}$. In sufficiently small samples, or with many
   overidentifying moments, this can erode or even reverse the asymptotic
   efficiency ranking of Proposition 3 relative to a well-chosen one-step
   estimator -- a concern that motivates alternatives such as
   continuously-updated GMM [@Hall2005, ch.\ 4].
3. **Weak or many instruments distort both estimation and inference.** When
   $\bm G$ is close to reduced rank (weak identification) the finite-sample
   distribution of $\hat{\bbeta}(\bm W_n)$ can be far from normal and biased
   toward the OLS estimator's plim, and the sandwich standard errors (6)
   and (8) can be badly miscalibrated even at moderate $n$
   [@ChausseIV2025].

The design in Section 3 deliberately avoids weak identification (caveat 3)
so that the simulation isolates caveats 1--2 -- ordinary finite-sample bias
and the cost of estimating the optimal weighting matrix -- against the
backdrop of Propositions 1--4.

# Monte Carlo Design

## Data-Generating Process

The design has one endogenous regressor and two excluded, exogenous
instruments, so $L=3$ instruments ($1,z_{1i},z_{2i}$) and $K=2$ regressors
($1,x_{2i}$): one overidentifying restriction. For each replication:
$$
z_{1i}, z_{2i} \overset{\text{iid}}{\sim} N(0,1), \qquad
u_i, w_i \overset{\text{iid}}{\sim} N(0,1), \quad
u_i, w_i \perp z_{1i}, z_{2i},
$$
$$
x_{2i} = \pi_0 + \pi_1 z_{1i} + \pi_2 z_{2i} + u_i, \qquad
e_i = \rho\, u_i + \sqrt{1-\rho^2}\, w_i, \qquad
\varepsilon_i = \sqrt{0.5+z_{1i}^2}\; e_i,
$$
$$
y_i = \beta_1 + \beta_2\, x_{2i} + \varepsilon_i.
$$
The first-stage error $u_i$ enters both $x_{2i}$ and (through $e_i$) the
structural error $\varepsilon_i$, so $x_{2i}$ is endogenous with
$\mathrm{Corr}(u_i, e_i) = \rho$; the parametrization $e_i = \rho u_i +
\sqrt{1-\rho^2}w_i$ with $u_i,w_i$ independent standard normal makes this
exact regardless of $\rho\in(-1,1)$. The multiplicative factor
$\sqrt{0.5+z_{1i}^2}$ makes $\Var(\varepsilon_i\mid z_{1i},z_{2i}) =
0.5+z_{1i}^2$ depend on the instrument $z_{1i}$: the errors are
conditionally heteroskedastic, so $\bm S=\E[\bm z_i\bm
z_i^\top\varepsilon_i^2]\neq\sigma^2\E[\bm z_i\bm z_i^\top]$ and the one-step
(2SLS-weighted) and two-step efficient GMM estimators of Section 2.5 have
genuinely different asymptotic variances -- without heteroskedasticity the
two would coincide asymptotically and the comparison in Section 4.1 would be
uninformative.

Parameter values, held fixed across all replications and sample sizes:
$$
(\beta_1,\beta_2) = (`r beta_true[["intercept"]]`, `r beta_true[["slope"]]`),
\qquad
(\pi_0,\pi_1,\pi_2) = (`r pi_coef[1]`, `r pi_coef[2]`, `r pi_coef[3]`),
\qquad \rho = `r rho`.
$$
The instruments are relevant ($\pi_1=\pi_2=1$, not weak) and the
endogeneity is moderate ($\rho=`r rho`$), consistent with the decision in
Section 2.6 to isolate ordinary finite-sample bias and weighting-matrix
estimation error rather than weak-instrument pathologies.

## Estimators Compared

For each simulated sample, two estimators are computed from the identical
data, exactly as in Section 2.5:

- **One-step GMM** ($\hat\beta^{(1)}$): $\bm W_n^{(1)} = (\bZ^\top\bZ/n)^{-1}$,
  numerically identical to 2SLS, with sandwich standard errors from (6).
- **Two-step efficient GMM** ($\hat\beta^{(2)}$): $\bm W_n^{(2)} =
  \hat{\bm S}_n^{-1}$, with $\hat{\bm S}_n$ the heteroskedasticity-robust
  estimator built from the $\hat\beta^{(1)}$ residuals, and standard errors
  from the simplified formula (8).

Both estimators are consistent for $\bbeta_0$ by Proposition 1; only their
asymptotic (and finite-sample) variance should differ. Hansen's $J$
statistic of Section 2.6 is computed at $\hat\beta^{(2)}$ using $\bm
W_n^{(2)}$, with $L-K=1$ degree of freedom.

## Simulation Parameters and Reported Statistics

The sample size grid is $n\in\{`r paste(n_grid, collapse=", ")`\}$, with
$N_{sim}=`r Nsim`$ independent replications drawn at each $n$ (seed fixed at
the start of the experiment for reproducibility). All reported quantities
concern the slope $\beta_2$, the coefficient on the endogenous regressor;
the intercept behaves analogously and is omitted. At each $n$ the script
`gmm_finite_sample_asymptotic_sim.R` computes, across the $N_{sim}$
replications:

- **Bias** and **variance** of $\hat\beta_2^{(1)}$ and $\hat\beta_2^{(2)}$
  around the true $\beta_2$ (Section 4.1: consistency and efficiency);
- **RMSE** of both estimators;
- **Coverage** of the nominal 95\% confidence interval
  $\hat\beta_2 \pm 1.96\times\widehat{\mathrm{se}}(\hat\beta_2)$, using the
  sandwich standard errors of (6) and (8) (Section 4.3);
- The rejection frequency of Hansen's $J_n$ statistic at the nominal 5\%
  level against $\chi^2(1)$ (Section 4.4).

In addition, at a small sample size ($n=`r n_small`$) and a large one
($n=`r n_large`$), the script retains every replication's studentized
statistic $(\hat\beta_2^{(2)}-\beta_2)/\widehat{\mathrm{se}}(\hat\beta_2^{(2)})$,
which Proposition 2 predicts converges to $N(0,1)$, to visualize how close
that limit is at each $n$ (Section 4.2).

# Results

## Consistency and Efficiency

@fig-bias plots Monte Carlo bias against $n$ (log scale) for both
estimators. Both are visually indistinguishable from zero once $n$ reaches a
few hundred, and neither series shows a systematic sign or trend as $n$
grows further -- exactly what Proposition 1 predicts, since both weighting
matrices satisfy Assumption 1.3 regardless of their efficiency. At the
smallest sample size, $n=`r n_grid[1]`$, the one-step estimator has bias
`r bias_of(n_grid[1], "1step")` and the two-step estimator
`r bias_of(n_grid[1], "2step")`; by $n=`r n_grid[length(n_grid)]`$ both have
shrunk an order of magnitude further, to
`r bias_of(n_grid[length(n_grid)], "1step")` and
`r bias_of(n_grid[length(n_grid)], "2step")` respectively -- consistent
with the $O(1/n)$ finite-sample bias flagged in Section 2.7, caveat 1, which
predicts bias shrinking in proportion to $n$, faster than the
$O(1/\sqrt n)$ rate of the standard error itself.

```{r fig-bias}
#| label: fig-bias
#| fig-cap: "Monte Carlo bias of the one-step (2SLS-weighted) and two-step efficient GMM slope estimators, by sample size."
#| fig-width: 5.6
#| fig-height: 3.8
p_bias
```

@fig-variance plots Monte Carlo variance against $n$ on a log-log scale.
Two features match Section 2 exactly. First, both curves are close to a
straight line of slope $-1$, the $O(1/n)$ rate implied by (5)/(7) (variance
$\approx \bm V/n$ for fixed asymptotic $\bm V$). Second, the two-step
efficient estimator's variance lies *below* the one-step estimator's at
**every** sample size in the grid -- the efficiency ranking of Proposition 3
is not just an asymptotic statement here but holds throughout the range
simulated. The efficiency gain,
$1-\widehat{\Var}(\hat\beta^{(2)})/\widehat{\Var}(\hat\beta^{(1)})$, is
`r efficiency_gain_of(n_grid[1])` at $n=`r n_grid[1]`$ and
`r efficiency_gain_of(n_grid[length(n_grid)])` at
$n=`r n_grid[length(n_grid)]`$: the efficient estimator's advantage, driven
by the conditional heteroskedasticity built into the design (Section 3.1),
is present even in small samples and if anything widens as $n$ grows and
$\hat{\bm S}_n$ is estimated more precisely.

```{r fig-variance}
#| label: fig-variance
#| fig-cap: "Monte Carlo variance of the two GMM slope estimators, log-log scale; the two-step efficient estimator (Proposition 3) attains lower variance at every sample size."
#| fig-width: 5.6
#| fig-height: 3.8
p_var
```

## Asymptotic Normality

Proposition 2 predicts that the studentized statistic
$(\hat\beta_2^{(2)}-\beta_2)/\widehat{\mathrm{se}}(\hat\beta_2^{(2)})$ is
approximately $N(0,1)$, with the approximation improving as $n\to\infty$.
@fig-density overlays the Monte Carlo density of this statistic against the
$N(0,1)$ density at $n=`r n_small`$ and $n=`r n_large`$, and @fig-qq shows
the corresponding QQ-plots against theoretical normal quantiles. At
$n=`r n_small`$ the histogram is visibly heavier in the centre and the
QQ-plot bows away from the 45-degree line in the tails -- a finite-sample
departure from normality consistent with the $O(1/n)$ bias and imperfectly
estimated $\hat{\bm S}_n$ discussed in Section 2.7. At $n=`r n_large`$ both
diagnostics track the normal reference closely, illustrating Proposition 2's
limit being approached, not merely asserted.

```{r fig-density}
#| label: fig-density
#| fig-cap: "Density of the studentized two-step GMM slope estimator at a small and a large sample size, against its N(0,1) asymptotic limit."
#| fig-width: 6.0
#| fig-height: 3.4
p_density
```

```{r fig-qq}
#| label: fig-qq
#| fig-cap: "QQ-plot of the studentized two-step GMM slope estimator against theoretical normal quantiles."
#| fig-width: 6.0
#| fig-height: 3.4
p_qq
```

## Coverage of Confidence Intervals

@fig-coverage plots the empirical coverage of the nominal 95\% confidence
interval, built from the sandwich standard errors (6) and (8), against $n$.
At $n=`r n_grid[1]`$ coverage is `r cov_of(n_grid[1], "1step")` for the
one-step estimator and `r cov_of(n_grid[1], "2step")` for the two-step
estimator -- both below the nominal 95\%, the direct consequence of the
finite-sample bias and non-normality documented in Sections 4.1--4.2 feeding
into an interval that assumes exact normality and a correctly centred
point estimate. Coverage rises toward the nominal level as $n$ grows,
reaching `r cov_of(n_grid[length(n_grid)], "1step")` and
`r cov_of(n_grid[length(n_grid)], "2step")` respectively at
$n=`r n_grid[length(n_grid)]`$, consistent with Proposition 2. This is the
practical stakes of Section 2.7: a researcher quoting a 95\% CI from GMM
estimated on a few dozen observations is, in this design, actually covering
the true slope somewhat less often than the label promises.

```{r fig-coverage}
#| label: fig-coverage
#| fig-cap: "Empirical coverage of the nominal 95% confidence interval for the GMM slope estimators, using the sandwich standard errors of equations (6) and (8)."
#| fig-width: 5.6
#| fig-height: 3.8
p_coverage
```

## Size of the Overidentification Test

@fig-jtest plots the rejection frequency of Hansen's $J_n$ statistic at the
nominal 5\% level against $\chi^2(1)$, as a function of $n$. Because the
model is correctly specified at every $n$ (the same DGP generates the data
used to estimate and to test it), Proposition 4 predicts a rejection
frequency of exactly 5\% in the limit; this is a test of **size**, not
power. The rejection frequency is `r jrej_of(n_grid[1])` at
$n=`r n_grid[1]`$ and settles near the nominal `r pc(alpha)` for
$n\geq250$, confirming that the $\chi^2(1)$ approximation of Proposition 4
is reliable well before the sample sizes at which the coverage results of
Section 4.3 fully converge -- the $J$ test's asymptotic approximation kicks
in faster here than the confidence-interval coverage does, because the $J$
statistic depends on $\hat{\bm S}_n$ only through its role in weighting,
whereas coverage also depends on the point estimate's own finite-sample
bias.

```{r fig-jtest}
#| label: fig-jtest
#| fig-cap: "Rejection frequency of Hansen's J overidentification test (df = 1) at the nominal 5% level, under correct specification."
#| fig-width: 5.6
#| fig-height: 3.8
p_jtest
```

## Summary Table

@tbl-summary collects all of the reported statistics by sample size.

```{r tbl-summary}
#| label: tbl-summary
#| tbl-cap: !expr paste0("Monte Carlo bias, variance, RMSE, 95% CI coverage, and J-test rejection frequency of the one-step and two-step GMM slope estimators (N_sim = ", Nsim, " replications per row).")
display_tbl <- data.frame(
  n = summary_table$n,
  Bias1 = summary_table$bias_1step,
  Bias2 = summary_table$bias_2step,
  Var1 = summary_table$var_1step,
  Var2 = summary_table$var_2step,
  RMSE1 = summary_table$rmse_1step,
  RMSE2 = summary_table$rmse_2step
)
colnames(display_tbl) <- c("$n$", "Bias (1-step)", "Bias (2-step)",
                           "Var (1-step)", "Var (2-step)",
                           "RMSE (1-step)", "RMSE (2-step)")
knitr::kable(display_tbl, digits = c(0, 4, 4, 5, 5, 4, 4),
            booktabs = TRUE, align = c("r", rep("c", 6)))
```

```{r tbl-inference}
#| label: tbl-inference
#| tbl-cap: "Coverage of the nominal 95% confidence interval and rejection frequency of Hansen's J test, by sample size."
display_tbl2 <- data.frame(
  n = summary_table$n,
  Cov1 = summary_table$coverage_1step,
  Cov2 = summary_table$coverage_2step,
  Jrej = summary_table$J_rejection
)
colnames(display_tbl2) <- c("$n$", "Coverage (1-step)", "Coverage (2-step)",
                            "J-test rejection")
knitr::kable(display_tbl2, digits = c(0, 3, 3, 3),
            booktabs = TRUE, align = c("r", "c", "c", "c"))
```

# Discussion and Conclusion

The theory of Section 2 and the simulation of Sections 3--4 tell the same
story from two directions. Propositions 1--4 are exact limiting results, and
the Monte Carlo evidence confirms every one of them is eventually a very
good approximation: bias and variance shrink toward zero, the two-step
estimator dominates the one-step estimator in variance exactly as the
efficiency bound (7) predicts, the studentized estimator's distribution
converges visibly to standard normal, confidence-interval coverage
approaches its nominal level, and the $J$ test is correctly sized. None of
this is automatic at the sample sizes an applied researcher actually has:
at $n=`r n_grid[1]`$, coverage is several percentage points below nominal
for both estimators, and the studentized distribution is visibly non-normal,
even though the design avoids the weak-instrument pathologies that make
GMM/IV inference unreliable for reasons unrelated to sample size (Section
2.7, caveat 3). The practical implication is not that GMM inference is
unreliable in general, but that the gap between "consistent and
asymptotically normal" and "trustworthy at my sample size" has to be
checked, not assumed -- exactly the exercise this handout carries out.

The companion script, `scripts/gmm_finite_sample_asymptotic_sim.R`, is the
single computational source for every number and figure in this handout: it
implements the closed-form estimator (3), the sandwich variances (6) and
(8), and Hansen's $J$ statistic exactly as derived in Section 2, and can be
re-run (`Rscript scripts/gmm_finite_sample_asymptotic_sim.R`) to reproduce
every result reported here, or edited to explore other sample-size grids,
degrees of heteroskedasticity, or instrument strength.

# References
