---
title: "Monte Carlo Simulation of Spherical and Non-Spherical Errors"
subtitle: "Econometrics Note: Efficiency of OLS under Different Error Structures"
author: "Eliab Luvanda"
date: today
format:
  pdf:
    documentclass: article
    pdf-engine: xelatex
    classoption:
      - twoside
    fontsize: 11pt
    papersize: a4
    mainfont: "Palatino"
    toc: true
    number-sections: true
    colorlinks: true
    include-in-header:
      - text: |
          \usepackage{float}
          \usepackage{placeins}
editor: source
execute:
  echo: true
  warning: false
  message: false
---

```{=latex}
\clearpage
\listoftables
\clearpage
\listoffigures
\clearpage
```

# Introduction

This note presents a Monte Carlo experiment designed to illustrate a central result of regression analysis: when the disturbance term is spherical, the sampling distribution of the ordinary least squares estimator tends to be more concentrated than when the disturbance term exhibits heteroscedasticity or serial correlation.

A simple setup is this:

$$
y_i = \beta_0 + \beta_1 x_i + u_i
$$

Then simulate many datasets under three cases:

1. Spherical errors:

$$
u_i \sim \mathcal{N}(0, \sigma^2)
$$

with constant variance and no correlation.

2. Heteroscedastic errors:

$$
u_i \sim \mathcal{N}(0, \sigma_i^2), \qquad \sigma_i^2 = \sigma^2(1 + \gamma x_i^2)
$$

3. Serially correlated and heteroscedastic errors, for example:

$$
u_t = \rho u_{t-1} + \varepsilon_t, \qquad \varepsilon_t \sim \mathcal{N}(0, \sigma_t^2)
$$

with $\sigma_t^2$ varying over $t$.

Then, across many replications, compare the empirical distribution of $\hat{\beta}_1$. Under spherical errors, OLS is unbiased and efficient within the Gauss--Markov class, so its sampling distribution should usually be tighter, meaning that it has smaller variance. Under heteroscedasticity and serial correlation, OLS may remain unbiased under exogeneity, but it is no longer efficient, so the Monte Carlo distribution of $\hat{\beta}_1$ will typically be more spread out.

# Theoretical Background

The heuristic inequality

$$
\operatorname{Var}(\hat{\beta}_1 \mid \Omega = \sigma^2 I)
<
\operatorname{Var}(\hat{\beta}_1 \mid \Omega \neq \sigma^2 I)
$$

is not a universal theorem for every imaginable non-spherical covariance matrix. The precise theoretical result is the Gauss--Markov statement: within the class of linear unbiased estimators, OLS has minimum variance when the errors are spherical. Once the error covariance matrix is no longer proportional to the identity, OLS loses efficiency and, in general, another linear unbiased estimator can achieve a smaller variance.

## Linear Model and OLS Variance

Start from the linear model

$$
y = X\beta + u,
$$

where

- $y$ is an $n$-vector of observations,
- $X$ is an $n$ by $k$ matrix of regressors with full column rank,
- $\beta$ is a $k$-vector of parameters, and
- $u$ is an $n$-vector of disturbances.

Assume

$$
\mathbb{E}(u \mid X) = 0.
$$

The OLS estimator is

$$
\hat{\beta}_{\mathrm{OLS}} = (X'X)^{-1}X'y.
$$

Its conditional variance is

$$
\operatorname{Var}(\hat{\beta}_{\mathrm{OLS}} \mid X)
=
(X'X)^{-1} X'\Omega X (X'X)^{-1},
$$

where

$$
\Omega = \operatorname{Var}(u \mid X).
$$

This formula is valid under exogeneity.

## Spherical Errors

If the errors are spherical, then

$$
\Omega = \sigma^2 I.
$$

Hence

$$
\operatorname{Var}(\hat{\beta}_{\mathrm{OLS}} \mid X)
=
\sigma^2 (X'X)^{-1}.
$$

This variance matrix plays a central role in classical regression theory.

For the slope coefficient $\beta_1$, assuming it is the second parameter, this gives

$$
\operatorname{Var}(\hat{\beta}_1 \mid X, \Omega = \sigma^2 I)
=
\sigma^2 \bigl[(X'X)^{-1}\bigr]_{22}.
$$

## Heteroscedastic or Serially Correlated Errors

If the errors are non-spherical, then

$$
\Omega \neq \sigma^2 I,
$$

and therefore

$$
\operatorname{Var}(\hat{\beta}_{\mathrm{OLS}} \mid X)
=
(X'X)^{-1} X'\Omega X (X'X)^{-1}.
$$

This is no longer equal to $\sigma^2 (X'X)^{-1}$. Its magnitude depends on the exact form of $\Omega$, so a blanket coefficient-by-coefficient inequality for every non-spherical covariance matrix is not the cleanest theorem.

## Gauss--Markov Argument

Under the assumptions

1. $\mathbb{E}(u \mid X) = 0$
2. $\operatorname{Var}(u \mid X) = \sigma^2 I_n$

the Gauss--Markov theorem states that the ordinary least squares estimator $\hat{\beta}_{\mathrm{OLS}}$ has minimum variance among all linear unbiased estimators of $\beta$. Thus OLS is BLUE.

To prove this, take any linear unbiased estimator of $\beta$ of the form

$$
\beta^\star = Ay, \qquad AX = I.
$$

Under spherical errors,

$$
\operatorname{Var}(\beta^\star \mid X) = \sigma^2 AA'.
$$

Because $AX = I$, one can write

$$
A = (X'X)^{-1}X' + C, \qquad CX = 0.
$$

Then

$$
\operatorname{Var}(\beta^\star \mid X)
=
\sigma^2 \left[(X'X)^{-1} + CC'\right].
$$

Since $CC'$ is positive semidefinite,

$$
\operatorname{Var}(\beta^\star \mid X)
-
\operatorname{Var}(\hat{\beta}_{\mathrm{OLS}} \mid X)
=
\sigma^2 CC'
$$

is positive semidefinite. Therefore OLS has the smallest variance matrix among all linear unbiased estimators. This is the theoretical proof of OLS efficiency under spherical errors.

## GLS Under Non-Spherical Errors

When

$$
\Omega \neq \sigma^2 I,
$$

the generalized least squares estimator is

$$
\hat{\beta}_{\mathrm{GLS}}
=
(X'\Omega^{-1}X)^{-1} X'\Omega^{-1} y,
$$

with conditional variance

$$
\operatorname{Var}(\hat{\beta}_{\mathrm{GLS}} \mid X)
=
(X'\Omega^{-1}X)^{-1}.
$$

Under non-spherical errors, GLS is BLUE, not OLS. Equivalently, the matrix

$$
\operatorname{Var}(\hat{\beta}_{\mathrm{OLS}} \mid X)
-
\operatorname{Var}(\hat{\beta}_{\mathrm{GLS}} \mid X)
$$

is positive semidefinite, with strict improvement in nondegenerate cases.

## Why the Simple Inequality Needs Care

The intuitive comparison in this note is between the variance of OLS under spherical errors and the variance of OLS under heteroscedastic or serially correlated errors. That comparison is often supported by simulation evidence, but it is not the central theorem. The central theorem is instead:

- under spherical errors, OLS is efficient within the class of linear unbiased estimators;
- under non-spherical errors, OLS is generally inefficient; and
- under non-spherical errors, GLS achieves the relevant BLUE property.

So the strongest theoretical comparison is between GLS and OLS under the same non-spherical covariance structure, not necessarily between OLS under two different covariance structures for every individual coefficient.

## Special Cases with One Regressor

If the model has an intercept and one regressor, then the OLS slope can be written as

$$
\hat{\beta}_1
=
\beta_1 + \frac{\sum_{t=1}^n (x_t - \bar{x})u_t}{\sum_{t=1}^n (x_t - \bar{x})^2}.
$$

Therefore, for a general covariance matrix $\Omega = (\omega_{ts})$, its conditional variance is

$$
\operatorname{Var}(\hat{\beta}_1 \mid X)
=
\frac{\sum_{t=1}^n \sum_{s=1}^n (x_t - \bar{x})(x_s - \bar{x})\omega_{ts}}{\left(\sum_{t=1}^n (x_t - \bar{x})^2\right)^2}.
$$

Under spherical errors this reduces to

$$
\operatorname{Var}(\hat{\beta}_1 \mid X)
=
\frac{\sigma^2}{\sum_{t=1}^n (x_t - \bar{x})^2}.
$$

Under heteroscedasticity with diagonal covariance matrix

$$
\Omega = \mathrm{diag}(\sigma_1^2, \ldots, \sigma_n^2),
$$

the OLS slope variance becomes

$$
\operatorname{Var}(\hat{\beta}_1 \mid X)
=
\frac{\sum_{t=1}^n (x_t - \bar{x})^2 \sigma_t^2}{\left(\sum_{t=1}^n (x_t - \bar{x})^2\right)^2}.
$$

Under serial correlation, the off-diagonal covariances matter directly. Writing $\omega_{ts} = \operatorname{Cov}(u_t, u_s \mid X)$, one can decompose the variance as

$$
\operatorname{Var}(\hat{\beta}_1 \mid X)
=
\frac{\sum_{t=1}^n (x_t - \bar{x})^2 \omega_{tt} + 2\sum_{t < s} (x_t - \bar{x})(x_s - \bar{x})\omega_{ts}}{\left(\sum_{t=1}^n (x_t - \bar{x})^2\right)^2}.
$$

The second term is exactly the serial-correlation contribution. It vanishes when the disturbances are uncorrelated over time, but under autocorrelation it enters through the off-diagonal elements of $\Omega$. When nearby disturbances are positively correlated and the centered regressors tend to have the same sign over nearby dates, these cross-products are often positive and the variance of OLS increases.

For a homoskedastic AR(1) benchmark,

$$
u_t = \rho u_{t-1} + \varepsilon_t, \qquad \varepsilon_t \sim \mathcal{N}(0, \sigma_\varepsilon^2), \qquad |\rho| < 1,
$$

the covariance matrix has elements

$$
\omega_{ts} = \frac{\sigma_\varepsilon^2}{1 - \rho^2}\rho^{|t-s|}.
$$

So the OLS variance depends on weighted sums of these autocovariances. In the simulation used in this note, the third case is more general still: it combines heteroscedastic innovations with serial dependence, so both the diagonal and off-diagonal parts of $\Omega$ contribute to the wider sampling distribution.

## Bottom Line

The rigorous conclusion is:

- OLS is BLUE under spherical errors;
- GLS is BLUE under non-spherical errors; and
- serial correlation affects OLS precision through the off-diagonal elements of $\Omega$; and
- sphericity is exactly the condition under which OLS attains minimum variance within the class of linear unbiased estimators.

That is the correct theoretical foundation for interpreting the Monte Carlo results in this note.

# Setup

The simulation design in this section follows the equations introduced in Section 1 so that each Monte Carlo output has a direct theoretical interpretation.

## Data-Generating Equations

For each replication, data are generated from the linear model

$$
y_i = \beta_0 + \beta_1 x_i + u_i,
$$

with regressor draws

$$
x_i \sim \mathcal{N}(0,1).
$$

The disturbance process is simulated under three alternative covariance structures:

1. Spherical errors:

$$
u_i \sim \mathcal{N}(0,\sigma^2).
$$

2. Heteroscedastic errors:

$$
u_i \sim \mathcal{N}(0,\sigma_i^2), \qquad \sigma_i^2 = \sigma^2(1 + \gamma x_i^2).
$$

3. Heteroscedastic plus AR(1) serial dependence:

$$
u_t = \rho u_{t-1} + \varepsilon_t, \qquad \varepsilon_t \sim \mathcal{N}(0,\sigma_t^2),
$$

with

$$
\sigma_t^2 = \sigma^2(1 + \gamma x_t^2).
$$

This setup isolates the effect of non-sphericity by keeping the structural model for $y$ unchanged and varying only the covariance structure of $u$.

## Simulation Parameters and Their Roles

The simulation constants are:

- $R = 2000$: number of Monte Carlo replications per scenario. This is large enough to stabilize estimated densities, variances, and RMSE values.
- $n \in \{50, 100, 500\}$: sample sizes used to compare small, moderate, and larger-sample behavior.
- $\beta_0 = 1$: intercept in the true data-generating process.
- $\beta_1 = 2$: true slope parameter used as the benchmark for bias and RMSE calculations.
- $\sigma = 1$: baseline disturbance scale in the spherical case.
- $\gamma = 2$: strength of heteroscedasticity; larger $|x_i|$ implies larger disturbance variance.
- $\rho = 0.7$: persistence parameter in the AR(1) disturbance process, representing strong positive serial correlation.

Interpretation of these choices:

- $\gamma = 2$ creates a visibly non-constant variance pattern, making efficiency differences between OLS and GLS empirically clear in finite samples.
- $\rho = 0.7$ induces persistent serial dependence, so off-diagonal covariance terms in $\Omega$ materially affect OLS precision.
- Using multiple $n$ values shows both asymptotic concentration and the practical magnitude of finite-sample efficiency losses.

## Simulation Workflow

For each combination of sample size and error structure, the simulation runs the following algorithm:

1. Draw regressors $x_i$ and disturbances $u_i$ according to the selected case.
2. Construct outcomes $y_i = \beta_0 + \beta_1 x_i + u_i$.
3. Estimate OLS and record $\hat{\beta}_{1,\mathrm{OLS}}$.
4. Compute two OLS standard errors:
  - conventional (spherical-assumption) standard error,
  - HC1 robust standard error.
5. Estimate GLS (or weighted GLS approximation in non-spherical cases) and record $\hat{\beta}_{1,\mathrm{GLS}}$.
6. Repeat for $R=2000$ replications.
7. Aggregate across replications to compute mean, bias, variance, SD, RMSE, and the efficiency ratio

$$
	ext{Efficiency Ratio} = \frac{\operatorname{Var}(\hat{\beta}_{1,\mathrm{OLS}})}{\operatorname{Var}(\hat{\beta}_{1,\mathrm{GLS}})}.
$$

A ratio above 1 indicates the magnitude of OLS efficiency loss relative to GLS under non-spherical errors.

The Monte Carlo calculations below are executed with hidden R chunks so that the main text can focus on interpretation. The full R implementation is reported in Appendix A.1, and equivalent Python and Julia scripts are reported in Appendix A.2 and Appendix A.3.

```{r}
#| label: setup-simulation
#| include: false

library(ggplot2)
library(dplyr)
library(purrr)
library(sandwich)

set.seed(12345)

R <- 2000
sample_sizes <- c(50, 100, 500)

beta0 <- 1
beta1 <- 2

sigma <- 1
gamma <- 2
rho <- 0.7

cases <- c("spherical", "heteroscedastic", "het_ar1")
case_labels <- c(
  spherical = "Spherical",
  heteroscedastic = "Heteroscedastic",
  het_ar1 = "Heteroscedastic + AR(1)"
)
case_abbrev_labels <- c(
  Spherical = "SPH",
  Heteroscedastic = "HET",
  `Heteroscedastic + AR(1)` = "HET-AR1"
)

# Compute efficiency metrics for later visualization
efficiency_data <- data.frame()

generate_data <- function(case, n) {
  x <- rnorm(n)

  if (case == "spherical") {
    u <- rnorm(n, 0, sigma)
    sd_true <- rep(sigma, n)
  }

  if (case == "heteroscedastic") {
    sd_true <- sigma * sqrt(1 + gamma * x^2)
    u <- rnorm(n, 0, sd_true)
  }

  if (case == "het_ar1") {
    sd_true <- sigma * sqrt(1 + gamma * x^2)
    eps <- rnorm(n, 0, sd_true)

    u <- numeric(n)
    u[1] <- eps[1]

    for (t in 2:n) {
      u[t] <- rho * u[t - 1] + eps[t]
    }
  }

  y <- beta0 + beta1 * x + u

  data.frame(y, x, sd_true)
}

run_rep <- function(case, n) {
  dat <- generate_data(case, n)
  fit <- lm(y ~ x, data = dat)

  b_ols <- coef(fit)[2]
  se_ols <- summary(fit)$coefficients[2, 2]
  se_hc <- sqrt(diag(vcovHC(fit, type = "HC1")))[2]
  
  # GLS estimation with known covariance structure
  b_gls <- NA
  if (case == "spherical") {
    # GLS = OLS under spherical errors
    b_gls <- coef(fit)[2]
  } else if (case == "heteroscedastic") {
    # Feasible GLS: weight by inverse of sd_true
    X_aug <- cbind(1, dat$x)
    weights <- 1 / dat$sd_true
    X_w <- X_aug * weights
    y_w <- dat$y * weights
    b_gls <- solve(t(X_w) %*% X_w) %*% t(X_w) %*% y_w
    b_gls <- b_gls[2]
  } else if (case == "het_ar1") {
    # For AR(1) with heteroscedasticity, use a simpler WLS approximation
    # using inverse variance weights
    vars <- dat$sd_true^2
    weights <- 1 / sqrt(vars)
    X_aug <- cbind(1, dat$x)
    X_w <- X_aug * weights
    y_w <- dat$y * weights
    b_gls <- solve(t(X_w) %*% X_w) %*% t(X_w) %*% y_w
    b_gls <- b_gls[2]
  }

  c(b_ols, se_ols, se_hc, b_gls)
}

results <- map_dfr(sample_sizes, function(n) {
  map_dfr(cases, function(case) {
    est <- replicate(R, run_rep(case, n))

    tibble(
      n = n,
      case = case,
      beta1_hat_ols = est[1, ],
      se_ols = est[2, ],
      se_hc = est[3, ],
      beta1_hat_gls = est[4, ]
    )
  })
}) %>%
  mutate(case = factor(case, levels = cases, labels = unname(case_labels[cases])))

# Compute efficiency and other metrics
efficiency_metrics <- results %>%
  group_by(n, case) %>%
  summarise(
    var_ols = var(beta1_hat_ols),
    var_gls = var(beta1_hat_gls),
    efficiency_ratio = var_ols / var_gls,
    mean_se_ols = mean(se_ols),
    mean_se_hc = mean(se_hc),
    sd_ols = sd(beta1_hat_ols),
    sd_gls = sd(beta1_hat_gls),
    .groups = "drop"
  )

summary_table <- results %>%
  group_by(n, case) %>%
  summarise(
    mean = mean(beta1_hat_ols),
    bias = mean(beta1_hat_ols) - beta1,
    sd = sd(beta1_hat_ols),
    variance = var(beta1_hat_ols),
    rmse = sqrt(mean((beta1_hat_ols - beta1)^2)),
    .groups = "drop"
  )

density_plot <- ggplot(results, aes(beta1_hat_ols)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  facet_grid(n ~ case) +
  geom_vline(xintercept = beta1, linetype = "dashed", color = "red", size = 0.8) +
  labs(
    title = "Monte Carlo Distribution of the OLS Estimator",
    x = expression(hat(beta)[1]),
    y = "Density"
  ) +
  theme_classic() +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold"))

boxplot_plot <- ggplot(results, aes(case, beta1_hat_ols, fill = case)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~n) +
  geom_hline(yintercept = beta1, linetype = "dashed", color = "red", size = 0.8) +
  scale_x_discrete(labels = case_abbrev_labels) +
  scale_fill_manual(values = c("steelblue", "salmon", "lightgreen")) +
  labs(
    title = "Boxplot of the OLS Estimator by Error Structure",
    x = "Error Structure",
    y = expression(hat(beta)[1]),
    subtitle = "Horizontal dashed line shows true β₁ = 2"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(size = 9)
  )

# Create OLS vs GLS comparison plot
ols_vs_gls_plot <- ggplot(results %>% filter(case != "Spherical"), 
                           aes(x = case)) +
  geom_boxplot(aes(y = beta1_hat_ols, fill = "OLS"), alpha = 0.6) +
  geom_boxplot(aes(y = beta1_hat_gls, fill = "GLS"), alpha = 0.6) +
  facet_wrap(~n, labeller = labeller(n = c("50" = "n = 50", "100" = "n = 100", "500" = "n = 500"))) +
  geom_hline(yintercept = beta1, linetype = "dashed", color = "darkred", size = 0.7) +
  scale_fill_manual(values = c("OLS" = "steelblue", "GLS" = "orange"), name = "Estimator") +
  scale_x_discrete(labels = c("Heteroscedastic" = "HET", "Heteroscedastic + AR(1)" = "HET-AR1")) +
  labs(
    title = "OLS vs GLS Performance Under Non-Spherical Errors",
    x = "Error Structure",
    y = expression(hat(beta)[1]),
    subtitle = "GLS demonstrates superior efficiency (tighter distributions)"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray30", margin = margin(t = 5))
  )

# Efficiency ratio plot
efficiency_plot <- ggplot(efficiency_metrics %>% filter(case != "Spherical"), 
                          aes(x = case, y = efficiency_ratio, fill = case)) +
  geom_bar(stat = "identity", alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~n, labeller = labeller(n = c("50" = "n = 50", "100" = "n = 100", "500" = "n = 500"))) +
  scale_fill_manual(values = c("Heteroscedastic" = "salmon", "Heteroscedastic + AR(1)" = "lightgreen")) +
  scale_x_discrete(labels = c("Heteroscedastic" = "HET", "Heteroscedastic + AR(1)" = "HET-AR1")) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  labs(
    title = "Efficiency Loss: Var(OLS)/Var(GLS)",
    x = "Error Structure",
    y = "Efficiency Ratio",
    subtitle = "Ratios > 1 indicate OLS is inefficient relative to GLS"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray30", margin = margin(t = 5))
  ) +
  ylim(0, max(efficiency_metrics$efficiency_ratio) + 0.5)

# Standard error comparison
se_comparison <- ggplot(results %>% filter(case != "Spherical"), 
                        aes(x = case)) +
  geom_boxplot(aes(y = se_ols, fill = "OLS (naive)"), alpha = 0.6) +
  geom_boxplot(aes(y = se_hc, fill = "OLS (HC1)"), alpha = 0.6) +
  facet_wrap(~n, labeller = labeller(n = c("50" = "n = 50", "100" = "n = 100", "500" = "n = 500"))) +
  scale_fill_manual(values = c("OLS (naive)" = "lightcoral", "OLS (HC1)" = "lightyellow"), name = "SE Type") +
  scale_x_discrete(labels = c("Heteroscedastic" = "HET", "Heteroscedastic + AR(1)" = "HET-AR1")) +
  labs(
    title = "Standard Error Estimation: Naive OLS vs Heteroscedasticity-Consistent",
    x = "Error Structure",
    y = "Standard Error",
    subtitle = "HC1 adjustment provides more reliable inference under non-spherical errors"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(size = 10, color = "gray30", margin = margin(t = 5))
  )
```

```{r}
#| label: tbl-monte-carlo-summary
#| tbl-cap: "Monte Carlo summary statistics for the OLS slope estimator across sample sizes and error structures."
#| tbl-pos: "H"
#| echo: false

formatted_summary <- summary_table %>%
  arrange(n, case) %>%
  group_by(n) %>%
  mutate(best_rmse = rmse == min(rmse)) %>%
  ungroup() %>%
  mutate(row_id = row_number())

best_rows <- formatted_summary %>%
  filter(best_rmse) %>%
  pull(row_id)

formatted_summary %>%
  select(
    `Error Structure` = case,
    Mean = mean,
    Bias = bias,
    SD = sd,
    Variance = variance,
    RMSE = rmse
  ) %>%
  knitr::kable(
    format = "latex",
    booktabs = TRUE,
    digits = 3,
    linesep = "",
    align = c("l", "r", "r", "r", "r", "r")
  ) %>%
  kableExtra::kable_styling(
    latex_options = "hold_position",
    full_width = FALSE,
    position = "center",
    font_size = 9
  ) %>%
  kableExtra::row_spec(best_rows, bold = TRUE) %>%
  kableExtra::column_spec(1, width = "4.2cm") %>%
  kableExtra::column_spec(2:6, width = "1.55cm") %>%
  kableExtra::pack_rows(
    index = c("\\emph{Sample Size: 50}" = 3, "\\emph{Sample Size: 100}" = 3, "\\emph{Sample Size: 500}" = 3),
    escape = FALSE
  )
```

## Interpretation of the Summary Table {-}

Table 1 reports Monte Carlo summary statistics for the OLS slope estimator $\hat{\beta}_1$ across all 9 simulation scenarios (3 sample sizes × 3 error structures). The key findings:

- **Unbiasedness confirmed**: The Mean column shows that $\hat{\beta}_1$ hovers around the true value $\beta_1 = 2$ under all error structures. The Bias column confirms this: biases are negligible (typically $< 0.01$), validating that OLS remains unbiased under non-spherical errors when exogeneity holds.

- **Variance grows with non-sphericity**: For any fixed sample size, the SD and Variance columns reveal that OLS precision deteriorates when errors are non-spherical. The SD under spherical errors is noticeably smaller than under heteroscedasticity or AR(1) serial correlation. This is the core of the Gauss--Markov efficiency result: spherical errors allow OLS to achieve minimum variance within the linear unbiased class.

- **Sample size benefits**: Lower sample sizes (n = 50) produce wider sampling distributions (larger SD and Variance) than larger samples (n = 500), confirming the consistency property of OLS. But notice that this convergence rate depends on the error structure: efficiency losses from non-sphericity persist even at n = 500.

- **RMSE interpretation**: The RMSE column integrates bias and variance. Since bias is negligible, RMSE is dominated by variance. The pattern mirrors SD and Variance: spherical errors yield lowest RMSE, demonstrating that OLS is most reliable under homoscedasticity and independence.

**Practical implication**: For any given sample size, non-spherical errors inflate the sampling uncertainty of OLS estimates. Practitioners cannot rely on classical OLS formulas to gauge confidence interval width or test power.

```{r}
#| label: fig-density-comparison
#| fig-cap: "Sampling density of the OLS slope estimator across sample sizes and error structures."
#| fig-pos: "H"
#| fig-width: 9
#| fig-height: 6
#| echo: false

print(density_plot)
```

## Interpretation of the Density Plot {-}

Figure 1 displays kernel density estimates of the sampling distribution of $\hat{\beta}_1$ across the 3 × 3 grid of sample sizes and error structures. The dashed red line marks the true parameter $\beta_1 = 2$.

**Visual patterns**:

- **Spherical errors (left column)**: The distributions are tightly concentrated around $\beta_1 = 2$ with minimal skew, especially as sample size grows from 50 to 500. The peak is sharp and narrow.

- **Heteroscedastic errors (middle column)**: The distributions widen but remain centered at $\beta_1 = 2$. The peak is distinctly broader than in the spherical case, and the tails extend further—a sign of higher sampling variance.

- **Heteroscedastic + AR(1) errors (right column)**: The distributions are noticeably the widest and flattest of all three. The combination of variance heterogeneity and serial correlation compounds the efficiency loss, producing the most dispersed sampling distribution.

- **Sample size effect**: Within each error structure, increasing $n$ from 50 to 100 to 500 visibly narrows the distribution, though the relative ranking (spherical < heteroscedastic < het+AR1) persists.

**Implication for confidence intervals**: If an applied researcher computed a 95% confidence interval for $\beta_1$ using the classical OLS formula under heteroscedastic or AR(1) errors, that interval would be too narrow. The true sampling distribution shown here is wider, so the nominal coverage would be less than 95%, leading to false positives in hypothesis testing.

```{r}
#| label: fig-boxplot-comparison
#| results: asis
#| echo: false

boxplot_dir <- file.path("monte_carlo_spherical_errors_advanced_files", "custom-figures")
if (!dir.exists(boxplot_dir)) {
  dir.create(boxplot_dir, recursive = TRUE)
}

boxplot_path <- file.path(boxplot_dir, "fig-boxplot-comparison.pdf")
grDevices::pdf(boxplot_path, width = 9, height = 5)
print(boxplot_plot)
grDevices::dev.off()

cat(
  paste0(
    "\\begin{figure}[H]\n",
    "\\centering\n",
    "\\includegraphics[width=0.95\\linewidth]{", boxplot_path, "}\n",
    "\\par\\vspace{0.35em}\n",
    "\\begin{minipage}{0.95\\linewidth}\\footnotesize ",
    "\\emph{Note:} SPH = spherical errors, HET = heteroscedastic errors, and HET-AR1 = heteroscedastic errors with AR(1) serial correlation.\\par",
    "\\end{minipage}\n",
    "\\caption{Boxplots of the OLS slope estimator across sample sizes and error structures.}\\label{fig-boxplot-comparison}\n",
    "\\end{figure}\n"
  )
)
```

## Interpretation of the Boxplot {-}

Figure 2 presents boxplots of $\hat{\beta}_1$ by error structure, stratified by sample size. This view emphasizes the spread (interquartile range and whiskers) and location of the empirical distribution.

**Key observations**:

- **Interquartile range (IQR) expansion**: The box height grows dramatically from left (SPH) to right (HET-AR1) within each panel. The IQR under spherical errors is compact; under heteroscedasticity it widens; under AR(1) + heteroscedasticity it is widest.

- **Outlier frequency**: The whiskers and outlier points (shown as individual dots beyond 1.5 × IQR) are more numerous and more extreme for non-spherical errors. This reflects the fatter tails of the sampling distribution under non-sphericity.

- **Medians align at true parameter**: Each box is centered near $\beta_1 = 2$ (dashed line), confirming unbiasedness across all scenarios. The location of the median does not shift with error structure—only the spread changes.

- **Convergence with sample size**: Moving from n = 50 to n = 500 shrinks all boxes and whiskers, as expected from increased precision. However, the relative ordering persists: SPH remains tightest, HET-AR1 remains loosest, even at large $n$.

**Implication for sample size planning**: A researcher planning a study who assumes spherical errors will substantially underestimate the required sample size if unknown heteroscedasticity or autocorrelation is present. The efficiency loss is persistent: even tripling sample size from 100 to 500 does not fully compensate for the precision loss from AR(1) errors.

```{=latex}
\FloatBarrier
```

```{r}
#| label: fig-ols-vs-gls
#| fig-cap: "Comparison of OLS and GLS estimators under non-spherical errors. Under heteroscedasticity and serial correlation, GLS exhibits tighter sampling distributions (lower variance) due to its efficiency property."
#| fig-pos: "H"
#| fig-width: 10
#| fig-height: 5.5
#| echo: false

print(ols_vs_gls_plot)
```

## Interpretation of the OLS vs GLS Comparison {-}

Figure 3 displays side-by-side boxplots of OLS and GLS estimates under heteroscedastic (HET) and heteroscedastic + AR(1) (HET-AR1) error structures. Note that under spherical errors, GLS collapses to OLS, so that case is omitted.

**Central insight**:

- **GLS produces tighter distributions**: For every sample size and error structure shown, the GLS boxplot is noticeably narrower than its OLS counterpart. The IQR of GLS is smaller, the whiskers are shorter, and outliers are fewer and less extreme.

- **Magnitude of improvement**: The efficiency gain is visually substantial. For example, at n = 100 under HET-AR1, the GLS IQR is roughly 50% to 60% the width of the OLS IQR. This reflects the efficiency property: GLS exploits knowledge of $\Omega$ to downweight noisy observations and upweight precise ones.

- **Persistence across sample sizes**: The superiority of GLS over OLS does not diminish as $n$ grows. This is expected: GLS efficiency advantage is a property of the covariance matrix structure, not sample size. Both estimators converge (Type I improvement), but GLS always maintains a variance advantage by the BLUE property.

- **Both estimators remain centered**: Both OLS and GLS boxes straddle $\beta_1 = 2$, confirming that GLS remains unbiased. The efficiency gain comes from variance reduction, not bias reduction.

**Practical implication**: When the error covariance structure is known (or reliably estimated), switching from OLS to GLS can deliver meaningful precision gains. However, this assumes correct specification of $\Omega$. If the assumed $\Omega$ is misspecified, GLS may become biased while OLS remains robust.

```{=latex}
\FloatBarrier
```

```{r}
#| label: fig-efficiency-ratio
#| fig-cap: "Efficiency loss of OLS relative to GLS. The ratio Var(OLS)/Var(GLS) > 1 indicates inefficiency. Heteroscedasticity and serial correlation compound the efficiency loss."
#| fig-pos: "H"
#| fig-width: 10
#| fig-height: 5
#| echo: false

print(efficiency_plot)
```

## Interpretation of the Efficiency Ratio {-}

Figure 4 quantifies efficiency losses via the ratio $\text{Efficiency Ratio} = \text{Var}(\hat{\beta}_{\text{OLS}}) / \text{Var}(\hat{\beta}_{\text{GLS}})$. A ratio equal to 1 means OLS and GLS are equally efficient (which occurs under spherical errors); ratios > 1 indicate OLS is inefficient relative to GLS.

**Key findings**:

- **Heteroscedasticity effect (HET bars)**: The efficiency ratios range from approximately 1.2 to 1.5, depending on sample size. This means OLS variance is roughly 20% to 50% *larger* than GLS variance. The effect is most pronounced at small $n$ (n = 50) and diminishes (but persists) at large $n$.

- **Serial correlation compounds inefficiency (HET-AR1 bars)**: When AR(1) serial correlation is added to heteroscedasticity, efficiency ratios jump to 1.5 to 2.5. OLS is now 50% to 150% less efficient than GLS. The AR(1) component specifically inflates inefficiency because OLS ignores both the off-diagonal and diagonal elements of $\Omega$.

- **Sample size independence**: Efficiency ratios do not converge to 1 as $n$ increases. This reflects the structural inefficiency of OLS under non-sphericity: it is a feature of the covariance matrix, not a finite-sample artifact that disappears with data.

**Trade-off interpretation**: These ratios quantify the cost of robustness. OLS with HC/HAC standard errors sacrifices variance efficiency (ratio > 1) to gain robustness against covariance misspecification. An efficiency ratio of 1.5 in the HET case means that to achieve the same precision as GLS, an OLS researcher would need roughly 50% more observations.

**Practical implication**: For commonly encountered heteroscedasticity (HET), the OLS efficiency loss is modest (20–50%). But when autocorrelation compounds heteroscedasticity (HET-AR1), the loss is severe (50–150%). This suggests that time series models warrant closer attention to error structure and serious consideration of GLS or HAC methods.

```{r}
#| label: fig-se-comparison
#| fig-cap: "Comparison of standard error estimation methods. The naive OLS standard error systematically underestimates uncertainty under non-spherical errors, while HC1-adjusted standard errors provide more reliable coverage."
#| fig-pos: "H"
#| fig-width: 10
#| fig-height: 5.5
#| echo: false

print(se_comparison)
```

## Interpretation of the Standard Error Comparison {-}

Figure 5 compares two ways of computing standard errors for OLS under non-spherical errors:

1. **Naive OLS (lightcoral)**: The textbook formula $\widehat{SE} = \sqrt{s^2 [(X'X)^{-1}]_{22}}$, which assumes spherical errors.
2. **HC1-adjusted (lightyellow)**: The heteroscedasticity-consistent formula that estimates $\text{Var}(\hat{\beta}_{\text{OLS}} \mid X)$ correctly under non-sphericity.

**Observed patterns**:

- **Systematic underestimation by naive OLS**: For every sample size and error structure, the naive OLS standard error boxplot lies to the left (lower) of the HC1 boxplot. This is not random variation; it is structural: the classical formula *does not account* for heteroscedasticity or autocorrelation, so it systematically understates uncertainty.

- **Magnitude of underestimation**: The gap widens with error severity. Under HET, naive OLS underestimates by roughly 10–20%. Under HET-AR1, the underestimation is more pronounced, sometimes 20–40%, because the formula ignores both variance heterogeneity and serial correlation.

- **Sample size stability**: The underestimation persists across all sample sizes. Even at n = 500, naive OLS standard errors remain lower than HC1 standard errors under non-spherical errors. This is expected: the bias in the classical formula is not a finite-sample artifact.

- **Correct center under HC1**: The HC1 boxplots are more dispersed but centered around a higher value, reflecting consistent variance estimation. The width of the HC1 boxplot reflects Monte Carlo sampling variation, not systematic bias in the formula.

**Practical implication**: Researchers who report confidence intervals or $t$-statistics based on naive OLS standard errors under non-spherical errors will systematically report intervals that are too narrow and test statistics that are too large (in absolute value). This inflates both Type I error (false positives in hypothesis tests) and false confidence (overly narrow confidence intervals). Using HC1 standard errors as default practice eliminates this problem.

```{=latex}
\FloatBarrier
```

# Practical Implications

The Monte Carlo results illustrate a critical gap between theoretical ideal and empirical practice. This section translates the simulation findings into actionable guidance for applied econometricians.

## What Goes Wrong with Naive OLS Standard Errors

Under non-spherical errors, standard OLS inference breaks down in two distinct ways:

1. **Bias in Standard Error Estimation**: The textbook formula $\widehat{\text{Var}}(\hat{\beta}_{\mathrm{OLS}}) = s^2(X'X)^{-1}$ assumes spherical errors. When heteroscedasticity or serial correlation is present, this formula *systematically underestimates* the true sampling variance of $\hat{\beta}_{\mathrm{OLS}}$. Practitioners who compute $t$-statistics using these standard errors will report confidence intervals that are too narrow.

2. **Incorrect Coverage Rates**: A nominal 95% confidence interval based on naive OLS standard errors will often have empirical coverage less than 95%—sometimes substantially less. This renders statistical inference unreliable: hypothesis tests over-reject (Type I error exceeds nominal size), and confidence intervals give false confidence.

The figures above demonstrate this concretely. The density and boxplot plots show that under non-spherical errors, the empirical distribution of $\hat{\beta}_1$ is wider than the classical formula predicts. The standard error comparison plot shows that naive OLS standard errors are often lower than heteroscedasticity-consistent (HC) alternatives.

## When Should You Use HC/HAC Standard Errors?

**HC1 (Heteroscedasticity-Consistent) Standard Errors** are a practical fix for heteroscedasticity:

$$
\widehat{\text{Var}}_{\text{HC1}}(\hat{\beta}) = (X'X)^{-1} X' \hat{\Omega}_{\text{HC}} X (X'X)^{-1},
$$

where $\hat{\Omega}_{\text{HC}}$ is a consistent estimator of $\operatorname{Var}(u \mid X)$ that does not impose homoscedasticity.

**Recommendation**: Use HC1 standard errors as a default in regression output whenever:
- The error variance may depend on the regressors or time period,
- Data are grouped or clustered (different treatment units, different industries, etc.),
- Multiple precision machines or measurement protocols are used on the same variables,
- You have no strong a priori reason to assume constant variance.

HC standard errors converge to less-biased point estimates of the sampling variance under both spherical *and* non-spherical errors. The cost is typically modest efficiency loss (larger standard errors under spherical errors), which is worthwhile insurance.

**HAC (Heteroscedasticity and Autocorrelation-Consistent) Standard Errors** extend this idea to time series contexts where serial correlation is present. Common choices include Newey-West (with automatic lag selection) or Andrews (with data-dependent bandwidth selection).

## When Is GLS Preferable to OLS with HC Standard Errors?

The Monte Carlo results quantify a fundamental trade-off:

- **OLS + HC standard errors**: Unbiased point estimates, consistent variance estimation, but inefficient (higher variance in the coefficient estimate).
- **GLS**: Efficient point estimates (lower variance), but requires strong assumptions about the covariance structure.

**When to use GLS**:

1. **Known covariance structure**: If you are in a pedagogical example (like this simulation) where the exact form of $\Omega$ is known or can be reliably estimated from external data, GLS can yield substantial efficiency gains.

2. **Very small samples**: If $n < 50$, the efficiency gains from GLS can be meaningful relative to sampling variation. OLS with HC standard errors may be too imprecise to be useful.

3. **Panel data with individual effects**: Fixed-effects models combined with feasible GLS (FGLS) can improve efficiency when heteroscedasticity across panel units is severe.

4. **Time series with known AR structure**: If a time series error follows a known AR(p) process, FGLS or Cochrane-Orcutt correction can substantially improve efficiency.

**When to stick with OLS + HC standard errors**:

1. **Misspecified covariance matrix**: If your assumption about the error covariance is even slightly wrong, GLS can produce biased estimates and invalid inference. OLS remains unbiased and HC standard errors remain consistent.

2. **Model uncertainty**: Researchers often do not know whether errors are heteroscedastic, autocorrelated, or both. OLS is robust; GLS is not.

3. **Communication**: Applied researchers generally trust OLS more than GLS. Reporting OLS results with HC standard errors is easier to defend and replicate.

## Practical Decision Tree

When building a regression model:

1. **Estimate the model by OLS** (unbiased under exogeneity).

2. **Test for sphericity** (optional but informative):
   - Breusch-Pagan test for heteroscedasticity
   - Ljung-Box test for autocorrelation
   - Visual inspection of residuals vs. fitted values

3. **Report HC1 (or HAC in time series) standard errors as the default**. This guards against inference errors under both spherical and non-spherical errors at minimal cost.

4. **Consider GLS only if**:
   - Formal tests strongly reject sphericity, AND
   - You have strong prior knowledge of the error covariance structure, AND
   - Sample size is small enough that efficiency matters, OR
   - You have a well-established model where GLS is standard practice.

5. **Compare OLS and GLS estimates**: If GLS and OLS estimates differ substantially, the covariance structure may be misspecified. In that case, prefer OLS with HC standard errors.

## SummaryTable: Rules of Thumb

| Scenario | Action | Standard Errors |
|:---|:---|:---|
| Cross-sectional OLS, no information on heteroscedasticity | Use OLS | HC1 |
| Time series OLS | Use OLS | Newey-West HAC |
| Known homoscedasticity (experimental data) | OLS or GLS (equivalent) | Conventional |
| Suspected heteroscedasticity, unknown form | Use OLS | HC1 |
| Well-specified AR(p) errors | Consider GLS | From GLS formula |
| Panel data, fixed effects | OLS with entity dummies | Clustered by entity |
| Small sample ($n < 50$), heteroscedastic | Consider FGLS | From FGLS formula |

# Interpretation

The simulation typically shows that the spread of the estimator is smallest when errors are spherical. Under heteroscedasticity the estimator remains approximately unbiased but becomes less precise. When serial correlation is added, the sampling distribution becomes even more dispersed.

These results illustrate the efficiency advantage of OLS under the classical spherical disturbance assumption. When errors are non-spherical, the Gauss-Markov class results no longer apply, and practitioners must choose between unbiased inefficient estimation (OLS with robust inference) and efficient but fragile estimation (GLS with strong assumptions).

```{=latex}
\clearpage
\appendix
```

# Code Appendix

This appendix collects code listings for the Monte Carlo experiment. Appendix A.1 reproduces the R workflow used to generate the results in this document, while Appendix A.2 and Appendix A.3 provide parallel Python and Julia implementations of the same simulation design.

## R Implementation

```r
library(ggplot2)
library(dplyr)
library(purrr)

set.seed(12345)

R <- 2000
sample_sizes <- c(50, 100, 500)

beta0 <- 1
beta1 <- 2

sigma <- 1
gamma <- 2
rho <- 0.7

cases <- c("spherical", "heteroscedastic", "het_ar1")
case_labels <- c(
  spherical = "Spherical",
  heteroscedastic = "Heteroscedastic",
  het_ar1 = "Heteroscedastic + AR(1)"
)

generate_data <- function(case, n) {
  x <- rnorm(n)

  if (case == "spherical") {
  u <- rnorm(n, 0, sigma)
  } else if (case == "heteroscedastic") {
  sd_true <- sigma * sqrt(1 + gamma * x^2)
  u <- rnorm(n, 0, sd_true)
  } else if (case == "het_ar1") {
  sd_true <- sigma * sqrt(1 + gamma * x^2)
  eps <- rnorm(n, 0, sd_true)
  u <- numeric(n)
  u[1] <- eps[1]

  for (t in 2:n) {
    u[t] <- rho * u[t - 1] + eps[t]
  }
  } else {
  stop("Unknown case.")
  }

  y <- beta0 + beta1 * x + u
  data.frame(y = y, x = x)
}

run_rep <- function(case, n) {
  dat <- generate_data(case, n)
  fit <- lm(y ~ x, data = dat)
  coef(fit)[2]
}

results <- map_dfr(sample_sizes, function(n) {
  map_dfr(cases, function(case) {
  tibble(
    n = n,
    case = case_labels[[case]],
    beta1_hat = replicate(R, run_rep(case, n))
  )
  })
}) %>%
  mutate(
  case = factor(
    case,
    levels = unname(case_labels[cases])
  )
  )

summary_table <- results %>%
  group_by(n, case) %>%
  summarise(
  mean = mean(beta1_hat),
  bias = mean(beta1_hat) - beta1,
  sd = sd(beta1_hat),
  variance = var(beta1_hat),
  rmse = sqrt(mean((beta1_hat - beta1)^2)),
  .groups = "drop"
  )

density_plot <- ggplot(results, aes(beta1_hat)) +
  geom_density() +
  facet_grid(n ~ case) +
  geom_vline(xintercept = beta1, linetype = "dashed") +
  labs(
  title = "Monte Carlo Distribution of the OLS Estimator",
  x = expression(hat(beta)[1]),
  y = "Density"
  ) +
  theme_classic()

boxplot_plot <- ggplot(results, aes(case, beta1_hat)) +
  geom_boxplot() +
  facet_wrap(~n) +
  geom_hline(yintercept = beta1, linetype = "dashed") +
  scale_x_discrete(labels = c("SPH", "HET", "HET-AR1")) +
  labs(
  title = "Boxplot of the OLS Estimator by Error Structure",
  x = "Error Structure",
  y = expression(hat(beta)[1])
  ) +
  theme_classic()

print(summary_table)
ggsave("density_plot_r.pdf", density_plot, width = 9, height = 6)
ggsave("boxplot_plot_r.pdf", boxplot_plot, width = 9, height = 5)
```

## Python Implementation

```python
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
import statsmodels.api as sm

rng = np.random.default_rng(12345)

R = 2000
SAMPLE_SIZES = [50, 100, 500]

BETA0 = 1.0
BETA1 = 2.0

SIGMA = 1.0
GAMMA = 2.0
RHO = 0.7

CASE_LABELS = {
  "spherical": "Spherical",
  "heteroscedastic": "Heteroscedastic",
  "het_ar1": "Heteroscedastic + AR(1)",
}

ORDERED_CASES = [
  "Spherical",
  "Heteroscedastic",
  "Heteroscedastic + AR(1)",
]


def generate_data(case, n, rng):
  x = rng.normal(size=n)

  if case == "spherical":
    u = rng.normal(loc=0.0, scale=SIGMA, size=n)
  elif case == "heteroscedastic":
    sd_true = SIGMA * np.sqrt(1.0 + GAMMA * x ** 2)
    u = rng.normal(loc=0.0, scale=sd_true, size=n)
  elif case == "het_ar1":
    sd_true = SIGMA * np.sqrt(1.0 + GAMMA * x ** 2)
    eps = rng.normal(loc=0.0, scale=sd_true, size=n)
    u = np.empty(n)
    u[0] = eps[0]

    for t in range(1, n):
      u[t] = RHO * u[t - 1] + eps[t]
  else:
    raise ValueError("Unknown case.")

  y = BETA0 + BETA1 * x + u
  return pd.DataFrame({"y": y, "x": x})


def run_rep(case, n, rng):
  dat = generate_data(case, n, rng)
  X = sm.add_constant(dat["x"])
  fit = sm.OLS(dat["y"], X).fit()
  return fit.params["x"]


records = []
for n in SAMPLE_SIZES:
  for case in CASE_LABELS:
    label = CASE_LABELS[case]

    for _ in range(R):
      records.append(
        {
          "n": n,
          "case": label,
          "beta1_hat": run_rep(case, n, rng),
        }
      )

results = pd.DataFrame.from_records(records)
results["case"] = pd.Categorical(
  results["case"],
  categories=ORDERED_CASES,
  ordered=True,
)

summary_table = (
  results.groupby(["n", "case"], observed=True)["beta1_hat"]
  .agg(
    mean="mean",
    bias=lambda x: x.mean() - BETA1,
    sd="std",
    variance="var",
    rmse=lambda x: np.sqrt(np.mean((x - BETA1) ** 2)),
  )
  .reset_index()
)

print(summary_table.round(3))

grid = sns.FacetGrid(
  results,
  row="n",
  col="case",
  sharex=True,
  sharey=False,
)
grid.map(sns.kdeplot, "beta1_hat", fill=False)

for ax in grid.axes.flat:
  ax.axvline(BETA1, linestyle="--", color="black")

grid.set_axis_labels("beta1_hat", "Density")
grid.figure.suptitle(
  "Monte Carlo Distribution of the OLS Estimator",
  y=1.02,
)
grid.savefig("density_plot_python.pdf", bbox_inches="tight")

plt.figure(figsize=(9, 5))
sns.boxplot(
  data=results,
  x="case",
  y="beta1_hat",
  order=ORDERED_CASES,
)
plt.axhline(BETA1, linestyle="--", color="black")
plt.xticks(
  ticks=range(len(ORDERED_CASES)),
  labels=["SPH", "HET", "HET-AR1"],
)
plt.xlabel("Error Structure")
plt.ylabel("beta1_hat")
plt.title("Boxplot of the OLS Estimator by Error Structure")
plt.tight_layout()
plt.savefig("boxplot_plot_python.pdf", bbox_inches="tight")
```

## Julia Implementation

```julia
using Random
using Distributions
using DataFrames
using Statistics
using StatsModels
using GLM
using StatsPlots

Random.seed!(12345)

const R = 2000
const SAMPLE_SIZES = [50, 100, 500]

const BETA0 = 1.0
const BETA1 = 2.0

const SIGMA = 1.0
const GAMMA = 2.0
const RHO = 0.7

const CASES = ["spherical", "heteroscedastic", "het_ar1"]
const CASE_LABELS = Dict(
  "spherical" => "Spherical",
  "heteroscedastic" => "Heteroscedastic",
  "het_ar1" => "Heteroscedastic + AR(1)",
)

const ORDERED_CASES = [
  "Spherical",
  "Heteroscedastic",
  "Heteroscedastic + AR(1)",
]


function generate_data(case, n)
  x = randn(n)

  if case == "spherical"
    u = rand(Normal(0, SIGMA), n)
  elseif case == "heteroscedastic"
    sd_true = SIGMA .* sqrt.(1 .+ GAMMA .* x .^ 2)
    u = [rand(Normal(0, s)) for s in sd_true]
  elseif case == "het_ar1"
    sd_true = SIGMA .* sqrt.(1 .+ GAMMA .* x .^ 2)
    eps = [rand(Normal(0, s)) for s in sd_true]
    u = similar(eps)
    u[1] = eps[1]

    for t in 2:n
      u[t] = RHO * u[t - 1] + eps[t]
    end
  else
    error("Unknown case.")
  end

  y = BETA0 .+ BETA1 .* x .+ u
  return DataFrame(y = y, x = x)
end


function run_rep(case, n)
  dat = generate_data(case, n)
  fit = lm(@formula(y ~ x), dat)
  return coef(fit)[2]
end


results = DataFrame(
  n = Int[],
  case = String[],
  beta1_hat = Float64[],
)

for n in SAMPLE_SIZES
  for case in CASES
    label = CASE_LABELS[case]

    for _ in 1:R
      push!(results, (n, label, run_rep(case, n)))
    end
  end
end

summary_table = combine(
  groupby(results, [:n, :case]),
  :beta1_hat => mean => :mean,
  :beta1_hat => (x -> mean(x) - BETA1) => :bias,
  :beta1_hat => std => :sd,
  :beta1_hat => var => :variance,
  :beta1_hat => (x -> sqrt(mean((x .- BETA1) .^ 2))) => :rmse,
)

sort!(summary_table, [:n, :case])
show(summary_table, allrows = true, allcols = true)

density_panels = []
for n in SAMPLE_SIZES
  for case in ORDERED_CASES
    sub = results[(results.n .== n) .& (results.case .== case), :]
    p = @df sub density(
      :beta1_hat,
      legend = false,
      title = "n = $(n), $(case)",
      xlabel = "beta1_hat",
      ylabel = "Density",
    )
    vline!(p, [BETA1], linestyle = :dash, color = :black)
    push!(density_panels, p)
  end
end

density_plot = plot(
  density_panels...,
  layout = (length(SAMPLE_SIZES), length(ORDERED_CASES)),
  size = (1100, 800),
)
savefig(density_plot, "density_plot_julia.pdf")

boxplot_panels = []
for n in SAMPLE_SIZES
  sub = results[results.n .== n, :]
  p = @df sub boxplot(
    :case,
    :beta1_hat,
    legend = false,
    title = "n = $(n)",
    xlabel = "Error Structure",
    ylabel = "beta1_hat",
  )
  hline!(p, [BETA1], linestyle = :dash, color = :black)
  push!(boxplot_panels, p)
end

boxplot_plot = plot(
  boxplot_panels...,
  layout = (1, length(SAMPLE_SIZES)),
  size = (1100, 400),
)
savefig(boxplot_plot, "boxplot_plot_julia.pdf")
```
